---
title: "Optimization methods for MLE"
author: "Maciej Beręsewicz"
format:
  html:
    embed-resources: true
    number-sections: true
    df-print: kable
execute:
  eval: true
  warning: true
  message: true
---

## Introduction

When maximizing the log-likelihood numerically, the optimizer may explore **invalid parameter regions** (e.g., $\lambda < 0$ for Poisson), producing `NaN` warnings. This notebook explains **why** these warnings occur, compares Newton-Raphson vs BFGS behavior, and shows how to fix them.

We use the **zero-truncated Poisson** log-likelihood from the MLE notebook:

$$
\log L(\lambda) = \sum_i x_i \log \lambda - n \log(e^\lambda - 1) - \sum_i \log(x_i!)
$$

## Data

::: panel-tabset
### R

```{r}
library(maxLik)
d <- c(1645, 183, 37, 13, 1, 1)
x <- rep(1:6, d)
```

### Python

```{python}
import numpy as np
from scipy.optimize import minimize
import warnings

d = np.array([1645, 183, 37, 13, 1, 1])
x = np.repeat(np.arange(1, 7), d)
```
:::

## Log-likelihood without safeguard

::: panel-tabset
### R

```{r}
ll <- function(par, x) {
  sum(x) * log(par) - length(x) * log(exp(par) - 1)
}

grad <- function(par, x) {
  sum(x) / par - length(x) * exp(par) / (exp(par) - 1)
}

hess <- function(par, x) {
  -sum(x) / par^2 + length(x) * exp(par) / (exp(par) - 1)^2
}
```

### Python

```{python}
def neg_ll(par, x):
    """Negative log-likelihood (for minimization)"""
    lam = par[0]
    return -(np.sum(x) * np.log(lam) - len(x) * np.log(np.exp(lam) - 1))

def neg_grad(par, x):
    """Negative gradient"""
    lam = par[0]
    return np.array([-(np.sum(x) / lam - len(x) * np.exp(lam) / (np.exp(lam) - 1))])
```
:::

## Why do warnings appear?

The log-likelihood contains `log(par)` and `log(exp(par) - 1)`. Both fail when `par <= 0`:

::: panel-tabset
### R

```{r}
## Evaluating at invalid parameter values
cat("log(-0.3) =", log(-0.3), "\n")
cat("log(exp(-0.3) - 1) = log(", exp(-0.3) - 1, ") =", log(exp(-0.3) - 1), "\n")
cat("\n")
cat("log(0.35) =", round(log(0.35), 4), " -- valid\n")
cat("log(exp(0.35) - 1) =", round(log(exp(0.35) - 1), 4), " -- valid\n")
```

### Python

```{python}
## Evaluating at invalid parameter values
with warnings.catch_warnings():
    warnings.simplefilter("always")
    print(f"log(-0.3) = {np.log(-0.3)}")
    print(f"log(exp(-0.3) - 1) = log({np.exp(-0.3) - 1:.4f}) = {np.log(np.exp(-0.3) - 1)}")
    print()
    print(f"log(0.35) = {np.log(0.35):.4f}  -- valid")
    print(f"log(exp(0.35) - 1) = {np.log(np.exp(0.35) - 1):.4f}  -- valid")
```
:::

## Step-by-step: how the optimizer goes negative

### Newton-Raphson

The Newton-Raphson update rule is:

$$
\lambda^{(t+1)} = \lambda^{(t)} - \frac{f'(\lambda^{(t)})}{f''(\lambda^{(t)})}
$$

Starting from $\lambda^{(0)} = 1$:

::: panel-tabset
### R

```{r}
## Manual Newton-Raphson step from lambda=1
lam0 <- 1.0
g <- grad(lam0, x)
h <- hess(lam0, x)
cat("At lambda = 1.0:\n")
cat("  Gradient =", round(g, 2), "(strongly negative: optimum is to the LEFT)\n")
cat("  Hessian  =", round(h, 2), "\n")
cat("  Full Newton step: lambda_new =", lam0, "- (", round(g, 2), "/", round(h, 2), ") =",
    round(lam0 - g/h, 4), "\n\n")

## The Newton step lands at ~0.31 directly, but during LINE SEARCH
## the algorithm may overshoot. With a step size > 1, it could try:
cat("If line search tries step size 1.5:\n")
overshoot <- lam0 - 1.5 * g/h
cat("  lambda_candidate =", round(overshoot, 4), "\n")
cat("  ll(", round(overshoot, 4), ") =", ll(overshoot, x), "\n")
cat("  -> If candidate < 0, log() produces NaN, line search backtracks\n")
```

### Python

```{python}
## Manual Newton-Raphson step from lambda=1
lam0 = 1.0
g = -(neg_grad([lam0], x)[0])  # positive gradient (for maximization)
## Hessian
h = -np.sum(x) / lam0**2 + len(x) * np.exp(lam0) / (np.exp(lam0) - 1)**2

print(f"At lambda = 1.0:")
print(f"  Gradient = {g:.2f} (strongly negative: optimum is to the LEFT)")
print(f"  Hessian  = {h:.2f}")
print(f"  Full Newton step: lambda_new = {lam0} - ({g:.2f} / {h:.2f}) = {lam0 - g/h:.4f}")
print()

## Line search overshoot example
print("If line search tries step size 1.5:")
overshoot = lam0 - 1.5 * g/h
print(f"  lambda_candidate = {overshoot:.4f}")
with warnings.catch_warnings():
    warnings.simplefilter("always")
    val = np.sum(x) * np.log(overshoot) - len(x) * np.log(np.exp(overshoot) - 1)
    print(f"  ll({overshoot:.4f}) = {val}")
    print("  -> If candidate < 0, log() produces NaN, line search backtracks")
```
:::

### Why BFGS produces more warnings

BFGS approximates the gradient using **finite differences**:

$$
f'(\lambda) \approx \frac{f(\lambda + \varepsilon) - f(\lambda - \varepsilon)}{2\varepsilon}
$$

When $\lambda$ is small, $\lambda - \varepsilon < 0$, producing NaN at **every iteration**:

::: panel-tabset
### R

```{r}
## Simulate what BFGS does near lambda = 0.01
eps <- 1e-4
lam_small <- 0.01
cat("At lambda =", lam_small, ":\n")
cat("  ll(lam + eps) = ll(", lam_small + eps, ") =", round(ll(lam_small + eps, x), 2), "\n")
cat("  ll(lam - eps) = ll(", lam_small - eps, ") =", round(ll(lam_small - eps, x), 2), "-- still valid\n\n")

lam_tiny <- 0.00005
cat("At lambda =", lam_tiny, ":\n")
cat("  ll(lam + eps) = ll(", lam_tiny + eps, ") =", round(ll(lam_tiny + eps, x), 2), "\n")
cat("  ll(lam - eps) = ll(", lam_tiny - eps, ") =", ll(lam_tiny - eps, x), "-- NaN!\n")
```

### Python

```{python}
## Simulate what BFGS does near lambda = 0.01
eps = 1e-4
lam_small = 0.01
print(f"At lambda = {lam_small}:")
print(f"  ll(lam + eps) = {-neg_ll([lam_small + eps], x):.2f}")
print(f"  ll(lam - eps) = {-neg_ll([lam_small - eps], x):.2f} -- still valid")
print()

lam_tiny = 0.00005
print(f"At lambda = {lam_tiny}:")
with warnings.catch_warnings():
    warnings.simplefilter("always")
    print(f"  ll(lam + eps) = {-neg_ll([lam_tiny + eps], x):.2f}")
    print(f"  ll(lam - eps) = {-neg_ll([lam_tiny - eps], x)} -- NaN!")
```
:::

## Comparison: Newton-Raphson vs BFGS

::: panel-tabset
### R

```{r}
cat("=== Newton-Raphson (default) ===\n")
res_nr <- maxLik(logLik = ll, start = 1, x = x)
cat("Iterations:", res_nr$iterations, "\n")
cat("Estimate:", round(coef(res_nr), 6), "\n")
cat("SE:", round(stdEr(res_nr), 5), "\n\n")

cat("=== BFGS ===\n")
res_bfgs <- maxLik(logLik = ll, start = 1, x = x, method = "BFGS")
cat("Iterations:", res_bfgs$iterations, "\n")
cat("Estimate:", round(coef(res_bfgs), 6), "\n")
cat("SE:", round(stdEr(res_bfgs), 5), "\n")
```

### Python

```{python}
print("=== Nelder-Mead (gradient-free) ===")
res_nm = minimize(neg_ll, x0=[0.5], args=(x,), method="Nelder-Mead")
print(f"Iterations: {res_nm.nit}")
print(f"Estimate: {res_nm.x[0]:.6f}")
print()

print("=== BFGS ===")
res_bfgs = minimize(neg_ll, x0=[0.5], args=(x,), method="BFGS", jac=lambda p, x: neg_grad(p, x), options={"disp": False})
print(f"Iterations: {res_bfgs.nit}")
print(f"Estimate: {res_bfgs.x[0]:.6f}")
print()

print("=== L-BFGS-B (bounded BFGS) ===")
res_lbfgsb = minimize(neg_ll, x0=[0.5], args=(x,), method="L-BFGS-B",
                       jac=lambda p, x: neg_grad(p, x), bounds=[(1e-8, None)])
print(f"Iterations: {res_lbfgsb.nit}")
print(f"Estimate: {res_lbfgsb.x[0]:.6f}")
print("No warnings -- bounds prevent lambda < 0")
```
:::

## Fix 1: safeguard in the log-likelihood

Return `NA` (R) or `np.inf` (Python) for invalid parameters. The optimizer treats this as "bad region, backtrack".

::: panel-tabset
### R

```{r}
ll_safe <- function(par, x) {
  if (par <= 0) return(NA)
  sum(x) * log(par) - length(x) * log(exp(par) - 1)
}

cat("=== Newton-Raphson with safeguard ===\n")
res_safe <- maxLik(logLik = ll_safe, start = 1, x = x)
cat("Estimate:", round(coef(res_safe), 6), "\n")
cat("No NaN warnings!\n")
```

### Python

```{python}
def neg_ll_safe(par, x):
    lam = par[0]
    if lam <= 0:
        return np.inf
    return -(np.sum(x) * np.log(lam) - len(x) * np.log(np.exp(lam) - 1))

print("=== Nelder-Mead with safeguard ===")
res_safe = minimize(neg_ll_safe, x0=[0.5], args=(x,), method="Nelder-Mead")
print(f"Estimate: {res_safe.x[0]:.6f}")
print("No NaN warnings!")
```
:::

## Fix 2: supply analytical gradient and Hessian

When the optimizer has exact derivatives, it takes precise steps and avoids invalid regions entirely.

::: panel-tabset
### R

```{r}
cat("=== Newton-Raphson with analytical gradient + Hessian ===\n")
res_exact <- maxLik(logLik = ll, start = 1, grad = grad, hess = hess, x = x)
cat("Iterations:", res_exact$iterations, "\n")
cat("Estimate:", round(coef(res_exact), 6), "\n")
cat("SE:", round(stdEr(res_exact), 5), "\n")
cat("No NaN warnings!\n")
```

### Python

```{python}
def neg_hess(par, x):
    lam = par[0]
    return np.array([[np.sum(x) / lam**2 - len(x) * np.exp(lam) / (np.exp(lam) - 1)**2]])

print("=== Newton-CG with analytical gradient + Hessian ===")
res_exact = minimize(neg_ll, x0=[0.5], args=(x,), method="Newton-CG",
                     jac=lambda p, x: neg_grad(p, x),
                     hessp=lambda p, v, x: neg_hess(p, x) @ v)
print(f"Iterations: {res_exact.nit}")
print(f"Estimate: {res_exact.x[0]:.6f}")
print("No NaN warnings!")
```
:::

## Fix 3: use bounded optimization

Constrain $\lambda > 0$ directly.

::: panel-tabset
### R

```{r}
cat("=== L-BFGS-B (bounded) via optim ===\n")
## maxLik does not support L-BFGS-B; use optim directly
res_b2 <- optim(1, function(p) -ll(p, x), method = "L-BFGS-B",
                lower = 1e-8, upper = 50)
cat("Estimate:", round(res_b2$par, 6), "\n")
cat("No NaN warnings -- lambda constrained to [1e-8, 50]\n")
```

### Python

```{python}
print("=== L-BFGS-B (bounded) ===")
res_bounded = minimize(neg_ll, x0=[0.5], args=(x,), method="L-BFGS-B",
                       bounds=[(1e-8, None)])
print(f"Estimate: {res_bounded.x[0]:.6f}")
print("No NaN warnings -- lambda constrained to (0, inf)")
```
:::

## Summary

| Approach | Warnings? | When to use |
|----------|-----------|-------------|
| No safeguard | Yes (harmless) | Quick prototyping; warnings don't affect the result |
| `if (par <= 0) return(NA/Inf)` | No | Simple fix; works with any optimizer |
| Analytical gradient + Hessian | No | Best convergence; recommended for production code |
| Bounded optimization (L-BFGS-B) | No | When parameter has known bounds |

All approaches converge to the same $\hat{\lambda} = 0.3086$. The choice depends on whether you want clean output and faster convergence (analytical derivatives) or just a quick result (safeguard).
