---
title: "Categorical variables in regression"
author: "Maciej Beręsewicz"
format:
  html:
    embed-resources: true
    table-of-contents: true
    number-sections: true
    df-print: kable
editor: source
execute:
  eval: true
  warning: false
  message: false
filters:
  - ripper
ripper:
  output-dir: "extracted"
  script-links-position: "none"
  include-yaml: true
---

```{r setup, echo = FALSE}
JuliaCall::julia_setup(JULIA_HOME = "/Applications/Julia-1.11.app/Contents/Resources/julia/bin/")
knitr::opts_chunk$set(engine.path = list(
  python = "/opt/anaconda3/bin/python3.12",
  julia = "/Applications/Julia-1.11.app/Contents/Resources/julia/bin/"
))
```

# Packages

::: panel-tabset
## R

```{r, eval = FALSE}
install.packages("car")
```

```{r}
library(car)
```

## Python

```{bash, eval = F}
!pip install statsmodels
```

```{python}
import pandas as pd
import numpy as np
import statsmodels.api as sm
import statsmodels.formula.api as smf
```

## Julia

```{julia, eval = F}
using Pkg
Pkg.add(["StatsBase", "CSV", "CategoricalArrays", "GLM"])
```

```{julia}
using StatsBase
using CSV
using CategoricalArrays
using GLM
using Statistics
using DataFrames
```
:::

# Data

Read in the dataset for the lecture and specify column classes (types).

::: panel-tabset
## R

```{r}
df <- read.csv("../../data/polish-jvs.csv", colClasses = c("character", "factor", rep("character", 4), "numeric"))
head(df)
```

Check the levels

```{r}
levels(df$woj)
```

Read in the dataset but without specifying classes/types, so `woj` is treated as a numeric variable.

```{r}
df_wrong <- read.csv("../../data/polish-jvs.csv")
head(df_wrong)
```

## Python

```{python}
df = pd.read_csv("../../data/polish-jvs.csv", dtype={"id": np.int64, "woj":str, "public":str,"size": str, "nace_division": str, "nace": str})
df.head()
```

Read in the dataset but without specifying classes/types, so `woj` is treated as a numeric variable.

```{python}
df_wrong = pd.read_csv("../../data/polish-jvs.csv")
df_wrong.head()
```

## Julia

```{julia}
df=CSV.read("../../data/polish-jvs.csv", DataFrame,
            types = Dict(:id => String, :woj=> String, :public=> String,
                         :size => String, :nace_division => String, :nace => String));
first(df, 5)
```

Read in the dataset but without specifying classes/types, so `woj` is treated as a numeric variable.

```{julia}
df_wrong=CSV.read("../../data/polish-jvs.csv", DataFrame);
first(df_wrong, 5)
```
:::

# Categorical variables in regression

Consider the following model that tries to explain level of `vacancies`
using `woj`. If `woj` would be treated as continuous than the postulated
model has the following form

$$
\text{vacancies} = \beta_0 + \beta_1\text{woj} + \epsilon.
$$ However, `woj` is categorical, so this model is incorrect. We should
consider the following model with *dummy variables*

$$
\text{vacancies} = \beta_0 + \beta_1I(\text{woj='04'})+ \beta_2I(\text{woj='06'}) + \cdots + \beta_{15}I(\text{woj='32'}) + \epsilon.
$$ Note that code `02` is missing, as it is the reference level.

::: panel-tabset
## R

```{r}
model1 <- lm(vacancies ~ woj, data = df_wrong)
summary(model1)
```

```{r}
model2 <- lm(vacancies ~ woj, data = df)
summary(model2)
```

Compare this model with the following summary. Why they are equal?
because OLS

```{r}
tab1 <- aggregate(vacancies ~ woj, data = df, FUN = mean)
tab1$coef_by_hand <- with(tab1, vacancies-vacancies[1])
tab1$coef <- coef(model2)
head(tab1)
```

```{r}
model3 <- lm(vacancies ~ relevel(woj, ref = "30"), data = df)
summary(model3)
```

```{r}
model4 <- lm(formula = vacancies ~ woj, data = df, contrasts = list(woj = contr.sum))
summary(model4)
```

0.82902 -- average of averages

```{r}
mean(tab1$vacancies)
```

## Python

```{python}
model1 = smf.ols("vacancies ~ woj", data = df_wrong).fit()
print(model1.summary())
```

```{python}
model2 = smf.ols("vacancies ~ woj", data = df).fit()
print(model2.summary())
```

```{python}
model3 = smf.ols("vacancies ~ C(woj, Treatment(reference='30'))", data = df).fit()
print(model3.summary())
```

```{python}
model4 = smf.ols("vacancies ~ C(woj, Sum)", data = df).fit()
print(model4.summary())
```

## Julia

```{julia}
model1 = lm(@formula(vacancies ~ woj), df_wrong)
```

```{julia}
model2 = lm(@formula(vacancies ~ woj), df)
```

```{julia}
model3 = lm(@formula(vacancies ~ woj), df, contrasts = Dict(:woj => DummyCoding(base="30")))
```

```{julia}
model4 = lm(@formula(vacancies ~ woj), df, contrasts = Dict(:woj => EffectsCoding()))
```
:::

## Contrast coding for `size` (ordinal variable)

The `size` variable has 3 levels: Large, Medium, Small. We demonstrate three
coding schemes and how each changes the interpretation. For a comprehensive
treatment, see Schad et al. (2020), *How to capitalize on a priori contrasts
in linear (mixed) models: A tutorial*, Journal of Memory and Language, 110.

### Treatment coding (default)

With treatment coding, Large is the reference level (alphabetical order).
Each coefficient represents the difference from the reference group mean.

::: panel-tabset
## R

```{r contrast-treat-r}
df$size <- factor(df$size)
contrasts(df$size)  ## default treatment contrasts
model_treat <- lm(vacancies ~ size, data = df)
summary(model_treat)

## Verify: coefficients equal group mean differences
group_means_size <- aggregate(vacancies ~ size, data = df, FUN = mean)
group_means_size
```

## Python

```{python contrast-treat-py}
model_treat = smf.ols("vacancies ~ C(size, Treatment)", data=df).fit()
print(model_treat.summary())

## Group means
print(df.groupby("size")["vacancies"].mean())
```

## Julia

```{julia contrast-treat-jl}
model_treat = lm(@formula(vacancies ~ size), df)  ## default DummyCoding
```
:::

### Sum (effect) coding

With sum coding, the intercept equals the grand mean of group means.
Each coefficient is the deviation of that group from the grand mean.
The omitted level's coefficient equals the negative sum of the shown coefficients.

::: panel-tabset
## R

```{r contrast-sum-r}
model_sum_size <- lm(vacancies ~ size, data = df, contrasts = list(size = contr.sum))
summary(model_sum_size)

## Intercept should equal the grand mean of group means
cat("Grand mean of group means:", mean(group_means_size$vacancies), "\n")
```

## Python

```{python contrast-sum-py}
model_sum_size = smf.ols("vacancies ~ C(size, Sum)", data=df).fit()
print(model_sum_size.summary())
```

## Julia

```{julia contrast-sum-jl}
model_sum_size = lm(@formula(vacancies ~ size), df, contrasts = Dict(:size => EffectsCoding()))
```
:::

### Polynomial coding (ordinal)

For ordinal variables with equally spaced levels, polynomial contrasts decompose
the effect into linear and quadratic (and higher) trends. The linear contrast
tests whether there is a monotonic increase/decrease. The quadratic tests for
curvature.

::: panel-tabset
## R

```{r contrast-poly-r}
## Create ordered factor: Small < Medium < Large
df$size_ord <- ordered(df$size, levels = c("Small", "Medium", "Large"))
model_poly <- lm(vacancies ~ size_ord, data = df)
summary(model_poly)
```

The linear coefficient (`.L`) is positive and significant: vacancies increase
with firm size. The quadratic coefficient (`.Q`) is also significant, indicating
the relationship is not purely linear.

## Python

```{python contrast-poly-py}
## Polynomial coding requires specifying the order
from patsy import Poly
model_poly = smf.ols("vacancies ~ C(size, Poly)", data=df).fit()
print(model_poly.summary())
```

## Julia

```{julia contrast-poly-jl}
## Julia does not have built-in polynomial coding in StatsModels.jl
## We can construct the contrast matrix manually for 3 ordered levels
using StatsModels
poly_contrasts = [-1/sqrt(2) 1/sqrt(6); 0 -2/sqrt(6); 1/sqrt(2) 1/sqrt(6)]
## Or use HelmertCoding as an alternative for ordered contrasts
model_helm = lm(@formula(vacancies ~ size), df, contrasts = Dict(:size => HelmertCoding()))
```
:::

# Interpreting interactions between two categorical variables

## The model

Consider the model `vacancies ~ size * public`, which expands to
`vacancies ~ size + public + size:public`. With `size` (3 levels: Large, Medium, Small
— alphabetical order, so Large is the reference) and `public` (2 levels: 0 as reference, 1), we get 6 parameters:

- $\beta_0$ — expected vacancies for Large, private (the baseline cell)
- $\beta_{\text{Medium}}$ — difference Medium vs Large **among private firms**
- $\beta_{\text{Small}}$ — difference Small vs Large **among private firms**
- $\beta_{\text{public}}$ — difference public vs private **among Large firms**
- $\beta_{\text{Medium:public}}$ — how much the public-vs-private difference changes for Medium relative to Large (departure from additivity)
- $\beta_{\text{Small:public}}$ — same for Small relative to Large

**Key insight:** "main effects" are conditional on the reference level of the other variable, not averaged across groups. The interaction terms are difference-in-differences.

## Fit the interaction model

::: panel-tabset
## R

```{r}
model_int1 <- lm(vacancies ~ size * public, data = df)
summary(model_int1)
```

## Python

```{python}
model_int1 = smf.ols("vacancies ~ size * public", data = df).fit()
print(model_int1.summary())
```

## Julia

```{julia}
model_int1 = lm(@formula(vacancies ~ size * public), df)
```
:::

## Verify with cell means

Compute the 6 cell means directly and confirm they match the coefficients:

::: panel-tabset
## R

```{r}
## Cell means
cell_means <- aggregate(vacancies ~ size + public, data = df, FUN = mean)
cell_means

## Reconstruct from coefficients
## Note: R uses alphabetical order, so Large is the reference level
b <- coef(model_int1)
cat("Large, private:", b[1], "\n")
cat("Medium, private:", b[1] + b[2], "\n")
cat("Small, private:", b[1] + b[3], "\n")
cat("Large, public:", b[1] + b[4], "\n")
cat("Medium, public:", b[1] + b[2] + b[4] + b[5], "\n")
cat("Small, public:", b[1] + b[3] + b[4] + b[6], "\n")
```

## Python

```{python}
## Cell means
cell_means = df.groupby(["size", "public"])["vacancies"].mean()
print(cell_means)

## Reconstruct from coefficients
## Note: Python/statsmodels uses alphabetical order, so Large is the reference level
b = model_int1.params
print(f"Large, private: {b['Intercept']:.4f}")
print(f"Medium, private: {b['Intercept'] + b['size[T.Medium]']:.4f}")
print(f"Small, private: {b['Intercept'] + b['size[T.Small]']:.4f}")
print(f"Large, public: {b['Intercept'] + b['public[T.1]']:.4f}")
print(f"Medium, public: {b['Intercept'] + b['size[T.Medium]'] + b['public[T.1]'] + b['size[T.Medium]:public[T.1]']:.4f}")
print(f"Small, public: {b['Intercept'] + b['size[T.Small]'] + b['public[T.1]'] + b['size[T.Small]:public[T.1]']:.4f}")
```

## Julia

```{julia}
## Cell means
cell_means = combine(groupby(df, [:size, :public]), :vacancies => mean => :mean_vac)
println(cell_means)

## Reconstruct from coefficients
## Note: Julia also uses alphabetical order, so Large is the reference level
b = coef(model_int1)
println("Large, private: ", b[1])
println("Medium, private: ", b[1] + b[2])
println("Small, private: ", b[1] + b[3])
println("Large, public: ", b[1] + b[4])
println("Medium, public: ", b[1] + b[2] + b[4] + b[5])
println("Small, public: ", b[1] + b[3] + b[4] + b[6])
```
:::

# ANOVA Type I, II, III

When a categorical predictor has $k$ levels, the model includes $k-1$ dummy variables. To test whether the variable **as a whole** is significant, we need a joint test of all $k-1$ coefficients. This is what ANOVA F-tests do. Three types of Sum of Squares (SS) decomposition exist, and they differ in what they adjust for.

## Type I (Sequential SS)

SS for each term is computed **in the order it appears** in the formula:

- $\text{SS}(A)$, $\text{SS}(B \mid A)$, $\text{SS}(A{:}B \mid A, B)$

The result is **order-dependent**: changing the formula order changes the SS for main effects. This is the default in R's `anova()`. Appropriate for balanced designs or when the order is theoretically motivated.

::: panel-tabset
## R

```{r}
## Type I (sequential) -- note: order matters!
anova(model_int1)
```

## Python

```{python}
## Type I (sequential)
print(sm.stats.anova_lm(model_int1, typ=1))
```

## Julia

```{julia}
## Type I via nested model comparison
model_null = lm(@formula(vacancies ~ 1), df)
model_size = lm(@formula(vacancies ~ size), df)
model_add  = lm(@formula(vacancies ~ size + public), df)
ftest(model_null.model, model_size.model, model_add.model, model_int1.model)
```
:::

## Type II SS

Each term is tested after all other terms of **equal or lower order**:

- $\text{SS}(A \mid B)$, $\text{SS}(B \mid A)$, $\text{SS}(A{:}B \mid A, B)$

Type II respects the **marginality principle** — main effects are not adjusted for
interactions. Order-independent. Recommended when there is no significant interaction.

::: panel-tabset
## R

```{r}
## Type II
Anova(model_int1, type = "II")
```

## Python

```{python}
## Type II
print(sm.stats.anova_lm(model_int1, typ=2))
```

## Julia

```{julia}
## Type II via manual model comparisons
## SS(size | public): compare (public) vs (size + public)
model_public = lm(@formula(vacancies ~ public), df)
ftest(model_public.model, model_add.model)

## SS(public | size): compare (size) vs (size + public)
ftest(model_size.model, model_add.model)

## SS(size:public | size, public): compare (size + public) vs (size * public)
ftest(model_add.model, model_int1.model)
```
:::

## Type III SS

Each term is tested after **all** other terms, **including** interactions:

- $\text{SS}(A \mid B, A{:}B)$, $\text{SS}(B \mid A, A{:}B)$, $\text{SS}(A{:}B \mid A, B)$

With treatment (dummy) coding, Type III tests the main effect of $A$ **at the reference level** of $B$. For main effects interpretable at the grand mean, use **sum contrasts**.

::: panel-tabset
## R

```{r}
## Type III -- with treatment contrasts (default)
Anova(model_int1, type = "III")

## Type III with sum contrasts for interpretability
df_sum <- df
df_sum$size <- factor(df_sum$size)
df_sum$public <- factor(df_sum$public)
contrasts(df_sum$size) <- contr.sum(3)
contrasts(df_sum$public) <- contr.sum(2)
model_int1_sum <- lm(vacancies ~ size * public, data = df_sum)
Anova(model_int1_sum, type = "III")
```

## Python

```{python}
## Type III
print(sm.stats.anova_lm(model_int1, typ=3))

## Type III with sum contrasts
model_int1_sum = smf.ols("vacancies ~ C(size, Sum) * C(public, Sum)", data = df).fit()
print(sm.stats.anova_lm(model_int1_sum, typ=3))
```

## Julia

```{julia}
## Type III via manual model comparisons
## SS(size | public, size:public): compare (public) vs (size * public)
## This tests whether adding size (and the interaction) improves over public alone
## For a proper Type III test, we need the model without size but with public and interaction
## In practice, Type III tests each term "last" -- for main effects in presence of interaction
## this requires dropping the main effect while keeping the interaction, which is unusual.
## A simpler approach: use the F-statistics from the coefficient table directly.
coeftable(model_int1)
```
:::

## Comparison

For **balanced** data (equal cell sizes), all three types give identical results.
For **unbalanced** data (which is typical in observational studies like our JVS data),
the results differ. Type II is generally preferred unless there is a significant
interaction, in which case Type III may be more appropriate (with sum contrasts).

For a detailed discussion of when to use Type II vs Type III, see Langsrud (2003),
*ANOVA for unbalanced data: Use Type II instead of Type III sums of squares*,
Statistics and Computing, 13, 163--167.

## Exercise: ANOVA with multiple covariates

Fit the model `vacancies ~ size * public + woj` and compare ANOVA types.

::: panel-tabset
## R

```{r anova-ex-r}
model_anova_ex <- lm(vacancies ~ size * public + woj, data = df)

## Type I: does the order matter?
cat("Type I -- size first:\n")
anova(lm(vacancies ~ size * public + woj, data = df))
cat("\nType I -- woj first:\n")
anova(lm(vacancies ~ woj + size * public, data = df))

## Type II
cat("\nType II (order-independent):\n")
Anova(model_anova_ex, type = "II")

## Type III with sum contrasts
df_ex <- df
df_ex$size <- factor(df_ex$size)
df_ex$public <- factor(df_ex$public)
contrasts(df_ex$size) <- contr.sum(3)
contrasts(df_ex$public) <- contr.sum(2)
model_anova_ex_sum <- lm(vacancies ~ size * public + woj, data = df_ex)
cat("\nType III (sum contrasts):\n")
Anova(model_anova_ex_sum, type = "III")
```

Now add `nace` to the model:

```{r anova-nace-r}
model_anova_nace <- lm(vacancies ~ size * public + woj + nace, data = df)
cat("Type II with nace:\n")
Anova(model_anova_nace, type = "II")
```

## Python

```{python anova-ex-py}
model_anova_ex = smf.ols("vacancies ~ size * public + woj", data=df).fit()

## Type I: order matters
print("Type I -- size first:")
print(sm.stats.anova_lm(smf.ols("vacancies ~ size * public + woj", data=df).fit(), typ=1))
print("\nType I -- woj first:")
print(sm.stats.anova_lm(smf.ols("vacancies ~ woj + size * public", data=df).fit(), typ=1))

## Type II
print("\nType II:")
print(sm.stats.anova_lm(model_anova_ex, typ=2))

## Type III with sum contrasts
model_anova_ex_sum = smf.ols("vacancies ~ C(size, Sum) * C(public, Sum) + woj", data=df).fit()
print("\nType III (sum contrasts):")
print(sm.stats.anova_lm(model_anova_ex_sum, typ=3))
```

```{python anova-nace-py}
## Add nace
model_anova_nace = smf.ols("vacancies ~ size * public + woj + nace", data=df).fit()
print("Type II with nace:")
print(sm.stats.anova_lm(model_anova_nace, typ=2))
```

## Julia

```{julia anova-ex-jl}
## Type I via nested models (order matters)
m_null = lm(@formula(vacancies ~ 1), df)
m_size = lm(@formula(vacancies ~ size), df)
m_sp = lm(@formula(vacancies ~ size + public), df)
m_spi = lm(@formula(vacancies ~ size * public), df)
m_spiw = lm(@formula(vacancies ~ size * public + woj), df)
println("Type I (sequential):")
ftest(m_null.model, m_size.model, m_sp.model, m_spi.model, m_spiw.model)
```
:::

# Categorical × continuous interaction and scaling

The topics of categorical × continuous interactions, centering, standardisation,
and Gelman's (2008) scaling approach are covered in the supplementary notebook:

**[07a: Interactions and scaling](07a-interactions-scaling.html)**

# Marginal effects

Marginal effects are covered in a separate notebook:

**[08: Marginal effects](08-marginal-effects.html)**

# Exercises

## Exercise 1: What happens when you forget to specify the variable type?

Fit two models with `vacancies` as the dependent variable and `woj` as the
independent variable:

- **Model A**: using `df_wrong` (where `woj` is treated as numeric),
- **Model B**: using `df` (where `woj` is treated as a factor/categorical variable).

Tasks:

1. Compare the number of estimated parameters in each model.
2. Compare $R^2$ values. Which model fits better? Why?
3. Compute the mean of `vacancies` for each `woj` level. Verify that the
   coefficients of Model B equal the difference between each group mean and
   the reference group mean.

::: panel-tabset
## R

```{r}
## Model A -- woj treated as numeric (wrong)
model_a <- lm(vacancies ~ woj, data = df_wrong)
summary(model_a)
```

```{r}
## Model B -- woj treated as factor (correct)
model_b <- lm(vacancies ~ woj, data = df)
summary(model_b)
```

```{r}
## Compare R-squared
cat("R-squared (numeric):", summary(model_a)$r.squared, "\n")
cat("R-squared (factor): ", summary(model_b)$r.squared, "\n")
```

```{r}
## Verify coefficients equal group mean differences
group_means <- aggregate(vacancies ~ woj, data = df, FUN = mean)
group_means$diff_from_ref <- group_means$vacancies - group_means$vacancies[1]
group_means$coef <- coef(model_b)
head(group_means)
```

## Python

```{python}
## Model A -- woj treated as numeric (wrong)
model_a = smf.ols("vacancies ~ woj", data = df_wrong).fit()
print(f"R-squared (numeric): {model_a.rsquared:.6f}")
print(f"Number of parameters: {len(model_a.params)}")
```

```{python}
## Model B -- woj treated as factor (correct)
model_b = smf.ols("vacancies ~ woj", data = df).fit()
print(f"R-squared (factor): {model_b.rsquared:.6f}")
print(f"Number of parameters: {len(model_b.params)}")
```

```{python}
## Verify coefficients equal group mean differences
group_means = df.groupby("woj")["vacancies"].mean()
ref_mean = group_means.iloc[0]
comparison = pd.DataFrame({
    "group_mean": group_means,
    "diff_from_ref": group_means - ref_mean,
    "coef": model_b.params.values
})
comparison.head()
```

## Julia

```{julia}
## Model A -- woj treated as numeric (wrong)
model_a = lm(@formula(vacancies ~ woj), df_wrong)
```

```{julia}
## Model B -- woj treated as factor (correct)
model_b = lm(@formula(vacancies ~ woj), df)
```

```{julia}
## Compare R-squared
println("R-squared (numeric): ", r2(model_a))
println("R-squared (factor):  ", r2(model_b))
```
:::

