---
title: "Categorical variables in regression and marginal effects"
author: "Maciej Beręsewicz"
format:
  html:
    embed-resources: true
    table-of-contents: true
    number-sections: true
    df-print: kable
editor: source
execute:
  eval: true
  warning: false
  message: false
filters:
  - ripper
ripper:
  output-dir: "extracted"
  script-links-position: "none"
  include-yaml: true
---

```{r setup, echo = FALSE}
JuliaCall::julia_setup(JULIA_HOME = "/Applications/Julia-1.11.app/Contents/Resources/julia/bin/")
knitr::opts_chunk$set(engine.path = list(
  python = "/opt/anaconda3/bin/python3.12",
  julia = "/Applications/Julia-1.11.app/Contents/Resources/julia/bin/"
))
```

# Packages

::: panel-tabset
## R

Install `marginaleffects` package if necessary.

```{r, eval = FALSE}
install.packages("marginaleffects")
install.packages("car")
```

```{r}
library(marginaleffects)
library(car)
```

## Python

```{bash, eval = F}
!pip install statsmodels
!pip install marginaleffects
```

```{python}
import pandas as pd
import numpy as np
import statsmodels.api as sm
import statsmodels.formula.api as smf
from marginaleffects import *
```

## Julia

```{julia, eval = F}
using Pkg
Pkg.add(["Effects", "StatsBase", "CSV", "CategoricalArrays", "GLM"])
```

```{julia}
using Effects
using StatsBase
using CSV
using CategoricalArrays
using GLM
using Statistics
using DataFrames
```
:::

# Data

Read in the dataset for the lecture and specify column classes (types).

::: panel-tabset
## R

```{r}
df <- read.csv("../../data/polish-jvs.csv", colClasses = c("character", "factor", rep("character", 4), "numeric"))
head(df)
```

Check the levels

```{r}
levels(df$woj)
```

Read in the dataset but without specifying classes/types, so `woj` is treated as a numeric variable.

```{r}
df_wrong <- read.csv("../../data/polish-jvs.csv")
head(df_wrong)
```

## Python

```{python}
df = pd.read_csv("../../data/polish-jvs.csv", dtype={"id": np.int64, "woj":str, "public":str,"size": str, "nace_division": str, "nace": str})
df.head()
```

Read in the dataset but without specifying classes/types, so `woj` is treated as a numeric variable.

```{python}
df_wrong = pd.read_csv("../../data/polish-jvs.csv")
df_wrong.head()
```

## Julia

```{julia}
df=CSV.read("../../data/polish-jvs.csv", DataFrame,
            types = Dict(:id => String, :woj=> String, :public=> String,
                         :size => String, :nace_division => String, :nace => String));
first(df, 5)
```

Read in the dataset but without specifying classes/types, so `woj` is treated as a numeric variable.

```{julia}
df_wrong=CSV.read("../../data/polish-jvs.csv", DataFrame);
first(df_wrong, 5)
```
:::

# Categorical variables in regression

Consider the following model that tries to explain level of `vacancies`
using `woj`. If `woj` would be treated as continuous than the postulated
model has the following form

$$
\text{vacancies} = \beta_0 + \beta_1\text{woj} + \epsilon.
$$ However, `woj` is categorical, so this model is incorrect. We should
consider the following model with *dummy variables*

$$
\text{vacancies} = \beta_0 + \beta_1I(\text{woj='04'})+ \beta_2I(\text{woj='06'}) + \cdots + \beta_{15}I(\text{woj='32'}) + \epsilon.
$$ Note that code `02` is missing, as it is the reference level.

::: panel-tabset
## R

```{r}
model1 <- lm(vacancies ~ woj, data = df_wrong)
summary(model1)
```

```{r}
model2 <- lm(vacancies ~ woj, data = df)
summary(model2)
```

Compare this model with the following summary. Why they are equal?
because OLS

```{r}
tab1 <- aggregate(vacancies ~ woj, data = df, FUN = mean)
tab1$coef_by_hand <- with(tab1, vacancies-vacancies[1])
tab1$coef <- coef(model2)
head(tab1)
```

```{r}
model3 <- lm(vacancies ~ relevel(woj, ref = "30"), data = df)
summary(model3)
```

```{r}
model4 <- lm(formula = vacancies ~ woj, data = df, contrasts = list(woj = contr.sum))
summary(model4)
```

0.82902 -- average of averages

```{r}
mean(tab1$vacancies)
```

## Python

```{python}
model1 = smf.ols("vacancies ~ woj", data = df_wrong).fit()
print(model1.summary())
```

```{python}
model2 = smf.ols("vacancies ~ woj", data = df).fit()
print(model2.summary())
```

```{python}
model3 = smf.ols("vacancies ~ C(woj, Treatment(reference='30'))", data = df).fit()
print(model3.summary())
```

```{python}
model4 = smf.ols("vacancies ~ C(woj, Sum)", data = df).fit()
print(model4.summary())
```

## Julia

```{julia}
model1 = lm(@formula(vacancies ~ woj), df_wrong)
```

```{julia}
model2 = lm(@formula(vacancies ~ woj), df)
```

```{julia}
model3 = lm(@formula(vacancies ~ woj), df, contrasts = Dict(:woj => DummyCoding(base="30")))
```

```{julia}
model4 = lm(@formula(vacancies ~ woj), df, contrasts = Dict(:woj => EffectsCoding()))
```
:::

# Interpreting interactions between two categorical variables

## The model

Consider the model `vacancies ~ size * public`, which expands to
`vacancies ~ size + public + size:public`. With `size` (3 levels: Large, Medium, Small
— alphabetical order, so Large is the reference) and `public` (2 levels: 0 as reference, 1), we get 6 parameters:

- $\beta_0$ — expected vacancies for Large, private (the baseline cell)
- $\beta_{\text{Medium}}$ — difference Medium vs Large **among private firms**
- $\beta_{\text{Small}}$ — difference Small vs Large **among private firms**
- $\beta_{\text{public}}$ — difference public vs private **among Large firms**
- $\beta_{\text{Medium:public}}$ — how much the public-vs-private difference changes for Medium relative to Large (departure from additivity)
- $\beta_{\text{Small:public}}$ — same for Small relative to Large

**Key insight:** "main effects" are conditional on the reference level of the other variable, not averaged across groups. The interaction terms are difference-in-differences.

## Fit the interaction model

::: panel-tabset
## R

```{r}
model_int1 <- lm(vacancies ~ size * public, data = df)
summary(model_int1)
```

## Python

```{python}
model_int1 = smf.ols("vacancies ~ size * public", data = df).fit()
print(model_int1.summary())
```

## Julia

```{julia}
model_int1 = lm(@formula(vacancies ~ size * public), df)
```
:::

## Verify with cell means

Compute the 6 cell means directly and confirm they match the coefficients:

::: panel-tabset
## R

```{r}
## Cell means
cell_means <- aggregate(vacancies ~ size + public, data = df, FUN = mean)
cell_means

## Reconstruct from coefficients
## Note: R uses alphabetical order, so Large is the reference level
b <- coef(model_int1)
cat("Large, private:", b[1], "\n")
cat("Medium, private:", b[1] + b[2], "\n")
cat("Small, private:", b[1] + b[3], "\n")
cat("Large, public:", b[1] + b[4], "\n")
cat("Medium, public:", b[1] + b[2] + b[4] + b[5], "\n")
cat("Small, public:", b[1] + b[3] + b[4] + b[6], "\n")
```

## Python

```{python}
## Cell means
cell_means = df.groupby(["size", "public"])["vacancies"].mean()
print(cell_means)

## Reconstruct from coefficients
## Note: Python/statsmodels uses alphabetical order, so Large is the reference level
b = model_int1.params
print(f"Large, private: {b['Intercept']:.4f}")
print(f"Medium, private: {b['Intercept'] + b['size[T.Medium]']:.4f}")
print(f"Small, private: {b['Intercept'] + b['size[T.Small]']:.4f}")
print(f"Large, public: {b['Intercept'] + b['public[T.1]']:.4f}")
print(f"Medium, public: {b['Intercept'] + b['size[T.Medium]'] + b['public[T.1]'] + b['size[T.Medium]:public[T.1]']:.4f}")
print(f"Small, public: {b['Intercept'] + b['size[T.Small]'] + b['public[T.1]'] + b['size[T.Small]:public[T.1]']:.4f}")
```

## Julia

```{julia}
## Cell means
cell_means = combine(groupby(df, [:size, :public]), :vacancies => mean => :mean_vac)
println(cell_means)

## Reconstruct from coefficients
## Note: Julia also uses alphabetical order, so Large is the reference level
b = coef(model_int1)
println("Large, private: ", b[1])
println("Medium, private: ", b[1] + b[2])
println("Small, private: ", b[1] + b[3])
println("Large, public: ", b[1] + b[4])
println("Medium, public: ", b[1] + b[2] + b[4] + b[5])
println("Small, public: ", b[1] + b[3] + b[4] + b[6])
```
:::

# ANOVA Type I, II, III

When a categorical predictor has $k$ levels, the model includes $k-1$ dummy variables. To test whether the variable **as a whole** is significant, we need a joint test of all $k-1$ coefficients. This is what ANOVA F-tests do. Three types of Sum of Squares (SS) decomposition exist, and they differ in what they adjust for.

## Type I (Sequential SS)

SS for each term is computed **in the order it appears** in the formula:

- $\text{SS}(A)$, $\text{SS}(B \mid A)$, $\text{SS}(A{:}B \mid A, B)$

The result is **order-dependent**: changing the formula order changes the SS for main effects. This is the default in R's `anova()`. Appropriate for balanced designs or when the order is theoretically motivated.

::: panel-tabset
## R

```{r}
## Type I (sequential) -- note: order matters!
anova(model_int1)
```

## Python

```{python}
## Type I (sequential)
print(sm.stats.anova_lm(model_int1, typ=1))
```

## Julia

```{julia}
## Type I via nested model comparison
model_null = lm(@formula(vacancies ~ 1), df)
model_size = lm(@formula(vacancies ~ size), df)
model_add  = lm(@formula(vacancies ~ size + public), df)
ftest(model_null.model, model_size.model, model_add.model, model_int1.model)
```
:::

## Type II SS

Each term is tested after all other terms of **equal or lower order**:

- $\text{SS}(A \mid B)$, $\text{SS}(B \mid A)$, $\text{SS}(A{:}B \mid A, B)$

Type II respects the **marginality principle** — main effects are not adjusted for
interactions. Order-independent. Recommended when there is no significant interaction.

::: panel-tabset
## R

```{r}
## Type II
Anova(model_int1, type = "II")
```

## Python

```{python}
## Type II
print(sm.stats.anova_lm(model_int1, typ=2))
```

## Julia

```{julia}
## Type II via manual model comparisons
## SS(size | public): compare (public) vs (size + public)
model_public = lm(@formula(vacancies ~ public), df)
ftest(model_public.model, model_add.model)

## SS(public | size): compare (size) vs (size + public)
ftest(model_size.model, model_add.model)

## SS(size:public | size, public): compare (size + public) vs (size * public)
ftest(model_add.model, model_int1.model)
```
:::

## Type III SS

Each term is tested after **all** other terms, **including** interactions:

- $\text{SS}(A \mid B, A{:}B)$, $\text{SS}(B \mid A, A{:}B)$, $\text{SS}(A{:}B \mid A, B)$

With treatment (dummy) coding, Type III tests the main effect of $A$ **at the reference level** of $B$. For main effects interpretable at the grand mean, use **sum contrasts**.

::: panel-tabset
## R

```{r}
## Type III -- with treatment contrasts (default)
Anova(model_int1, type = "III")

## Type III with sum contrasts for interpretability
df_sum <- df
contrasts(df_sum$size) <- contr.sum(3)
contrasts(df_sum$public) <- contr.sum(2)
model_int1_sum <- lm(vacancies ~ size * public, data = df_sum)
Anova(model_int1_sum, type = "III")
```

## Python

```{python}
## Type III
print(sm.stats.anova_lm(model_int1, typ=3))

## Type III with sum contrasts
model_int1_sum = smf.ols("vacancies ~ C(size, Sum) * C(public, Sum)", data = df).fit()
print(sm.stats.anova_lm(model_int1_sum, typ=3))
```

## Julia

```{julia}
## Type III via manual model comparisons
## SS(size | public, size:public): compare (public) vs (size * public)
## This tests whether adding size (and the interaction) improves over public alone
## For a proper Type III test, we need the model without size but with public and interaction
## In practice, Type III tests each term "last" -- for main effects in presence of interaction
## this requires dropping the main effect while keeping the interaction, which is unusual.
## A simpler approach: use the F-statistics from the coefficient table directly.
coeftable(model_int1)
```
:::

## Comparison

For **balanced** data (equal cell sizes), all three types give identical results.
For **unbalanced** data (which is typical in observational studies like our JVS data),
the results differ. Type II is generally preferred unless there is a significant
interaction, in which case Type III may be more appropriate (with sum contrasts).

# Categorical × continuous interaction

## Creating a continuous predictor

Our dataset has only categorical predictors. For this example, we create a continuous
variable `n_firms` — the number of observations (firms) per `nace_division`, a proxy
for the size of the economic sector.

::: panel-tabset
## R

```{r}
nace_counts <- as.data.frame(table(df$nace_division))
names(nace_counts) <- c("nace_division", "n_firms")
df <- merge(df, nace_counts, by = "nace_division")
df$n_firms <- as.numeric(df$n_firms)
head(df[, c("nace_division", "size", "n_firms", "vacancies")])
```

## Python

```{python}
nace_counts = df.groupby("nace_division").size().reset_index(name="n_firms")
df = df.merge(nace_counts, on="nace_division")
print(df[["nace_division", "size", "n_firms", "vacancies"]].head())
```

## Julia

```{julia}
nace_counts = combine(groupby(df, :nace_division), nrow => :n_firms)
df = leftjoin(df, nace_counts, on = :nace_division)
first(df[:, [:nace_division, :size, :n_firms, :vacancies]], 5)
```
:::

## Model without interaction (parallel lines)

The model `vacancies ~ size + n_firms` assumes the same slope for `n_firms` across
all size groups — only the intercept differs. The lines are parallel.

::: panel-tabset
## R

```{r}
model_par <- lm(vacancies ~ size + n_firms, data = df)
summary(model_par)
```

## Python

```{python}
model_par = smf.ols("vacancies ~ size + n_firms", data = df).fit()
print(model_par.summary())
```

## Julia

```{julia}
model_par = lm(@formula(vacancies ~ size + n_firms), df)
```
:::

## Model with interaction (different slopes)

The model `vacancies ~ size * n_firms` allows a different slope for `n_firms` in each
size group. The coefficient of `n_firms` is the slope for the reference group (Large).
The interaction terms measure how much the slope differs for other groups.

::: panel-tabset
## R

```{r}
model_slopes <- lm(vacancies ~ size * n_firms, data = df)
summary(model_slopes)

## Slopes by group
b <- coef(model_slopes)
cat("Slope for Large:", b["n_firms"], "\n")
cat("Slope for Medium:", b["n_firms"] + b["sizeMedium:n_firms"], "\n")
cat("Slope for Small:", b["n_firms"] + b["sizeSmall:n_firms"], "\n")
```

## Python

```{python}
model_slopes = smf.ols("vacancies ~ size * n_firms", data = df).fit()
print(model_slopes.summary())

## Slopes by group
b = model_slopes.params
print(f"Slope for Large: {b['n_firms']:.6f}")
print(f"Slope for Medium: {b['n_firms'] + b['size[T.Medium]:n_firms']:.6f}")
print(f"Slope for Small: {b['n_firms'] + b['size[T.Small]:n_firms']:.6f}")
```

## Julia

```{julia}
model_slopes = lm(@formula(vacancies ~ size * n_firms), df)

## Slopes by group
b = coef(model_slopes)
println("Slope for Large: ", b[4])
println("Slope for Medium: ", b[4] + b[5])
println("Slope for Small: ", b[4] + b[6])
```
:::

# Centering continuous variables

## Why center?

In a model with interaction `vacancies ~ size * n_firms`, the "main effect" of `size`
is the effect **when `n_firms = 0`**. But `n_firms = 0` is outside the observed range
(there are no sectors with zero firms), so this main effect is an extrapolation with
no practical meaning.

**Grand-mean centering** replaces `n_firms` with `n_firms - mean(n_firms)`. Now the
main effect of `size` is the effect **at the average sector size**, which is much more
interpretable.

## Grand-mean centering

::: panel-tabset
## R

```{r}
df$n_firms_c <- df$n_firms - mean(df$n_firms)

model_cent <- lm(vacancies ~ size * n_firms_c, data = df)
summary(model_cent)
```

## Python

```{python}
df["n_firms_c"] = df["n_firms"] - df["n_firms"].mean()

model_cent = smf.ols("vacancies ~ size * n_firms_c", data = df).fit()
print(model_cent.summary())
```

## Julia

```{julia}
df.n_firms_c = df.n_firms .- mean(df.n_firms)

model_cent = lm(@formula(vacancies ~ size * n_firms_c), df)
```
:::

## What changes and what stays the same?

::: panel-tabset
## R

```{r}
## R-squared: identical
cat("R² (uncentered):", summary(model_slopes)$r.squared, "\n")
cat("R² (centered):  ", summary(model_cent)$r.squared, "\n")

## Interaction coefficients: identical
cat("\nInteraction coefs (uncentered):\n")
print(coef(model_slopes)[5:6])
cat("Interaction coefs (centered):\n")
print(coef(model_cent)[5:6])

## Main-effect coefficients: different
cat("\nMain effects (uncentered):\n")
print(coef(model_slopes)[1:3])
cat("Main effects (centered):\n")
print(coef(model_cent)[1:3])

## Fitted values: identical
cat("\nFitted values identical?", all.equal(fitted(model_slopes), fitted(model_cent)), "\n")
```

## Python

```{python}
## R-squared: identical
print(f"R² (uncentered): {model_slopes.rsquared:.6f}")
print(f"R² (centered):   {model_cent.rsquared:.6f}")

## Interaction coefficients: identical
print(f"\nInteraction coefs (uncentered):")
print(model_slopes.params.filter(like=":"))
print(f"Interaction coefs (centered):")
print(model_cent.params.filter(like=":"))

## Main-effect coefficients: different
print(f"\nMain effects (uncentered):")
print(model_slopes.params.iloc[:3])
print(f"Main effects (centered):")
print(model_cent.params.iloc[:3])
```

## Julia

```{julia}
## R-squared: identical
println("R² (uncentered): ", r2(model_slopes))
println("R² (centered):   ", r2(model_cent))

## Interaction coefficients: identical
println("\nInteraction coefs (uncentered): ", coef(model_slopes)[5:6])
println("Interaction coefs (centered):   ", coef(model_cent)[5:6])

## Main-effect coefficients: different
println("\nMain effects (uncentered): ", coef(model_slopes)[1:3])
println("Main effects (centered):   ", coef(model_cent)[1:3])
```
:::

## Standardisation

Standardising (`z-score`) additionally rescales to standard deviation units, so
coefficients become comparable across variables with different scales.

::: panel-tabset
## R

```{r}
df$n_firms_s <- as.numeric(scale(df$n_firms))

model_std <- lm(vacancies ~ size * n_firms_s, data = df)
summary(model_std)
```

## Python

```{python}
df["n_firms_s"] = (df["n_firms"] - df["n_firms"].mean()) / df["n_firms"].std()

model_std = smf.ols("vacancies ~ size * n_firms_s", data = df).fit()
print(model_std.summary())
```

## Julia

```{julia}
df.n_firms_s = (df.n_firms .- mean(df.n_firms)) ./ std(df.n_firms)

model_std = lm(@formula(vacancies ~ size * n_firms_s), df)
```
:::

# Marginal effects

## Why marginal effects?

In models with interactions or polynomial terms, the effect of a variable on the
response depends on the values of other variables. The regression coefficients
alone no longer have a straightforward interpretation. Marginal effects provide
a way to summarise the effect of each variable in a meaningful way.

## Model with interactions

Consider the following model

$$
\text{vacancies} = \beta_0 + \boldsymbol{\beta}_{\text{woj}} + \boldsymbol{\beta}_{\text{size}} + \beta_{\text{public}} + \boldsymbol{\beta}_{\text{woj} \times \text{size}} + \boldsymbol{\beta}_{\text{woj} \times \text{public}} + \varepsilon,
$$

where $\times$ denotes interactions between two variables.

::: panel-tabset
## R

```{r}
model5 <- lm(vacancies ~ woj*size + woj*public, df)
summary(model5)
```

Calculate **Average Marginal Effects** (AME) -- the marginal effect is computed
for each observation, then averaged across the sample.

```{r}
avg_slopes(model5)
```

Calculate **Marginal Effects at the Mean** (MEM) -- the marginal effect evaluated
at the mean values of all other variables.

```{r}
avg_slopes(model5, newdata = "mean")
```

## Python

```{python}
model5 = smf.ols("vacancies ~ woj*size + woj*public", data = df).fit()
print(model5.summary())
```

Calculate **Average Marginal Effects** (AME).

```{python}
model5_ame = avg_slopes(model5)
model5_ame.to_pandas().head()
```

Calculate **Marginal Effects at the Mean** (MEM).

```{python}
model5_mem = avg_slopes(model5, newdata="mean")
model5_mem.to_pandas().head()
```

## Julia

```{julia}
model5 = lm(@formula(vacancies ~ woj*size + woj*public), df)
```

`Effects.jl` returns marginal effects for each level separately, not as
deviations from the reference level.

```{julia}
vcat(
  effects(Dict(:woj => sort(unique(df.woj))), model5),
  effects(Dict(:public => sort(unique(df.public))), model5),
  effects(Dict(:size => sort(unique(df.size))), model5),
  cols = :union
)
```
:::

## AME by subgroups

We can compute AME of one variable separately for subgroups defined by another
variable. For example, does the effect of `size` differ between public and
private firms?

::: panel-tabset
## R

```{r}
avg_slopes(model5, variables = "size", by = "public")
```

## Python

```{python}
avg_slopes(model5, variables = "size", by = "public").to_pandas()
```

## Julia

```{julia}
effects(Dict(:size => sort(unique(df.size))), model5)
```
:::

## Marginal Effects at Representative Values (MERV)

We can also compute fitted values at specific combinations of variable
values. For example, expected vacancies for each voivodeship, holding
`size = "Medium"` and `public = "0"`.

::: panel-tabset
## R

```{r}
predictions(model5,
            newdata = datagrid(woj = unique(df$woj), size = "Medium", public = "0"))
```

## Python

```{python}
from marginaleffects import predictions, datagrid
pred = predictions(model5,
                   newdata = datagrid(model5, woj = df["woj"].unique(), size = "Medium", public = "0"))
pred.to_pandas().head()
```

## Julia

```{julia}
effects(Dict(:woj => sort(unique(df.woj)),
             :size => ["Medium"],
             :public => ["0"]), model5)
```
:::

# Exercises

## Exercise 1: What happens when you forget to specify the variable type?

Fit two models with `vacancies` as the dependent variable and `woj` as the
independent variable:

- **Model A**: using `df_wrong` (where `woj` is treated as numeric),
- **Model B**: using `df` (where `woj` is treated as a factor/categorical variable).

Tasks:

1. Compare the number of estimated parameters in each model.
2. Compare $R^2$ values. Which model fits better? Why?
3. Compute the mean of `vacancies` for each `woj` level. Verify that the
   coefficients of Model B equal the difference between each group mean and
   the reference group mean.

::: panel-tabset
## R

```{r}
## Model A -- woj treated as numeric (wrong)
model_a <- lm(vacancies ~ woj, data = df_wrong)
summary(model_a)
```

```{r}
## Model B -- woj treated as factor (correct)
model_b <- lm(vacancies ~ woj, data = df)
summary(model_b)
```

```{r}
## Compare R-squared
cat("R-squared (numeric):", summary(model_a)$r.squared, "\n")
cat("R-squared (factor): ", summary(model_b)$r.squared, "\n")
```

```{r}
## Verify coefficients equal group mean differences
group_means <- aggregate(vacancies ~ woj, data = df, FUN = mean)
group_means$diff_from_ref <- group_means$vacancies - group_means$vacancies[1]
group_means$coef <- coef(model_b)
head(group_means)
```

## Python

```{python}
## Model A -- woj treated as numeric (wrong)
model_a = smf.ols("vacancies ~ woj", data = df_wrong).fit()
print(f"R-squared (numeric): {model_a.rsquared:.6f}")
print(f"Number of parameters: {len(model_a.params)}")
```

```{python}
## Model B -- woj treated as factor (correct)
model_b = smf.ols("vacancies ~ woj", data = df).fit()
print(f"R-squared (factor): {model_b.rsquared:.6f}")
print(f"Number of parameters: {len(model_b.params)}")
```

```{python}
## Verify coefficients equal group mean differences
group_means = df.groupby("woj")["vacancies"].mean()
ref_mean = group_means.iloc[0]
comparison = pd.DataFrame({
    "group_mean": group_means,
    "diff_from_ref": group_means - ref_mean,
    "coef": model_b.params.values
})
comparison.head()
```

## Julia

```{julia}
## Model A -- woj treated as numeric (wrong)
model_a = lm(@formula(vacancies ~ woj), df_wrong)
```

```{julia}
## Model B -- woj treated as factor (correct)
model_b = lm(@formula(vacancies ~ woj), df)
```

```{julia}
## Compare R-squared
println("R-squared (numeric): ", r2(model_a))
println("R-squared (factor):  ", r2(model_b))
```
:::

## Exercise 2: Average Marginal Effect of firm size

Fit the model: `vacancies ~ size * public + woj`.

Tasks:

1. How many parameters does this model estimate?
2. Compute the AME of `size`. Interpret: what is the average difference in
   expected vacancies between Large and Small firms, averaging over all
   values of `public` and `woj`?
3. Compute the AME of `size` separately for `public = 0` and `public = 1`.
   Does firm size matter more in the public or private sector?

::: panel-tabset
## R

```{r}
model6 <- lm(vacancies ~ size * public + woj, data = df)
summary(model6)
```

```{r}
cat("Number of parameters:", length(coef(model6)), "\n")
```

AME of `size`:

```{r}
avg_slopes(model6, variables = "size")
```

AME of `size` by `public`:

```{r}
avg_slopes(model6, variables = "size", by = "public")
```

## Python

```{python}
model6 = smf.ols("vacancies ~ size * public + woj", data = df).fit()
print(f"Number of parameters: {len(model6.params)}")
```

AME of `size`:

```{python}
avg_slopes(model6, variables = "size").to_pandas()
```

AME of `size` by `public`:

```{python}
avg_slopes(model6, variables = "size", by = "public").to_pandas()
```

## Julia

```{julia}
model6 = lm(@formula(vacancies ~ size * public + woj), df)
println("Number of parameters: ", length(coef(model6)))
```

```{julia}
effects(Dict(:size => sort(unique(df.size))), model6)
```
:::

## Exercise 3: Fitted values across NACE sectors

Fit the model: `vacancies ~ nace + size + public`.

Tasks:

1. Compute the fitted values for each NACE section, holding
   `size = "Medium"` and `public = "0"`.
2. Which 3 sectors have the highest expected vacancies? Which 3 have the
   lowest?
3. Add the interaction `nace * size` and repeat. Do the rankings change
   across size categories?

::: panel-tabset
## R

```{r}
model7 <- lm(vacancies ~ nace + size + public, data = df)
summary(model7)
```

Fitted values at representative values:

```{r}
pred7 <- predictions(model7,
                     newdata = datagrid(nace = unique(df$nace), size = "Medium", public = "0"))
pred7 <- as.data.frame(pred7)
pred7_sorted <- pred7[order(-pred7$estimate), c("nace", "estimate")]
head(pred7_sorted, 3)
tail(pred7_sorted, 3)
```

Now add the interaction `nace * size`:

```{r}
model8 <- lm(vacancies ~ nace * size + public, data = df)
```

```{r}
pred8_medium <- predictions(model8,
                            newdata = datagrid(nace = unique(df$nace), size = "Medium", public = "0"))
pred8_large <- predictions(model8,
                           newdata = datagrid(nace = unique(df$nace), size = "Large", public = "0"))
pred8_medium <- as.data.frame(pred8_medium)
pred8_large <- as.data.frame(pred8_large)
cat("Top 3 NACE (Medium):\n")
head(pred8_medium[order(-pred8_medium$estimate), "nace"], 3)
cat("Top 3 NACE (Large):\n")
head(pred8_large[order(-pred8_large$estimate), "nace"], 3)
```

## Python

```{python}
model7 = smf.ols("vacancies ~ nace + size + public", data = df).fit()
```

```{python}
from marginaleffects import predictions, datagrid
pred7 = predictions(model7,
                    newdata = datagrid(model7, nace = df["nace"].unique(), size = "Medium", public = "0"))
pred7_df = pred7.to_pandas().sort_values("estimate", ascending=False)
print("Top 3:", pred7_df.head(3)[["nace", "estimate"]].to_string())
print("Bottom 3:", pred7_df.tail(3)[["nace", "estimate"]].to_string())
```

```{python}
model8 = smf.ols("vacancies ~ nace * size + public", data = df).fit()
pred8 = predictions(model8,
                    newdata = datagrid(model8, nace = df["nace"].unique(), size = "Medium", public = "0"))
pred8.to_pandas().sort_values("estimate", ascending=False).head(3)[["nace", "estimate"]]
```

## Julia

```{julia}
model7 = lm(@formula(vacancies ~ nace + size + public), df)
```

```{julia}
effects(Dict(:nace => sort(unique(df.nace)),
             :size => ["Medium"],
             :public => ["0"]), model7)
```

```{julia}
model8 = lm(@formula(vacancies ~ nace * size + public), df)
```

```{julia}
effects(Dict(:nace => sort(unique(df.nace)),
             :size => ["Medium"],
             :public => ["0"]), model8)
```
:::
