---
title: "Interactions and scaling of regression inputs"
author: "Maciej Beręsewicz"
format:
  html:
    embed-resources: true
    number-sections: true
    df-print: kable
execute:
  eval: true
  warning: false
  message: false
---

```{r setup, echo = FALSE}
JuliaCall::julia_setup(JULIA_HOME = "/Applications/Julia-1.11.app/Contents/Resources/julia/bin/")
knitr::opts_chunk$set(engine.path = list(
  python = "/opt/anaconda3/bin/python3.12",
  julia = "/Applications/Julia-1.11.app/Contents/Resources/julia/bin/"
))
```

## Introduction

This supplementary notebook covers categorical $\times$ continuous interactions,
centering, standardisation, and Gelman's (2008) approach to scaling regression
inputs. These techniques improve coefficient interpretability, especially in
models with interactions.

**References:**

- Gelman, A. (2008). Scaling regression inputs by dividing by two standard
  deviations. *Statistics in Medicine*, 27(15), 2865--2873.
  [PDF](https://sites.stat.columbia.edu/gelman/research/published/standardizing7.pdf)

## Data

::: panel-tabset
### R

```{r data-r, message=FALSE, warning=FALSE}
library(ggplot2)
df <- read.csv("data/polish-jvs.csv", colClasses = c("character", "factor", rep("character", 4), "numeric"))
nace_counts <- as.data.frame(table(df$nace_division))
names(nace_counts) <- c("nace_division", "n_firms")
df <- merge(df, nace_counts, by = "nace_division")
df$n_firms <- as.numeric(df$n_firms) / 1000  ## in thousands for readable coefficients
head(df[, c("nace_division", "size", "n_firms", "vacancies")])
```

### Python

```{python data-py}
import pandas as pd
import numpy as np
import statsmodels.formula.api as smf
import matplotlib.pyplot as plt

df = pd.read_csv("data/polish-jvs.csv", dtype={"id": np.int64, "woj":str, "public":str,"size": str, "nace_division": str, "nace": str})
nace_counts = df.groupby("nace_division").size().reset_index(name="n_firms")
df = df.merge(nace_counts, on="nace_division")
df["n_firms"] = df["n_firms"] / 1000  ## in thousands for readable coefficients
print(df[["nace_division", "size", "n_firms", "vacancies"]].head())
```

### Julia

```{julia data-jl}
using CSV, DataFrames, GLM, Statistics

df = CSV.read("data/polish-jvs.csv", DataFrame,
              types = Dict(:id => String, :woj=> String, :public=> String,
                           :size => String, :nace_division => String, :nace => String))
nace_counts = combine(groupby(df, :nace_division), nrow => :n_firms)
df = leftjoin(df, nace_counts, on = :nace_division)
df.n_firms = df.n_firms ./ 1000  ## in thousands for readable coefficients
first(df[:, [:nace_division, :size, :n_firms, :vacancies]], 5)
```
:::

## Categorical $\times$ continuous interaction

### Model without interaction (parallel lines)

The model `vacancies ~ size + n_firms` assumes the same slope for `n_firms` across
all size groups -- only the intercept differs. The lines are parallel.

::: panel-tabset
### R

```{r model-parallel-r}
model_par <- lm(vacancies ~ size + n_firms, data = df)
summary(model_par)
```

### Python

```{python model-parallel-py}
model_par = smf.ols("vacancies ~ size + n_firms", data = df).fit()
print(model_par.summary())
```

### Julia

```{julia model-parallel-jl}
model_par = lm(@formula(vacancies ~ size + n_firms), df)
```
:::

### Model with interaction (different slopes)

The model `vacancies ~ size * n_firms` allows a different slope for `n_firms` in each
size group. The coefficient of `n_firms` is the slope for the reference group (Large).
The interaction terms measure how much the slope differs for other groups.

::: panel-tabset
### R

```{r model-interaction-r}
model_slopes <- lm(vacancies ~ size * n_firms, data = df)
summary(model_slopes)

## Slopes by group
b <- coef(model_slopes)
cat("Slope for Large:", b["n_firms"], "\n")
cat("Slope for Medium:", b["n_firms"] + b["sizeMedium:n_firms"], "\n")
cat("Slope for Small:", b["n_firms"] + b["sizeSmall:n_firms"], "\n")
```

### Python

```{python model-interaction-py}
model_slopes = smf.ols("vacancies ~ size * n_firms", data = df).fit()
print(model_slopes.summary())

## Slopes by group
b = model_slopes.params
print(f"Slope for Large: {b['n_firms']:.6f}")
print(f"Slope for Medium: {b['n_firms'] + b['size[T.Medium]:n_firms']:.6f}")
print(f"Slope for Small: {b['n_firms'] + b['size[T.Small]:n_firms']:.6f}")
```

### Julia

```{julia model-interaction-jl}
model_slopes = lm(@formula(vacancies ~ size * n_firms), df)

## Slopes by group
b = coef(model_slopes)
println("Slope for Large: ", b[4])
println("Slope for Medium: ", b[4] + b[5])
println("Slope for Small: ", b[4] + b[6])
```
:::

### Visualising parallel vs. crossing lines

::: panel-tabset
### R

```{r plot-interaction-r, fig.width=10, fig.height=4}
df$fitted_par <- fitted(model_par)
df$fitted_int <- fitted(model_slopes)

## Use distinct n_firms values per size group for clean lines
plot_data <- aggregate(cbind(fitted_par, fitted_int) ~ n_firms + size, data = df, FUN = mean)

p1 <- ggplot(plot_data, aes(x = n_firms, y = fitted_par, color = size)) +
  geom_line(linewidth = 1) +
  labs(title = "Without interaction (parallel lines)",
       y = "Fitted vacancies", x = "Number of firms in sector (thousands)") +
  theme_minimal()

p2 <- ggplot(plot_data, aes(x = n_firms, y = fitted_int, color = size)) +
  geom_line(linewidth = 1) +
  labs(title = "With interaction (different slopes)",
       y = "Fitted vacancies", x = "Number of firms in sector (thousands)") +
  theme_minimal()

library(patchwork)
p1 + p2
```

### Python

```{python plot-interaction-py}
df["fitted_par"] = model_par.fittedvalues
df["fitted_int"] = model_slopes.fittedvalues

plot_data = df.groupby(["n_firms", "size"])[["fitted_par", "fitted_int"]].mean().reset_index()

fig, axes = plt.subplots(1, 2, figsize=(12, 4))
for name, grp in plot_data.groupby("size"):
    axes[0].plot(grp["n_firms"], grp["fitted_par"], label=name)
    axes[1].plot(grp["n_firms"], grp["fitted_int"], label=name)
axes[0].set_title("Without interaction (parallel lines)")
axes[1].set_title("With interaction (different slopes)")
for ax in axes:
    ax.set_xlabel("Number of firms in sector (thousands)")
    ax.set_ylabel("Fitted vacancies")
    ax.legend()
plt.tight_layout()
plt.show()
```
:::

## Centering continuous variables

### Why center?

In a model with interaction `vacancies ~ size * n_firms`, the "main effect" of `size`
is the effect **when `n_firms = 0`**. But `n_firms = 0` is outside the observed range
(there are no sectors with zero firms), so this main effect is an extrapolation with
no practical meaning.

**Grand-mean centering** replaces `n_firms` with `n_firms - mean(n_firms)`. Now the
main effect of `size` is the effect **at the average sector size**, which is much more
interpretable.

### Grand-mean centering

::: panel-tabset
### R

```{r centering-r}
df$n_firms_c <- df$n_firms - mean(df$n_firms)

model_cent <- lm(vacancies ~ size * n_firms_c, data = df)
summary(model_cent)
```

### Python

```{python centering-py}
df["n_firms_c"] = df["n_firms"] - df["n_firms"].mean()

model_cent = smf.ols("vacancies ~ size * n_firms_c", data = df).fit()
print(model_cent.summary())
```

### Julia

```{julia centering-jl}
df.n_firms_c = df.n_firms .- mean(df.n_firms)

model_cent = lm(@formula(vacancies ~ size * n_firms_c), df)
```
:::

### What changes and what stays the same?

::: panel-tabset
### R

```{r centering-compare-r}
## R-squared: identical
cat("R² (uncentered):", summary(model_slopes)$r.squared, "\n")
cat("R² (centered):  ", summary(model_cent)$r.squared, "\n")

## Interaction coefficients: identical
cat("\nInteraction coefs (uncentered):\n")
print(coef(model_slopes)[5:6])
cat("Interaction coefs (centered):\n")
print(coef(model_cent)[5:6])

## Main-effect coefficients: different
cat("\nMain effects (uncentered):\n")
print(coef(model_slopes)[1:3])
cat("Main effects (centered):\n")
print(coef(model_cent)[1:3])

## Fitted values: identical
cat("\nFitted values identical?", all.equal(fitted(model_slopes), fitted(model_cent)), "\n")
```

### Python

```{python centering-compare-py}
## R-squared: identical
print(f"R² (uncentered): {model_slopes.rsquared:.6f}")
print(f"R² (centered):   {model_cent.rsquared:.6f}")

## Interaction coefficients: identical
print(f"\nInteraction coefs (uncentered):")
print(model_slopes.params.filter(like=":"))
print(f"Interaction coefs (centered):")
print(model_cent.params.filter(like=":"))

## Main-effect coefficients: different
print(f"\nMain effects (uncentered):")
print(model_slopes.params.iloc[:3])
print(f"Main effects (centered):")
print(model_cent.params.iloc[:3])
```

### Julia

```{julia centering-compare-jl}
## R-squared: identical
println("R² (uncentered): ", r2(model_slopes))
println("R² (centered):   ", r2(model_cent))

## Interaction coefficients: identical
println("\nInteraction coefs (uncentered): ", coef(model_slopes)[5:6])
println("Interaction coefs (centered):   ", coef(model_cent)[5:6])

## Main-effect coefficients: different
println("\nMain effects (uncentered): ", coef(model_slopes)[1:3])
println("Main effects (centered):   ", coef(model_cent)[1:3])
```
:::

### Standardisation (z-score)

Standardising (`z-score`) additionally rescales to standard deviation units, so
coefficients become comparable across variables with different scales.

::: panel-tabset
### R

```{r standardisation-r}
df$n_firms_s <- as.numeric(scale(df$n_firms))

model_std <- lm(vacancies ~ size * n_firms_s, data = df)
summary(model_std)
```

### Python

```{python standardisation-py}
df["n_firms_s"] = (df["n_firms"] - df["n_firms"].mean()) / df["n_firms"].std()

model_std = smf.ols("vacancies ~ size * n_firms_s", data = df).fit()
print(model_std.summary())
```

### Julia

```{julia standardisation-jl}
df.n_firms_s = (df.n_firms .- mean(df.n_firms)) ./ std(df.n_firms)

model_std = lm(@formula(vacancies ~ size * n_firms_s), df)
```
:::

## Gelman's approach: scaling by two standard deviations

### The problem with standard standardisation

When comparing continuous and binary predictors in the same model, dividing the
continuous input by 1 SD does not make coefficients directly comparable:

- A binary (0/1) variable like `public` has a natural 1-unit change (from 0 to 1),
  representing its full range.
- For a continuous variable, a 1-SD change is only a moderate shift -- roughly
  covering 68% of the distribution.

Gelman (2008) proposes a simple fix: **divide continuous inputs by 2 SD** instead
of 1 SD. Then a 1-unit change in the rescaled continuous variable corresponds to
moving from 1 SD below to 1 SD above the mean -- a "typical" swing that is more
comparable in magnitude to the 0-to-1 swing of a binary predictor.

### Centering binary variables by their means

Gelman also recommends **centering binary inputs by their means**. Instead of
leaving `public` as 0/1 with the intercept representing the `public = 0` group:

- Replace `public` with `public - mean(public)`
- The intercept now represents the expected value at the "average" mix of
  public/private firms, not just at the reference level

This is especially useful in models with interactions, where the "main effect"
of a categorical variable is conditional on the other variable being at its
reference level.

### Implementation

::: panel-tabset
### R

```{r gelman-r}
## Gelman scaling: continuous by 2*SD
df$n_firms_g <- (df$n_firms - mean(df$n_firms)) / (2 * sd(df$n_firms))

## Center binary by its mean
df$public_num <- as.numeric(df$public == "1")
df$public_c <- df$public_num - mean(df$public_num)

## Model with Gelman-scaled inputs
model_gelman <- lm(vacancies ~ size + n_firms_g + public_c, data = df)
summary(model_gelman)
```

### Python

```{python gelman-py}
## Gelman scaling
df["n_firms_g"] = (df["n_firms"] - df["n_firms"].mean()) / (2 * df["n_firms"].std())

## Center binary by its mean
df["public_num"] = (df["public"] == "1").astype(float)
df["public_c"] = df["public_num"] - df["public_num"].mean()

model_gelman = smf.ols("vacancies ~ size + n_firms_g + public_c", data=df).fit()
print(model_gelman.summary())
```

### Julia

```{julia gelman-jl}
df.n_firms_g = (df.n_firms .- mean(df.n_firms)) ./ (2 * std(df.n_firms))
df.public_num = Float64.(df.public .== "1")
df.public_c = df.public_num .- mean(df.public_num)

model_gelman = lm(@formula(vacancies ~ size + n_firms_g + public_c), df)
```
:::

### Comparison: coefficient magnitudes

Compare the same model under three scaling approaches to see how coefficient
comparability changes.

::: panel-tabset
### R

```{r gelman-compare-r}
## Original (no scaling)
model_orig <- lm(vacancies ~ size + n_firms + public_num, data = df)

## Standard z-score (1 SD)
model_1sd <- lm(vacancies ~ size + n_firms_s + public_num, data = df)

## Gelman (2 SD for continuous, mean-centered binary)
model_2sd <- lm(vacancies ~ size + n_firms_g + public_c, data = df)

comparison <- data.frame(
  original = coef(model_orig),
  z_score_1sd = coef(model_1sd),
  gelman_2sd = coef(model_2sd)
)
round(comparison, 4)
```

Note how the Gelman-scaled coefficient for `n_firms_g` and `public_c` are now
on a more comparable scale: both represent the effect of a "typical" change
in the predictor.

### Python

```{python gelman-compare-py}
model_orig = smf.ols("vacancies ~ size + n_firms + public_num", data=df).fit()
model_1sd = smf.ols("vacancies ~ size + n_firms_s + public_num", data=df).fit()
model_2sd = smf.ols("vacancies ~ size + n_firms_g + public_c", data=df).fit()

comparison = pd.DataFrame({
    "original": model_orig.params,
    "z_score_1sd": model_1sd.params,
    "gelman_2sd": model_2sd.params
})
print(comparison.round(4))
```

### Julia

```{julia gelman-compare-jl}
model_orig = lm(@formula(vacancies ~ size + n_firms + public_num), df)
model_1sd = lm(@formula(vacancies ~ size + n_firms_s + public_num), df)
model_2sd = lm(@formula(vacancies ~ size + n_firms_g + public_c), df)

println("Original:    ", round.(coef(model_orig), digits=4))
println("Z-score 1SD: ", round.(coef(model_1sd), digits=4))
println("Gelman 2SD:  ", round.(coef(model_2sd), digits=4))
```
:::
