---
title: "Marginal effects"
author: "Maciej Beręsewicz"
format:
  html:
    embed-resources: true
    table-of-contents: true
    number-sections: true
    df-print: kable
editor: source
execute:
  eval: true
  warning: false
  message: false
filters:
  - ripper
ripper:
  output-dir: "extracted"
  script-links-position: "none"
  include-yaml: true
---

```{r setup, echo = FALSE}
JuliaCall::julia_setup(JULIA_HOME = "/Applications/Julia-1.11.app/Contents/Resources/julia/bin/")
knitr::opts_chunk$set(engine.path = list(
  python = "/opt/anaconda3/bin/python3.12",
  julia = "/Applications/Julia-1.11.app/Contents/Resources/julia/bin/"
))
```

# Packages

::: panel-tabset
## R

Install `marginaleffects` package if necessary.

```{r pkg-install-r, eval = FALSE}
install.packages("marginaleffects")
install.packages("car")
```

```{r pkg-load-r}
library(marginaleffects)
library(car)
```

## Python

```{bash pkg-install-py, eval = F}
!pip install statsmodels
!pip install marginaleffects
```

```{python pkg-load-py}
import pandas as pd
import numpy as np
import statsmodels.api as sm
import statsmodels.formula.api as smf
from marginaleffects import *
```

## Julia

```{julia pkg-install-jl, eval = F}
using Pkg
Pkg.add(["Effects", "StatsBase", "CSV", "CategoricalArrays", "GLM"])
```

```{julia pkg-load-jl}
using Effects
using StatsBase
using CSV
using CategoricalArrays
using GLM
using Statistics
using DataFrames
```
:::

# Data

Read in the dataset for the lecture and specify column classes (types).

::: panel-tabset
## R

```{r data-r}
df <- read.csv("../../data/polish-jvs.csv", colClasses = c("character", "factor", rep("character", 4), "numeric"))
head(df)
```

## Python

```{python data-py}
df = pd.read_csv("../../data/polish-jvs.csv", dtype={"id": np.int64, "woj":str, "public":str,"size": str, "nace_division": str, "nace": str})
df.head()
```

## Julia

```{julia data-jl}
df=CSV.read("../../data/polish-jvs.csv", DataFrame,
            types = Dict(:id => String, :woj=> String, :public=> String,
                         :size => String, :nace_division => String, :nace => String));
first(df, 5)
```
:::

# Why marginal effects?

In models with interactions or polynomial terms, the effect of a variable on the
response depends on the values of other variables. The regression coefficients
alone no longer have a straightforward interpretation. Marginal effects provide
a way to summarise the effect of each variable in a meaningful way.

For a comprehensive treatment of marginal effects in R and Python, see
Arel-Bundock, Greifer, and Heiss (2024), *How to Interpret Statistical Models
Using marginaleffects for R and Python*, Journal of Statistical Software, 111(9), 1--32.

# Model with interactions

Consider the following model

$$
\text{vacancies} = \beta_0 + \boldsymbol{\beta}_{\text{woj}} + \boldsymbol{\beta}_{\text{size}} + \beta_{\text{public}} + \boldsymbol{\beta}_{\text{woj} \times \text{size}} + \boldsymbol{\beta}_{\text{woj} \times \text{public}} + \varepsilon,
$$

where $\times$ denotes interactions between two variables.

::: panel-tabset
## R

```{r model-int-r}
model5 <- lm(vacancies ~ woj*size + woj*public, df)
summary(model5)
```

Calculate **Average Marginal Effects** (AME) -- the marginal effect is computed
for each observation, then averaged across the sample.

```{r ame-r}
avg_slopes(model5)
```

Calculate **Marginal Effects at the Mean** (MEM) -- the marginal effect evaluated
at the mean values of all other variables.

```{r mem-r}
avg_slopes(model5, newdata = "mean")
```

## Python

```{python model-int-py}
model5 = smf.ols("vacancies ~ woj*size + woj*public", data = df).fit()
print(model5.summary())
```

Calculate **Average Marginal Effects** (AME).

```{python ame-py}
model5_ame = avg_slopes(model5)
model5_ame.to_pandas().head()
```

Calculate **Marginal Effects at the Mean** (MEM).

```{python mem-py}
model5_mem = avg_slopes(model5, newdata="mean")
model5_mem.to_pandas().head()
```

## Julia

```{julia model-int-jl}
model5 = lm(@formula(vacancies ~ woj*size + woj*public), df)
```

`Effects.jl` returns marginal effects for each level separately, not as
deviations from the reference level.

```{julia effects-jl}
vcat(
  effects(Dict(:woj => sort(unique(df.woj))), model5),
  effects(Dict(:public => sort(unique(df.public))), model5),
  effects(Dict(:size => sort(unique(df.size))), model5),
  cols = :union
)
```
:::

# AME by subgroups

We can compute AME of one variable separately for subgroups defined by another
variable. For example, does the effect of `size` differ between public and
private firms?

::: panel-tabset
## R

```{r ame-by-r}
avg_slopes(model5, variables = "size", by = "public")
```

## Python

```{python ame-by-py}
avg_slopes(model5, variables = "size", by = "public").to_pandas()
```

## Julia

```{julia ame-by-jl}
effects(Dict(:size => sort(unique(df.size))), model5)
```
:::

# Marginal Effects at Representative Values (MERV)

We can also compute fitted values at specific combinations of variable
values. For example, expected vacancies for each voivodeship, holding
`size = "Medium"` and `public = "0"`.

::: panel-tabset
## R

```{r merv-r}
predictions(model5,
            newdata = datagrid(woj = unique(df$woj), size = "Medium", public = "0"))
```

## Python

```{python merv-py}
from marginaleffects import predictions, datagrid
pred = predictions(model5,
                   newdata = datagrid(model5, woj = df["woj"].unique(), size = "Medium", public = "0"))
pred.to_pandas().head()
```

## Julia

```{julia merv-jl}
effects(Dict(:woj => sort(unique(df.woj)),
             :size => ["Medium"],
             :public => ["0"]), model5)
```
:::

# Exercises

## Exercise 1: Average Marginal Effect of firm size

Fit the model: `vacancies ~ size * public + woj`.

Tasks:

1. How many parameters does this model estimate?
2. Compute the AME of `size`. Interpret: what is the average difference in
   expected vacancies between Large and Small firms, averaging over all
   values of `public` and `woj`?
3. Compute the AME of `size` separately for `public = 0` and `public = 1`.
   Does firm size matter more in the public or private sector?

::: panel-tabset
## R

```{r ex1-model-r}
model6 <- lm(vacancies ~ size * public + woj, data = df)
summary(model6)
```

```{r ex1-npar-r}
cat("Number of parameters:", length(coef(model6)), "\n")
```

AME of `size`:

```{r ex1-ame-r}
avg_slopes(model6, variables = "size")
```

AME of `size` by `public`:

```{r ex1-ame-by-r}
avg_slopes(model6, variables = "size", by = "public")
```

## Python

```{python ex1-model-py}
model6 = smf.ols("vacancies ~ size * public + woj", data = df).fit()
print(f"Number of parameters: {len(model6.params)}")
```

AME of `size`:

```{python ex1-ame-py}
avg_slopes(model6, variables = "size").to_pandas()
```

AME of `size` by `public`:

```{python ex1-ame-by-py}
avg_slopes(model6, variables = "size", by = "public").to_pandas()
```

## Julia

```{julia ex1-model-jl}
model6 = lm(@formula(vacancies ~ size * public + woj), df)
println("Number of parameters: ", length(coef(model6)))
```

```{julia ex1-effects-jl}
effects(Dict(:size => sort(unique(df.size))), model6)
```
:::

## Exercise 2: Fitted values across NACE sectors

Fit the model: `vacancies ~ nace + size + public`.

Tasks:

1. Compute the fitted values for each NACE section, holding
   `size = "Medium"` and `public = "0"`.
2. Which 3 sectors have the highest expected vacancies? Which 3 have the
   lowest?
3. Add the interaction `nace * size` and repeat. Do the rankings change
   across size categories?

::: panel-tabset
## R

```{r ex2-model-r}
model7 <- lm(vacancies ~ nace + size + public, data = df)
summary(model7)
```

Fitted values at representative values:

```{r ex2-pred-r}
pred7 <- predictions(model7,
                     newdata = datagrid(nace = unique(df$nace), size = "Medium", public = "0"))
pred7 <- as.data.frame(pred7)
pred7_sorted <- pred7[order(-pred7$estimate), c("nace", "estimate")]
head(pred7_sorted, 3)
tail(pred7_sorted, 3)
```

Now add the interaction `nace * size`:

```{r ex2-int-model-r}
model8 <- lm(vacancies ~ nace * size + public, data = df)
```

```{r ex2-int-pred-r}
pred8_medium <- predictions(model8,
                            newdata = datagrid(nace = unique(df$nace), size = "Medium", public = "0"))
pred8_large <- predictions(model8,
                           newdata = datagrid(nace = unique(df$nace), size = "Large", public = "0"))
pred8_medium <- as.data.frame(pred8_medium)
pred8_large <- as.data.frame(pred8_large)
cat("Top 3 NACE (Medium):\n")
head(pred8_medium[order(-pred8_medium$estimate), "nace"], 3)
cat("Top 3 NACE (Large):\n")
head(pred8_large[order(-pred8_large$estimate), "nace"], 3)
```

## Python

```{python ex2-model-py}
model7 = smf.ols("vacancies ~ nace + size + public", data = df).fit()
```

```{python ex2-pred-py}
from marginaleffects import predictions, datagrid
pred7 = predictions(model7,
                    newdata = datagrid(model7, nace = df["nace"].unique(), size = "Medium", public = "0"))
pred7_df = pred7.to_pandas().sort_values("estimate", ascending=False)
print("Top 3:", pred7_df.head(3)[["nace", "estimate"]].to_string())
print("Bottom 3:", pred7_df.tail(3)[["nace", "estimate"]].to_string())
```

```{python ex2-int-py}
model8 = smf.ols("vacancies ~ nace * size + public", data = df).fit()
pred8 = predictions(model8,
                    newdata = datagrid(model8, nace = df["nace"].unique(), size = "Medium", public = "0"))
pred8.to_pandas().sort_values("estimate", ascending=False).head(3)[["nace", "estimate"]]
```

## Julia

```{julia ex2-model-jl}
model7 = lm(@formula(vacancies ~ nace + size + public), df)
```

```{julia ex2-pred-jl}
effects(Dict(:nace => sort(unique(df.nace)),
             :size => ["Medium"],
             :public => ["0"]), model7)
```

```{julia ex2-int-model-jl}
model8 = lm(@formula(vacancies ~ nace * size + public), df)
```

```{julia ex2-int-pred-jl}
effects(Dict(:nace => sort(unique(df.nace)),
             :size => ["Medium"],
             :public => ["0"]), model8)
```
:::
