---
title: "Generalized linear models: count data"
author: "Maciej Beręsewicz"
format:
  html:
    embed-resources: true
    table-of-contents: true
    number-sections: true
    df-print: kable
editor: source
execute:
  eval: true
  warning: false
  message: false
filters:
  - ripper
ripper:
  output-dir: "extracted"
  script-links-position: "none"
  include-yaml: true
---

```{r setup, echo = FALSE}
JuliaCall::julia_setup(JULIA_HOME = "/Applications/Julia-1.11.app/Contents/Resources/julia/bin/")
knitr::opts_chunk$set(engine.path = list(
  python = "/opt/anaconda3/bin/python3.12",
  julia = "/Applications/Julia-1.11.app/Contents/Resources/julia/bin/"
))
```

# Packages

::: panel-tabset
## R

```{r pkg-install-r, eval = FALSE}
install.packages("MASS")
install.packages("AER")
install.packages("lmtest")
install.packages("performance")
install.packages("marginaleffects")
install.packages("modelsummary")
```

```{r pkg-load-r}
library(MASS)
library(AER)
library(lmtest)
library(performance)
library(marginaleffects)
library(modelsummary)
```

## Python

```{bash pkg-install-py, eval = FALSE}
!pip install statsmodels
!pip install scipy
!pip install marginaleffects
```

```{python pkg-load-py}
import numpy as np
import pandas as pd
import statsmodels.api as sm
import statsmodels.formula.api as smf
```

## Julia

```{julia pkg-install-jl, eval = FALSE}
using Pkg
Pkg.add(["GLM", "CSV", "DataFrames", "CategoricalArrays", "Statistics", "Distributions"])
```

```{julia pkg-load-jl}
using GLM
using CSV
using DataFrames
using CategoricalArrays
using Statistics
```
:::

# Data

The Polish Job Vacancy Survey (JVS): one row per firm, response variable `vacancies`.

::: panel-tabset
## R

```{r data-r}
df <- read.csv("data/polish-jvs.csv",
               colClasses = c("character", "factor", rep("character", 4), "numeric"))
df$size   <- factor(df$size,   levels = c("Large", "Medium", "Small"))
df$public <- factor(df$public)
df$nace   <- factor(df$nace)
head(df)
```

```{r data-summary-r}
c(mean      = mean(df$vacancies),
  var       = var(df$vacancies),
  prop_zero = mean(df$vacancies == 0))
```

## Python

```{python data-py}
df = pd.read_csv("data/polish-jvs.csv",
                 dtype={"id": np.int64, "woj": str, "public": str,
                        "size": str, "nace_division": str, "nace": str})
df["size"] = pd.Categorical(df["size"], categories=["Large", "Medium", "Small"])
df.head()
```

```{python data-summary-py}
{"mean":      df["vacancies"].mean(),
 "var":       df["vacancies"].var(),
 "prop_zero": (df["vacancies"] == 0).mean()}
```

## Julia

```{julia data-jl}
df = CSV.read("data/polish-jvs.csv", DataFrame,
              types = Dict(:id => String, :woj => String, :public => String,
                           :size => String, :nace_division => String, :nace => String));
df.size   = categorical(df.size, levels = ["Large", "Medium", "Small"]);
df.public = categorical(df.public);
df.nace   = categorical(df.nace);
first(df, 5)
```

```{julia data-summary-jl}
(mean      = mean(df.vacancies),
 var       = var(df.vacancies),
 prop_zero = mean(df.vacancies .== 0))
```
:::

The mean is roughly $1.0$, the variance is roughly $96$, and about 80% of firms post zero vacancies. Variance is two orders of magnitude larger than the mean — far from a Gaussian situation.

# Why not OLS? Negative predictions on a count

We model `vacancies` with OLS using firm `size`, `public` status, and the NACE sector letter. Then we check: how many fitted values are negative?

::: panel-tabset
## R

```{r ols-neg-r}
ols <- lm(vacancies ~ size + public + nace, data = df)
preds_ols <- predict(ols)
cat("Min OLS prediction:", round(min(preds_ols), 3), "\n")
cat("Number negative   :", sum(preds_ols < 0),
    "out of", length(preds_ols), "\n")
```

## Python

```{python ols-neg-py}
ols = smf.ols("vacancies ~ C(size) + C(public) + C(nace)", data=df).fit()
preds_ols = ols.predict(df)
print("Min OLS prediction:", round(preds_ols.min(), 3))
print("Number negative   :", int((preds_ols < 0).sum()),
      "out of", len(preds_ols))
```

## Julia

```{julia ols-neg-jl}
ols = lm(@formula(vacancies ~ size + public + nace), df)
preds_ols = predict(ols)
println("Min OLS prediction: ", round(minimum(preds_ols), digits = 3))
println("Number negative   : ", sum(preds_ols .< 0), " out of ", length(preds_ols))
```
:::

OLS produces negative fitted means for some sub-groups — impossible for a count. The variance and distribution of residuals are also wrong: $\operatorname{Var}(Y\mid X)$ grows with $E(Y\mid X)$ for counts, so a single $\sigma^2$ misrepresents the data.

# Why `log(y + c)` is the wrong fix

A common workaround is to fit OLS on $\log(Y + c)$. Two problems:

1. coefficients **depend on the choice of $c$**;
2. back-transformed effects **disagree with Poisson** (the consistent estimator of multiplicative effects on the mean — Wooldridge 1999).

::: panel-tabset
## R

```{r logy-r}
fit_log_c <- function(c_val) {
  m <- lm(log(vacancies + c_val) ~ size + public, data = df)
  exp(coef(m))
}
cmp <- cbind(`c=1`     = fit_log_c(1),
             `c=0.5`   = fit_log_c(0.5),
             `c=0.001` = fit_log_c(0.001))
round(cmp, 3)
```

```{r poisson-baseline-r}
pois_simple <- glm(vacancies ~ size + public, data = df, family = poisson())
round(exp(coef(pois_simple)), 3)
```

## Python

```{python logy-py}
def fit_log_c(c_val):
    df["log_y"] = np.log(df["vacancies"] + c_val)
    m = smf.ols("log_y ~ C(size) + C(public)", data=df).fit()
    return np.exp(m.params)

cmp = pd.DataFrame({"c=1":     fit_log_c(1),
                    "c=0.5":   fit_log_c(0.5),
                    "c=0.001": fit_log_c(0.001)})
cmp.round(3)
```

```{python poisson-baseline-py}
pois_simple = smf.glm("vacancies ~ C(size) + C(public)", data=df,
                      family=sm.families.Poisson()).fit()
np.exp(pois_simple.params).round(3)
```

## Julia

```{julia logy-jl}
function fit_log_c(c_val)
    df.log_y = log.(df.vacancies .+ c_val)
    m = lm(@formula(log_y ~ size + public), df)
    return exp.(coef(m))
end

(c_1    = fit_log_c(1),
 c_05   = fit_log_c(0.5),
 c_0001 = fit_log_c(0.001))
```

```{julia poisson-baseline-jl}
pois_simple = glm(@formula(vacancies ~ size + public), df, Poisson(), LogLink())
exp.(coef(pois_simple))
```
:::

The columns for $c=1$, $c=0.5$, $c=0.001$ disagree with each other, and none coincides with the Poisson estimates. There is no principled way to pick $c$, and back-transforming OLS on $\log(Y+c)$ gives a biased estimate of multiplicative effects on $E[Y\mid X]$ (Jensen's inequality).

# Poisson regression

Poisson regression with the canonical log link models
$$
\log E(Y_i \mid \mathbf{x}_i) = \mathbf{x}_i^\top \boldsymbol{\beta}, \qquad
E(Y_i \mid \mathbf{x}_i) = \mu_i = \exp(\mathbf{x}_i^\top \boldsymbol{\beta}).
$$

::: panel-tabset
## R

```{r pois-r}
pois <- glm(vacancies ~ size + public + nace, data = df,
            family = poisson(link = "log"))
summary(pois)
```

`exp(coef)` gives **incidence rate ratios** (IRR) — the multiplicative effect on $E[Y]$.

```{r pois-irr-r}
round(exp(coef(pois)), 3)
```

```{r pois-modelplot-r}
modelplot(pois, exponentiate = TRUE) +
  ggplot2::geom_vline(xintercept = 1, linetype = 2)
```

## Python

```{python pois-py}
pois = smf.glm("vacancies ~ C(size) + C(public) + C(nace)", data=df,
               family=sm.families.Poisson()).fit()
print(pois.summary())
```

```{python pois-irr-py}
np.exp(pois.params).round(3)
```

## Julia

```{julia pois-jl}
pois = glm(@formula(vacancies ~ size + public + nace), df, Poisson(), LogLink())
```

```{julia pois-irr-jl}
exp.(coef(pois))
```
:::

**Interpretation.** For example, `sizeMedium = 0.14` means that, holding `public` and `nace` fixed, the expected number of vacancies for medium-sized firms is about 14% of that for large firms (the reference). Equivalently, `(0.14 - 1) * 100 ≈ -86%`.

# Detecting overdispersion

Pure Poisson assumes $\operatorname{Var}(Y\mid X) = E(Y\mid X)$. The Pearson dispersion statistic
$$
\hat{\phi} = \frac{1}{n - p} \sum_{i=1}^{n} \frac{(y_i - \hat{\mu}_i)^2}{\hat{\mu}_i}
$$
is close to 1 if Poisson is correct; $\hat{\phi} \gg 1$ signals overdispersion.

::: panel-tabset
## R

```{r phi-r}
phi <- sum(residuals(pois, type = "pearson")^2) / df.residual(pois)
round(phi, 2)
```

Cameron–Trivedi regression test:

```{r dispersiontest-r}
AER::dispersiontest(pois, trafo = 1)
```

`performance` wraps a similar diagnostic:

```{r perf-od-r}
performance::check_overdispersion(pois)
```

## Python

```{python phi-py}
mu_hat = pois.fittedvalues
pearson = (df["vacancies"] - mu_hat) / np.sqrt(mu_hat)
phi_py = float((pearson ** 2).sum() / pois.df_resid)
round(phi_py, 2)
```

## Julia

```{julia phi-jl}
mu_hat = predict(pois)
phi_jl = sum((df.vacancies .- mu_hat) .^ 2 ./ mu_hat) / dof_residual(pois)
round(phi_jl, digits = 2)
```
:::

The dispersion statistic is far above 1. Pure-Poisson SEs are anti-conservative by roughly $\sqrt{\hat{\phi}}$.

# Quasi-Poisson and Negative Binomial

Three options share the exponential mean specification but model the variance differently.

| Model         | $\operatorname{Var}(Y_i)$               | Likelihood | Dispersion estimated by  |
|---------------|-----------------------------------------|------------|--------------------------|
| Poisson       | $\mu_i$                                 | full       | --                       |
| Quasi-Poisson | $\phi\, \mu_i$                          | quasi-MLE  | Pearson statistic        |
| NB1           | $\mu_i + \mu_i / k = (1 + 1/k)\, \mu_i$ | full       | MLE for $k$              |
| NB2           | $\mu_i + \mu_i^2 / k$                   | full       | MLE for $k$              |

Quasi-Poisson and NB1 share the variance form $\operatorname{Var} = \phi\mu$ (with $\phi = 1 + 1/k$) and produce numerically similar $\hat{\boldsymbol{\beta}}$, but they are **not** the same model: the standard errors come from different assumptions.

::: panel-tabset
## R

```{r qp-r}
qp  <- glm(vacancies ~ size + public + nace, data = df,
           family = quasipoisson(link = "log"))
```

```{r nb-r}
nb2 <- MASS::glm.nb(vacancies ~ size + public + nace, data = df)
nb2$theta  # this is k in our notation
```

Side-by-side comparison (IRR with SE):

```{r modelsummary-r}
modelsummary(list(Poisson       = pois,
                  `Quasi-Poiss` = qp,
                  NB2           = nb2),
             exponentiate = TRUE,
             statistic    = "({std.error})",
             gof_omit     = "F|RMSE")
```

## Python

```{python nb-py}
## Iterative NB2 via GLM: estimate alpha from Poisson residuals, then refit
alpha_hat = ((pois.resid_pearson ** 2).sum() / pois.df_resid - 1) / pois.fittedvalues.mean()
for _ in range(25):
    nb2 = smf.glm("vacancies ~ C(size) + C(public) + C(nace)", data=df,
                  family=sm.families.NegativeBinomial(alpha=alpha_hat)).fit()
    alpha_hat = ((nb2.resid_pearson ** 2).sum() / nb2.df_resid - 1) / nb2.fittedvalues.mean()

print(nb2.summary())
print(f"\nEstimated alpha (1/k): {alpha_hat:.4f}")
```

`statsmodels` exposes quasi-Poisson via the `Poisson` family with `scale = "X2"`:

```{python qp-py}
qp = smf.glm("vacancies ~ C(size) + C(public) + C(nace)",
             data = df, family = sm.families.Poisson()).fit(scale = "X2")
qp.bse.head()
```

## Julia

```{julia nb-jl}
nb2 = negbin(@formula(vacancies ~ size + public + nace), df, LogLink())
```
:::

# Model comparison: Poisson vs. Negative Binomial

How do we decide whether the extra parameter in NB2 is justified? Three complementary tools:

| Tool | What it answers | Requirement |
|------|-----------------|-------------|
| LR test | Is the improvement *significant*? | Nested models |
| AIC / BIC | Which model fits *better*? | $\geq 2$ models (same response) |
| Rootogram | Where does the model *fail*? | Visual diagnostic |

## Likelihood ratio test

Poisson is nested within NB2 (it is the limiting case as $k \to \infty$). Under $H_0\colon k = \infty$ the LR statistic $2[\ell_{\text{NB2}} - \ell_{\text{Poisson}}]$ follows a $\chi^2_1$ distribution (one extra parameter).

::: panel-tabset
## R

```{r lr-test-r}
lr_stat <- 2 * (logLik(nb2) - logLik(pois))
lr_pval <- pchisq(as.numeric(lr_stat), df = 1, lower.tail = FALSE)
cat("LR statistic:", round(as.numeric(lr_stat), 1), "\n")
cat("p-value     :", format.pval(lr_pval), "\n")
```

Alternatively, `lmtest::lrtest()` automates this:

```{r lrtest-r}
lmtest::lrtest(pois, nb2)
```

## Python

```{python lr-test-py}
## NB2 GLM log-likelihood is conditional on estimated alpha;
## valid for LR test once alpha has converged
lr_stat = 2 * (nb2.llf - pois.llf)
from scipy.stats import chi2
lr_pval = chi2.sf(lr_stat, df=1)
print(f"LR statistic: {lr_stat:.1f}")
print(f"p-value     : {lr_pval:.2e}")
```

## Julia

```{julia lr-test-jl}
using Distributions
lr_stat = 2 * (loglikelihood(nb2) - loglikelihood(pois))
lr_pval = ccdf(Chisq(1), lr_stat)
println("LR statistic: ", round(lr_stat, digits = 1))
println("p-value     : ", lr_pval)
```
:::

The $p$-value is essentially zero — the data strongly reject the Poisson restriction.

## Information criteria

AIC and BIC penalise model complexity differently. Lower values indicate better fit–parsimony trade-off. AIC penalises by $2s$; BIC penalises by $\log(n)\cdot s$, so BIC is stricter for large $n$.

::: panel-tabset
## R

```{r aic-bic-r}
data.frame(
  Model = c("Poisson", "NB2"),
  AIC   = round(c(AIC(pois), AIC(nb2)), 1),
  BIC   = round(c(BIC(pois), BIC(nb2)), 1)
)
```

## Python

```{python aic-bic-py}
pd.DataFrame({
    "Model": ["Poisson", "NB2"],
    "AIC":   [round(pois.aic, 1), round(nb2.aic, 1)],
    "BIC":   [round(pois.bic, 1), round(nb2.bic, 1)]
})
```

## Julia

```{julia aic-bic-jl}
DataFrame(
    Model = ["Poisson", "NB2"],
    AIC   = round.([aic(pois), aic(nb2)], digits = 1),
    BIC   = round.([bic(pois), bic(nb2)], digits = 1)
)
```
:::

NB2 is preferred by both criteria — the massive improvement in log-likelihood easily offsets the one additional parameter.

## Summary

When overdispersion is detected ($\hat\phi \gg 1$):

- The LR test rejects Poisson in favour of NB2.
- AIC/BIC both favour NB2.
- Quasi-Poisson gives the same point estimates as Poisson but adjusts SEs; it does **not** provide a proper likelihood, so AIC/BIC are not available.

However, rejecting the Poisson *distribution* does not invalidate Poisson as a *mean estimator*. See the companion notebook **09a-poisson-vs-negbin** for a detailed discussion of when each model is appropriate.

# Marginal effects from a count model

Coefficients are multiplicative on the log scale. To express effects on the original (count) scale we use marginal effects.

::: panel-tabset
## R

Average Marginal Effect of `size`:

```{r ame-r}
avg_slopes(nb2, variables = "size")
```

`avg_slopes` differentiates $\hat\mu_i = \exp(\mathbf{x}_i^\top\hat{\boldsymbol{\beta}})$ for each observation, then averages — the right thing to do once the link is non-identity.

Predicted vacancies for each NACE letter, holding `size = "Medium"` and `public = "0"`:

```{r preds-by-nace-r}
predictions(nb2,
            newdata = datagrid(nace   = unique(df$nace),
                               size   = "Medium",
                               public = "0")) |>
  as.data.frame() |>
  subset(select = c("nace", "estimate", "conf.low", "conf.high"))
```

## Python

```{python ame-py, eval = FALSE}
from marginaleffects import avg_slopes
avg_slopes(pois, variables = "size").to_pandas()
```

## Julia

`Effects.jl` works with `GLM.jl` models; for a Poisson model it returns predictions on the response scale.

```{julia eff-jl, eval = FALSE}
using Effects
effects(Dict(:size => ["Large", "Medium", "Small"]), pois)
```
:::

# Exercises

## Exercise 1 — Poisson vs. NB on a sub-sample

Restrict the data to firms in NACE letter `C` (manufacturing). Refit the Poisson and NB2 models. Does the NB shape parameter $k$ become smaller or larger compared to the full-sample fit? What does that say about the dispersion of vacancies in this sector?

## Exercise 2 — OLS vs. Poisson predictions

For each combination of `size` and `public`, compute (i) the OLS-fitted mean from the model in *Why not OLS?*, and (ii) the Poisson-fitted mean from *Poisson regression*. Plot both against the empirical group means. Where do they disagree the most?

## Exercise 3 — Robust SEs for Poisson

Use the `sandwich` package to compute heteroskedasticity-robust SEs for the Poisson model. Compare the resulting CIs with the quasi-Poisson and NB2 CIs. Which intervals are widest? Narrowest?

# Further reading

- **Theory.** Cameron, A. C., & Trivedi, P. K. (2013). *Regression Analysis of Count Data* (2nd ed.). Cambridge University Press.
- **Why not log-transform.** O'Hara, R. B., & Kotze, D. J. (2010). Do not log-transform count data. *Methods in Ecology and Evolution*, 1(2), 118–122.
- **Logs with zeros.** Bellégo, C., Benatia, D., & Pape, L.-D. (2022). Dealing with logs and zeros in regression models. arXiv:2203.11820. Chen, J., & Roth, J. (2024). Logs with zeros? Some problems and solutions. *Quarterly Journal of Economics*, 139(2), 891–936.
- **Quasi-likelihood for counts.** Wooldridge, J. M. (1999). Distribution-free estimation of some nonlinear panel data models. *Journal of Econometrics*, 90(1), 77–97.
- **Poisson QMLE vs. NB.** Wooldridge, J. M. (2010). *Econometric Analysis of Cross Section and Panel Data* (2nd ed.), Chapter 18. MIT Press. — argues that Poisson with robust SEs is the preferred estimator for conditional means under minimal assumptions.
- **Diagnostics packages.** `DHARMa` (simulated residuals), `countreg` / `topmodels` (rootograms), `performance` (assumption checks).
