---
title: "Poisson QMLE vs. Negative Binomial"
author: "Maciej Beręsewicz"
format:
  html:
    embed-resources: true
    table-of-contents: true
    number-sections: true
    df-print: kable
editor: source
execute:
  eval: true
  warning: false
  message: false
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(engine.path = list(
  python = "/opt/anaconda3/bin/python3.12"
))
```

## Introduction

The likelihood ratio test and AIC/BIC favour NB2 over Poisson when overdispersion is present. But **rejecting the Poisson distribution does not invalidate Poisson as a mean estimator**. This notebook explains when each model is appropriate.

The key insight comes from Wooldridge (1999, 2010): Poisson regression with robust standard errors is a *quasi-maximum likelihood estimator* (QMLE) that is consistent for $\boldsymbol{\beta}$ under minimal assumptions.

## Setup

::: panel-tabset
### R

```{r}
library(MASS)
library(lmtest)
library(sandwich)
library(modelsummary)
```

### Python

```{python}
import numpy as np
import pandas as pd
import statsmodels.api as sm
import statsmodels.formula.api as smf
from scipy.stats import chi2
```
:::

## Data

::: panel-tabset
### R

```{r}
df <- read.csv("data/polish-jvs.csv",
               colClasses = c("character", "factor", rep("character", 4), "numeric"))
df$size   <- factor(df$size,   levels = c("Large", "Medium", "Small"))
df$public <- factor(df$public)
df$nace   <- factor(df$nace)
```

### Python

```{python}
df = pd.read_csv("data/polish-jvs.csv",
                 dtype={"id": np.int64, "woj": str, "public": str,
                        "size": str, "nace_division": str, "nace": str})
df["size"] = pd.Categorical(df["size"], categories=["Large", "Medium", "Small"])
```
:::

## Fit both models

::: panel-tabset
### R

```{r}
pois <- glm(vacancies ~ size + public + nace, data = df,
            family = poisson(link = "log"))
nb2  <- MASS::glm.nb(vacancies ~ size + public + nace, data = df)
```

### Python

```{python}
pois = smf.glm("vacancies ~ C(size) + C(public) + C(nace)", data=df,
               family=sm.families.Poisson()).fit()

## Iterative NB2 via GLM: estimate alpha from Poisson residuals, then refit
alpha_hat = ((pois.resid_pearson ** 2).sum() / pois.df_resid - 1) / pois.fittedvalues.mean()
for _ in range(25):
    nb2 = smf.glm("vacancies ~ C(size) + C(public) + C(nace)", data=df,
                  family=sm.families.NegativeBinomial(alpha=alpha_hat)).fit()
    alpha_hat = ((nb2.resid_pearson ** 2).sum() / nb2.df_resid - 1) / nb2.fittedvalues.mean()

print(f"Estimated alpha (1/k): {alpha_hat:.4f}")
```
:::

## The core argument: what are you estimating?

### If your goal is the conditional mean $E(Y \mid X)$

Poisson QMLE is consistent for $\boldsymbol{\beta}$ as long as the mean function is correctly specified:
$$
E(Y_i \mid \mathbf{x}_i) = \exp(\mathbf{x}_i^\top \boldsymbol{\beta}).
$$

This holds **regardless of whether the data are truly Poisson**. Overdispersion does not bias $\hat{\boldsymbol{\beta}}$ — it only inflates the naïve SEs. Robust (sandwich) SEs correct this.

Wooldridge's analogy: *Poisson QMLE is to the exponential mean what OLS is to the linear mean — a robust workhorse requiring no distributional assumptions beyond the first moment.*

### If your goal involves the full distribution

When you need $P(Y = k \mid X)$, quantiles, or simulation, the NB2 model captures the variance structure and provides a proper likelihood for all moments. Here the LR test and AIC/BIC are the right model selection tools.

### Decision table

| Goal | Recommended model |
|------|-------------------|
| Estimate effects on $E(Y \mid X)$ | Poisson + robust SEs |
| Full distributional inference ($P(Y=k)$, quantiles) | NB2 (or other explicit distribution) |
| Quick robustness check of SEs | Quasi-Poisson |

## Why not always use NB2?

Standard NB2 MLE assumes the quadratic variance function:
$$
\operatorname{Var}(Y \mid X) = \mu + \mu^2 / k.
$$

If this specific mean–variance relationship fails, NB2 MLE can be **inconsistent for the mean parameters**. Poisson QMLE avoids this risk entirely — it only requires correct specification of $E(Y \mid X)$.

## Empirical comparison: robust SEs vs. NB2 SEs

::: panel-tabset
### R

```{r}
pois_robust <- coeftest(pois, vcov = sandwich::vcovHC(pois, type = "HC0"))
```

Compare the three sets of standard errors:

```{r}
se_compare <- data.frame(
  Poisson_naive  = summary(pois)$coefficients[1:5, "Std. Error"],
  Poisson_robust = pois_robust[1:5, "Std. Error"],
  NB2            = summary(nb2)$coefficients[1:5, "Std. Error"]
)
round(se_compare, 4)
```

### Python

```{python}
pois_robust = smf.glm("vacancies ~ C(size) + C(public) + C(nace)", data=df,
                      family=sm.families.Poisson()).fit(cov_type="HC0")
idx = pois.bse.index[:5]
pd.DataFrame({
    "Poisson_naive":  pois.bse[idx].values,
    "Poisson_robust": pois_robust.bse[idx].values,
    "NB2":           nb2.bse[idx].values
}, index=idx).round(4)
```
:::

The robust SEs are larger than naïve Poisson SEs (correcting for overdispersion) and typically close to the NB2 SEs. Both approaches give similar inference for the conditional mean, but Poisson + robust SEs achieves this **without assuming a specific variance function**.

## Point estimates are (nearly) identical

The $\hat{\boldsymbol{\beta}}$ from Poisson and NB2 should be very close when both models specify the same mean function:

::: panel-tabset
### R

```{r}
coef_compare <- data.frame(
  Poisson = coef(pois)[1:5],
  NB2     = coef(nb2)[1:5],
  Diff    = coef(pois)[1:5] - coef(nb2)[1:5]
)
round(coef_compare, 5)
```

### Python

```{python}
idx = pois.params.index[:5]
pd.DataFrame({
    "Poisson": pois.params[idx].values,
    "NB2":     nb2.params[idx].values,
    "Diff":    (pois.params[idx] - nb2.params[idx]).values
}, index=idx).round(5)
```
:::


## Summary

- Rejecting equidispersion ($\hat\phi \gg 1$) tells you the data are *not Poisson-distributed*. It does **not** tell you that Poisson regression gives wrong $\hat{\boldsymbol{\beta}}$.
- If $E(Y \mid X) = \exp(\mathbf{x}^\top\boldsymbol{\beta})$ is correct, Poisson QMLE with robust SEs consistently estimates $\boldsymbol{\beta}$ — no matter how overdispersed the data are.
- Use NB2 when you need distributional predictions. Use Poisson + robust SEs when you only need effects on the mean.
- Do **not** choose NB2 over Poisson solely because a dispersion test is significant.

## References

- Wooldridge, J. M. (1999). Distribution-free estimation of some nonlinear panel data models. *Journal of Econometrics*, 90(1), 77–97.
- Wooldridge, J. M. (2010). *Econometric Analysis of Cross Section and Panel Data* (2nd ed.), Chapter 18. MIT Press.
- Cameron, A. C., & Trivedi, P. K. (2013). *Regression Analysis of Count Data* (2nd ed.). Cambridge University Press.
