---
title: "Rootograms for count regressions"
author: "Maciej Beręsewicz"
format:
  html:
    embed-resources: true
    table-of-contents: true
    number-sections: true
    df-print: kable
editor: source
execute:
  eval: true
  warning: false
  message: false
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(engine.path = list(
  python = "/opt/anaconda3/bin/python3.12"
))
```

## Introduction

A **rootogram** (Tukey, 1977; Kleiber & Zeileis, 2016) is a graphical
goodness-of-fit display for count models. For each observed count $k$
it compares

- the **observed frequency** $n_k = \#\{i : y_i = k\}$,
- the **expected frequency** under the fitted model
  $\hat n_k = \sum_{i=1}^{n} \hat P(Y_i = k \mid \mathbf{x}_i;\hat{\boldsymbol\theta})$.

In a **hanging** rootogram the bar of height $\sqrt{n_k}$ hangs from
the curve $\sqrt{\hat n_k}$, so its bottom sits at
$\sqrt{\hat n_k} - \sqrt{n_k}$. The square-root scale stretches the
tails so small frequencies are still visible. Reading the picture is
easy:

- bar bottom on the zero line → model fits this count perfectly,
- bar bottom **below** zero → model **under-predicts** that count
  (the data have more cases than expected),
- bar bottom **above** zero → model **over-predicts**.

In R rootograms are available via the `topmodels` package
(successor to `countreg`) and accept any fitted count model
(`glm`, `glm.nb`, `hurdle`, `zeroinfl`, …). In Python there is no
mainstream package that does this, so we write a small helper that
reproduces what `topmodels::rootogram` shows.

## Setup

::: panel-tabset
### R

```{r}
## install.packages("topmodels", repos = "https://R-Forge.R-project.org")
#remotes::install_svn("svn://R-Forge.R-project.org/svnroot/topmodels/pkg/topmodels")

library(topmodels)
library(MASS)
library(here)
```

### Python

```{python}
import numpy as np
import pandas as pd
import statsmodels.api as sm
import statsmodels.formula.api as smf
from scipy.stats import poisson, nbinom
import matplotlib.pyplot as plt
```
:::

## Data

We reuse the Polish Job Vacancy Survey (`vacancies` per company), the
same dataset as in lecture 9.

::: panel-tabset
### R

```{r}
df <- read.csv(here("data/polish-jvs.csv"),
               colClasses = c("character", "factor", rep("character", 4), "numeric"))
df$size   <- factor(df$size,   levels = c("Large", "Medium", "Small"))
df$public <- factor(df$public)
df$nace   <- factor(df$nace)
```

### Python

```{python}
from pathlib import Path
csv_path = Path("data/polish-jvs.csv")
if not csv_path.exists():
    csv_path = Path("../../data/polish-jvs.csv")

df = pd.read_csv(csv_path,
                 dtype={"id": np.int64, "woj": str, "public": str,
                        "size": str, "nace_division": str, "nace": str})
df["size"] = pd.Categorical(df["size"], categories=["Large", "Medium", "Small"])
```
:::

## Fit Poisson and Negative Binomial models

::: panel-tabset
### R

```{r}
pois <- glm(vacancies ~ size + public + nace, data = df,
            family = poisson(link = "log"))
nb2  <- MASS::glm.nb(vacancies ~ size + public + nace, data = df)
```

### Python

```{python}
pois = smf.glm("vacancies ~ C(size) + C(public) + C(nace)", data=df,
               family=sm.families.Poisson()).fit()

# Joint MLE for NB2 (matches MASS::glm.nb)
start = np.append(pois.params.values, 1.0)
nb2 = smf.negativebinomial("vacancies ~ C(size) + C(public) + C(nace)",
                           data=df).fit(start_params=start,
                                        method="bfgs", maxiter=500, disp=0)
```
:::

## Rootograms

We display counts up to $k = 20$. Vacancies in the JVS are very
right-skewed (≈ 80% zeros, mean ≈ 1, max in the hundreds), so 20 is
enough to show what matters: zeros, the small-count region, and the
beginning of the tail.

### R — `topmodels::rootogram`

`topmodels::rootogram()` takes a fitted model and produces a hanging
rootogram by default. 

```{r}
#| fig-width: 8
#| fig-height: 4

rootogram(pois, plot = TRUE, 
          main = "Poisson", 
          xlim = c(0,20))
```

```{r}
#| fig-width: 8
#| fig-height: 4

rootogram(nb2, plot = TRUE, 
          main = "Negative binomial (NB2)",
          xlim = c(0,20))
```

### Python — manual implementation

Same recipe as `topmodels`: observed counts, expected counts on the
$\sqrt{\cdot}$ scale, bars hanging from the expected curve. The
expected count for value $k$ is the **sum across observations** of the
fitted PMF evaluated at $k$.

For NB2, statsmodels parametrises overdispersion via $\alpha = 1/\theta$
so $\operatorname{Var}(Y) = \mu + \alpha\mu^2$. To map to
`scipy.stats.nbinom(n, p)` use $n = 1/\alpha$ and $p = n/(n + \mu)$.

```{python}
def rootogram(model, y, max_count=20, dist="poisson", ax=None,
              title=None):
    """Hanging rootogram for a fitted Poisson or NB2 GLM.

    model : statsmodels results object with .predict() returning fitted means
    y     : observed response (1d array-like of non-negative ints)
    dist  : "poisson" or "nbinom"
    """
    mu = np.asarray(model.predict())
    y  = np.asarray(y)
    ks = np.arange(max_count + 1)

    observed = np.array([(y == k).sum() for k in ks], dtype=float)

    if dist == "poisson":
        expected = np.array([poisson.pmf(k, mu).sum() for k in ks])
    elif dist == "nbinom":
        alpha = float(model.params["alpha"])
        n = 1.0 / alpha
        p = n / (n + mu)
        expected = np.array([nbinom.pmf(k, n, p).sum() for k in ks])
    else:
        raise ValueError("dist must be 'poisson' or 'nbinom'")

    obs_root = np.sqrt(observed)
    exp_root = np.sqrt(expected)

    if ax is None:
        _, ax = plt.subplots(figsize=(8, 4))

    ax.bar(ks, obs_root, bottom=exp_root - obs_root, width=0.7,
           color="lightgray", edgecolor="black", label="observed")
    ax.plot(ks, exp_root, "ro-", linewidth=1.5, markersize=4,
            label=r"$\sqrt{\mathrm{expected}}$")
    ax.axhline(0, color="black", linewidth=0.8)
    ax.set_xlabel("count")
    ax.set_ylabel(r"$\sqrt{\mathrm{frequency}}$")
    ax.set_xticks(ks[::2])
    if title:
        ax.set_title(title)
    ax.legend(loc="upper right")
    return ax
```

```{python}
#| fig-width: 8
#| fig-height: 4
fig, ax = plt.subplots(figsize=(8, 4))
rootogram(pois, df["vacancies"], dist="poisson", ax=ax, title="Poisson")
plt.show()
```

```{python}
#| fig-width: 8
#| fig-height: 4
fig, ax = plt.subplots(figsize=(8, 4))
rootogram(nb2, df["vacancies"], dist="nbinom", ax=ax,
          title="Negative binomial (NB2)")
plt.show()
```

## Numerical comparison

The picture is also visible in the raw observed/expected counts.

::: panel-tabset
### R

```{r}
ks <- 0:10
obs <- as.numeric(table(factor(df$vacancies, levels = ks)))

mu_p <- predict(pois, type = "response")
exp_p <- sapply(ks, function(k) sum(dpois(k, lambda = mu_p)))

mu_n  <- predict(nb2, type = "response")
theta <- nb2$theta
exp_n <- sapply(ks, function(k) sum(dnbinom(k, mu = mu_n, size = theta)))

round(data.frame(k = ks,
                 observed   = obs,
                 exp_pois   = exp_p,
                 exp_nb2    = exp_n,
                 diff_pois  = obs - exp_p,
                 diff_nb2   = obs - exp_n), 1)
```

### Python

```{python}
ks = np.arange(11)
obs = np.array([(df["vacancies"] == k).sum() for k in ks], dtype=float)

mu_p  = pois.predict()
exp_p = np.array([poisson.pmf(k, mu_p).sum() for k in ks])

alpha = float(nb2.params["alpha"])
n_par = 1.0 / alpha
mu_n  = nb2.predict()
p_par = n_par / (n_par + mu_n)
exp_n = np.array([nbinom.pmf(k, n_par, p_par).sum() for k in ks])

pd.DataFrame({"k": ks,
              "observed":  obs,
              "exp_pois":  exp_p,
              "exp_nb2":   exp_n,
              "diff_pois": obs - exp_p,
              "diff_nb2":  obs - exp_n}).round(1)
```
:::

## What the rootograms tell us

Both models have the same mean structure (Poisson QMLE and NB2 give
nearly identical $\hat{\boldsymbol\beta}$, see notebook 09a), but they
imply different distributions:

- **Poisson** badly **under-predicts the zeros** (bars dip well below
  the line at $k = 0$) and **over-predicts** the small positive counts
  ($k = 1, 2, 3$). This is the signature of overdispersion: the
  Poisson variance $\operatorname{Var}(Y) = \mu$ is too small, so mass
  gets concentrated near $\mu$ instead of being spread over zeros and
  the long tail.
- **NB2** has $\operatorname{Var}(Y) = \mu + \mu^2/\theta$. The extra
  variance lets it match the zeros and the right tail. Bar bottoms sit
  much closer to the zero line.

Conclusion (consistent with notebook 09a): use **NB2** (or another
distribution with extra variance) when you need the **distribution**
of $Y$ — for example to predict the proportion of firms with zero
vacancies, or to simulate from the fitted model. For the **mean**
$E(Y \mid X)$ alone, Poisson with robust standard errors is enough.

## References

- Tukey, J. W. (1977). *Exploratory Data Analysis*. Addison-Wesley.
- Kleiber, C., & Zeileis, A. (2016). Visualizing count data
  regressions using rootograms. *The American Statistician*, 70(3),
  296–303. <https://doi.org/10.1080/00031305.2016.1173590>
- Zeileis, A., Lang, M. N., & Stauffer, R. (2024). *topmodels:
  Infrastructure for Forecasting and Assessment of Probabilistic
  Models*. R package, <https://topmodels.R-Forge.R-project.org/>.
