---
title: "Generalized linear models: logistic regression"
author: "Maciej Beręsewicz"
format:
  html:
    embed-resources: true
    table-of-contents: true
    number-sections: true
    df-print: kable
editor: source
execute:
  eval: true
  warning: false
  message: false
filters:
  - ripper
ripper:
  output-dir: "extracted"
  script-links-position: "none"
  include-yaml: true
---

```{r setup, echo = FALSE}
JuliaCall::julia_setup(JULIA_HOME = "/Applications/Julia-1.11.app/Contents/Resources/julia/bin/")
knitr::opts_chunk$set(engine.path = list(
  python = "/opt/anaconda3/bin/python3.12",
  julia = "/Applications/Julia-1.11.app/Contents/Resources/julia/bin/"
))
```

# Packages

::: panel-tabset
## R

```{r pkg-install-r, eval = FALSE}
install.packages("marginaleffects")
install.packages("modelsummary")
install.packages("performance")
install.packages("pROC")
install.packages("car")
install.packages("broom")
```

```{r pkg-load-r}
library(marginaleffects)
library(modelsummary)
library(performance)
library(pROC)
library(broom)
```

## Python

```{bash pkg-install-py, eval = FALSE}
!pip install statsmodels scikit-learn
```

```{python pkg-load-py}
import numpy as np
import pandas as pd
import statsmodels.api as sm
import statsmodels.formula.api as smf
from sklearn.metrics import roc_auc_score, roc_curve
```

## Julia

```{julia pkg-install-jl, eval = FALSE}
using Pkg
Pkg.add(["GLM", "CSV", "DataFrames", "CategoricalArrays", "Statistics", "StatsBase", "Effects"])
```

```{julia pkg-load-jl}
using GLM
using CSV
using DataFrames
using CategoricalArrays
using Statistics
using StatsBase
```
:::

# Data

The Polish Job Vacancy Survey (JVS): one row per firm. We model whether a firm posts at least one vacancy:
$$
\text{has\_vacancy}_i = \mathbf{1}\{\text{vacancies}_i > 0\}, \qquad Y_i \in \{0, 1\}.
$$

::: panel-tabset
## R

```{r data-r}
df <- read.csv("data/polish-jvs.csv",
               colClasses = c("character", "factor", rep("character", 4), "numeric"))
df$size        <- factor(df$size,   levels = c("Large", "Medium", "Small"))
df$public      <- factor(df$public)
df$nace        <- factor(df$nace)
df$has_vacancy <- as.integer(df$vacancies > 0)
head(df)
```

```{r data-summary-r}
c(n         = nrow(df),
  prop_yes  = mean(df$has_vacancy),
  prop_pub  = mean(df$public == "1"))
```

## Python

```{python data-py}
df = pd.read_csv("data/polish-jvs.csv",
                 dtype={"id": np.int64, "woj": str, "public": str,
                        "size": str, "nace_division": str, "nace": str})
df["size"]        = pd.Categorical(df["size"], categories = ["Large", "Medium", "Small"])
df["has_vacancy"] = (df["vacancies"] > 0).astype(int)
df.head()
```

```{python data-summary-py}
{"n":        len(df),
 "prop_yes": df["has_vacancy"].mean(),
 "prop_pub": (df["public"] == "1").mean()}
```

## Julia

```{julia data-jl}
df = CSV.read("data/polish-jvs.csv", DataFrame,
              types = Dict(:id => String, :woj => String, :public => String,
                           :size => String, :nace_division => String, :nace => String));
df.size        = categorical(df.size, levels = ["Large", "Medium", "Small"]);
df.public      = categorical(df.public);
df.nace        = categorical(df.nace);
df.has_vacancy = ifelse.(df.vacancies .> 0, 1, 0);
first(df, 5)
```

```{julia data-summary-jl}
(n        = nrow(df),
 prop_yes = mean(df.has_vacancy),
 prop_pub = mean(df.public .== "1"))
```
:::

About 12% of firms post at least one vacancy.

# Why not LPM?

The **linear probability model** (LPM) fits $P(Y_i = 1 \mid \mathbf{x}_i) = \beta_0 + \mathbf{x}_i^\top\boldsymbol{\beta}$ by OLS. Three issues:

1. fitted probabilities can fall outside $[0, 1]$;
2. residual variance is intrinsically heteroskedastic ($\operatorname{Var}(Y\mid X) = \pi(1-\pi)$);
3. the Gaussian likelihood is misspecified, so likelihood-based inference (LR tests, AIC, normal-theory prediction intervals) does not apply.

Point estimates remain consistent for the **linear approximation** to $\pi(\mathbf{x})$; robust (HC) SEs give asymptotically valid Wald CIs. But the constant marginal-effect assumption and out-of-range predictions are not repairable without changing the model.

::: panel-tabset
## R

```{r lpm-r}
lpm <- lm(has_vacancy ~ size + public + nace, data = df)
preds_lpm <- predict(lpm)
c(min      = round(min(preds_lpm), 4),
  max      = round(max(preds_lpm), 4),
  n_below0 = sum(preds_lpm < 0),
  n_above1 = sum(preds_lpm > 1))
```

## Python

```{python lpm-py}
lpm = smf.ols("has_vacancy ~ C(size) + C(public) + C(nace)", data = df).fit()
preds_lpm = lpm.predict(df)
{"min":      round(preds_lpm.min(), 4),
 "max":      round(preds_lpm.max(), 4),
 "n_below0": int((preds_lpm < 0).sum()),
 "n_above1": int((preds_lpm > 1).sum())}
```

## Julia

```{julia lpm-jl}
lpm = lm(@formula(has_vacancy ~ size + public + nace), df)
preds_lpm = predict(lpm)
(min      = round(minimum(preds_lpm), digits = 4),
 max      = round(maximum(preds_lpm), digits = 4),
 n_below0 = sum(preds_lpm .< 0),
 n_above1 = sum(preds_lpm .> 1))
```
:::

LPM produces fitted probabilities outside $[0,1]$ for sub-groups with very low base rates.

# Odds, log-odds, and odds ratios

For a binary response with $\pi = \Pr(Y = 1)$:
$$
\text{odds} = \frac{\pi}{1-\pi}, \qquad \operatorname{logit}(\pi) = \log\frac{\pi}{1-\pi}.
$$

For two groups with success probabilities $\pi_1, \pi_2$, the **odds ratio** is
$$
\theta \;=\; \frac{\pi_1/(1-\pi_1)}{\pi_2/(1-\pi_2)}, \qquad \log \theta \;=\; \operatorname{logit}(\pi_1) - \operatorname{logit}(\pi_2).
$$

Compute it from the 2×2 cross-tabulation `public × has_vacancy`:

::: panel-tabset
## R

```{r or-r}
tab <- table(public = df$public, has_vacancy = df$has_vacancy)
tab
odds_pub  <- tab["1", "1"] / tab["1", "0"]
odds_priv <- tab["0", "1"] / tab["0", "0"]
c(odds_public  = round(odds_pub,  4),
  odds_private = round(odds_priv, 4),
  OR           = round(odds_pub / odds_priv, 4),
  log_OR       = round(log(odds_pub / odds_priv), 4))
```

## Python

```{python or-py}
tab = pd.crosstab(df["public"], df["has_vacancy"])
tab
```

```{python or-stats-py}
odds_pub  = tab.loc["1", 1] / tab.loc["1", 0]
odds_priv = tab.loc["0", 1] / tab.loc["0", 0]
{"odds_public":  round(odds_pub,  4),
 "odds_private": round(odds_priv, 4),
 "OR":           round(odds_pub / odds_priv, 4),
 "log_OR":       round(np.log(odds_pub / odds_priv), 4)}
```

## Julia

```{julia or-jl}
tab = combine(groupby(df, [:public, :has_vacancy]), nrow => :n)
unstack(tab, :public, :has_vacancy, :n)
```

```{julia or-stats-jl}
n11 = sum((df.public .== "1") .& (df.has_vacancy .== 1));
n10 = sum((df.public .== "1") .& (df.has_vacancy .== 0));
n01 = sum((df.public .== "0") .& (df.has_vacancy .== 1));
n00 = sum((df.public .== "0") .& (df.has_vacancy .== 0));
odds_pub  = n11 / n10;
odds_priv = n01 / n00;
(odds_public  = round(odds_pub,  digits = 4),
 odds_private = round(odds_priv, digits = 4),
 OR           = round(odds_pub / odds_priv, digits = 4),
 log_OR       = round(log(odds_pub / odds_priv), digits = 4))
```
:::

The OR is about $1.13$: the odds of having a vacancy are roughly $13\%$ higher for public firms than for private ones. The log-OR ($\approx 0.12$) is exactly what a logistic regression on `public` recovers as $\hat{\beta}_1$.

# Logistic regression: fit and interpret

Logistic regression specifies
$$
\operatorname{logit}(\pi_i) = \beta_0 + \mathbf{x}_i^\top\boldsymbol{\beta}, \qquad
\pi_i = \frac{\exp(\beta_0 + \mathbf{x}_i^\top\boldsymbol{\beta})}{1 + \exp(\beta_0 + \mathbf{x}_i^\top\boldsymbol{\beta})}.
$$

::: panel-tabset
## R

```{r fit-public-r}
m1 <- glm(has_vacancy ~ public, data = df, family = binomial(link = "logit"))
summary(m1)
```

```{r m1-or-r}
round(exp(coef(m1)), 4)
```

The exponentiated coefficient on `public1` matches the OR computed from the 2×2 table.

```{r fit-full-r}
m_full <- glm(has_vacancy ~ public + size + nace, data = df,
              family = binomial(link = "logit"))
summary(m_full)
```

`modelsummary` with `exponentiate = TRUE` prints odds ratios with CIs:

```{r modelsummary-r}
modelsummary(list(`logit (public)` = m1,
                  `logit (full)`   = m_full),
             exponentiate = TRUE,
             statistic    = "({std.error})",
             gof_omit     = "F|RMSE")
```

```{r anova-r}
car::Anova(m_full, type = 2) |> broom::tidy()
```

## Python

```{python fit-public-py}
m1 = smf.glm("has_vacancy ~ C(public)", data = df,
             family = sm.families.Binomial()).fit()
print(m1.summary())
```

```{python m1-or-py}
np.exp(m1.params).round(4)
```

```{python fit-full-py}
m_full = smf.glm("has_vacancy ~ C(public) + C(size) + C(nace)", data = df,
                 family = sm.families.Binomial()).fit()
print(m_full.summary())
```

## Julia

```{julia fit-public-jl}
m1 = glm(@formula(has_vacancy ~ public), df, Bernoulli(), LogitLink())
```

```{julia m1-or-jl}
exp.(coef(m1))
```

```{julia fit-full-jl}
m_full = glm(@formula(has_vacancy ~ public + size + nace), df,
             Bernoulli(), LogitLink())
```
:::

**Interpretation.** A unit change in $X_j$ (others fixed) shifts the log-odds by $\beta_j$:
$$
\beta_j = \operatorname{logit}(\pi_+) - \operatorname{logit}(\pi_-) = \log \theta_j.
$$
So $\hat{\beta}_j$ is the log odds ratio, and $\exp(\hat{\beta}_j)$ is the odds ratio.

# Other links: probit and cloglog

The link $g$ in $g(\pi_i) = \beta_0 + \mathbf{x}_i^\top\boldsymbol{\beta}$ does not have to be the logit:

- **logit**: $g(\pi) = \log\frac{\pi}{1-\pi}$.
- **probit**: $g(\pi) = \Phi^{-1}(\pi)$.
- **cloglog**: $g(\pi) = \log(-\log(1-\pi))$ — asymmetric, used when the positive (or negative) outcome is rare.

Logit and probit produce nearly identical fitted probabilities (probit coefficients are about $\beta^{\text{logit}}/1.6$); only the logit gives log-odds-ratio coefficients.

::: panel-tabset
## R

```{r probit-r}
m_logit   <- glm(has_vacancy ~ public + size + nace, data = df,
                 family = binomial(link = "logit"))
m_probit  <- glm(has_vacancy ~ public + size + nace, data = df,
                 family = binomial(link = "probit"))
m_cloglog <- glm(has_vacancy ~ public + size + nace, data = df,
                 family = binomial(link = "cloglog"))

# Coefficients side-by-side (link scale)
modelsummary(list(logit   = m_logit,
                  probit  = m_probit,
                  cloglog = m_cloglog),
             statistic = "({std.error})",
             gof_omit  = "F|RMSE")
```

```{r ratio-logit-probit-r}
round(coef(m_logit) / coef(m_probit), 3)
```

## Python

```{python probit-py}
m_logit  = smf.glm("has_vacancy ~ C(public) + C(size) + C(nace)", data = df,
                   family = sm.families.Binomial(sm.families.links.Logit())).fit()
m_probit = smf.glm("has_vacancy ~ C(public) + C(size) + C(nace)", data = df,
                   family = sm.families.Binomial(sm.families.links.Probit())).fit()
ratio = (m_logit.params / m_probit.params).round(3)
print(ratio)
```

## Julia

```{julia probit-jl}
m_logit  = glm(@formula(has_vacancy ~ public + size + nace), df,
               Bernoulli(), LogitLink())
m_probit = glm(@formula(has_vacancy ~ public + size + nace), df,
               Bernoulli(), ProbitLink())
round.(coef(m_logit) ./ coef(m_probit), digits = 3)
```
:::

The logit-to-probit coefficient ratio is close to $1.6$, as expected.

# Marginal effects

Coefficients live on the link scale. To express effects on the **probability scale**, differentiate $\pi_i = G(\beta_0 + \mathbf{x}_i^\top\boldsymbol{\beta})$:
$$
\frac{\partial \pi_i}{\partial x_{ij}} \;=\; g(\hat{\eta}_i)\,\hat{\beta}_j,
$$
where $g$ is the inverse-link density. Three flavors:

- **AME** (average marginal effect): average of $g(\hat{\eta}_i)\hat{\beta}_j$ over the sample;
- **MEM** (at the mean): $g(\hat{\beta}_0 + \bar{\mathbf{x}}^\top\hat{\boldsymbol{\beta}})\hat{\beta}_j$;
- **MER** (at representative values): evaluated at user-chosen $\mathbf{x}^\star$.

The AME is the default in `marginaleffects`, `statsmodels`, and Stata's `margins`.

::: panel-tabset
## R

```{r ame-r}
avg_slopes(m_full)
```

`marginaleffects` differentiates $\pi$ for each observation, then averages — the right thing to do when the link is non-identity.

```{r mem-r}
slopes(m_full, newdata = "mean")
```

```{r mer-r}
predictions(m_full,
            newdata = datagrid(public = c("0", "1"),
                               size   = "Medium",
                               nace   = "C")) |>
  as.data.frame() |>
  subset(select = c("public", "size", "nace", "estimate", "conf.low", "conf.high"))
```

## Python

```{python ame-py}
mfx = m_full.get_margeff(at = "overall", method = "dydx")
print(mfx.summary())
```

```{python mem-py}
mfx_mean = m_full.get_margeff(at = "mean", method = "dydx")
print(mfx_mean.summary())
```

## Julia

`Effects.jl` provides predicted probabilities at user-chosen design points; for AME you can differentiate manually:

```{julia ame-jl}
pi_hat = predict(m_full)
g_hat  = pi_hat .* (1 .- pi_hat)         # logistic density g(eta) = pi*(1-pi)
beta   = coef(m_full)
ame_public = mean(g_hat) * beta[2]
round(ame_public, digits = 4)
```
:::

The AME for `public` is about $0.013$ — public firms are roughly $1.3$ percentage points more likely to post a vacancy, against a baseline rate of $\sim 12\%$.

# Goodness of fit and discrimination

Fit, discrimination, and calibration can all disagree: a model can have low deviance but poor AUC, or high AUC but bad calibration. Always look at multiple summaries.

::: panel-tabset
## R

Likelihood-based summaries:

```{r gof-r}
c(deviance     = round(deviance(m_full), 1),
  AIC          = round(AIC(m_full), 1),
  BIC          = round(BIC(m_full), 1),
  null_dev     = round(m_full$null.deviance, 1),
  df_residual  = m_full$df.residual)
```

McFadden pseudo-$R^2$ and friends via `performance`:

```{r perf-r2-r}
performance::r2_mcfadden(m_full)
performance::r2_tjur(m_full)
```

LR test against a reduced model (drop `nace`):

```{r lr-r}
m_red <- glm(has_vacancy ~ public + size, data = df,
             family = binomial(link = "logit"))
anova(m_red, m_full, test = "LRT")
```

ROC and AUC:

```{r auc-r}
df$pi_hat <- predict(m_full, type = "response")
roc_obj <- pROC::roc(df$has_vacancy, df$pi_hat, quiet = TRUE)
auc(roc_obj)
plot(roc_obj, main = "ROC for full logistic model")
```

Calibration plot (deciles of $\hat{\pi}$ vs.\ observed proportion):

```{r calib-r}
df$bin <- cut(df$pi_hat,
              breaks = quantile(df$pi_hat, probs = seq(0, 1, by = 0.1)),
              include.lowest = TRUE)
calib <- aggregate(cbind(pi_hat, has_vacancy) ~ bin, data = df, FUN = mean)
plot(calib$pi_hat, calib$has_vacancy,
     xlim = c(0, max(calib$pi_hat) * 1.05),
     ylim = c(0, max(calib$has_vacancy) * 1.05),
     xlab = "Predicted probability (decile mean)",
     ylab = "Observed proportion",
     pch  = 19)
abline(0, 1, lty = 2)
```

## Python

```{python gof-py}
{"deviance":     round(m_full.deviance, 1),
 "AIC":          round(m_full.aic, 1),
 "BIC":          round(m_full.bic_llf, 1),
 "null_dev":     round(m_full.null_deviance, 1),
 "df_residual":  int(m_full.df_resid)}
```

```{python lr-py}
m_red = smf.glm("has_vacancy ~ C(public) + C(size)", data = df,
                family = sm.families.Binomial()).fit()
LR = 2 * (m_full.llf - m_red.llf)
df_diff = int(m_red.df_resid - m_full.df_resid)
from scipy import stats
{"LR_stat": round(LR, 2),
 "df":      df_diff,
 "p_value": stats.chi2.sf(LR, df_diff)}
```

```{python auc-py}
pi_hat = m_full.predict(df)
auc = roc_auc_score(df["has_vacancy"], pi_hat)
round(auc, 4)
```

## Julia

```{julia gof-jl}
(deviance    = round(deviance(m_full), digits = 1),
 aic         = round(aic(m_full), digits = 1),
 bic         = round(bic(m_full), digits = 1),
 df_residual = dof_residual(m_full))
```
:::

A high AUC means the model **ranks** observations well; calibration tells whether the predicted probabilities are on the right level. They can come apart, so check both.

# Aggregated data

Logistic regression accepts grouped/aggregated counts as a `cbind(successes, failures)` response in R or `freq_weights` / explicit count columns in Python. The point estimates are identical to fitting on individual data.

::: panel-tabset
## R

```{r agg-r}
agg <- aggregate(cbind(yes = has_vacancy, n = 1) ~ public + size,
                 data = df, FUN = sum)
agg$no <- agg$n - agg$yes
head(agg)
```

```{r agg-fit-r}
m_agg <- glm(cbind(yes, no) ~ public + size, data = agg,
             family = binomial(link = "logit"))
round(coef(m_agg), 4)
```

```{r agg-vs-individual-r}
m_ind <- glm(has_vacancy ~ public + size, data = df,
             family = binomial(link = "logit"))
round(coef(m_ind), 4)
```

## Python

```{python agg-py}
agg = (df.groupby(["public", "size"], observed = True)["has_vacancy"]
         .agg(yes = "sum", n = "size")
         .reset_index())
agg["no"] = agg["n"] - agg["yes"]
agg
```

```{python agg-fit-py}
m_agg = smf.glm("yes + no ~ C(public) + C(size)", data = agg,
                family = sm.families.Binomial()).fit()
m_agg.params.round(4)
```

## Julia

```{julia agg-jl}
agg = combine(groupby(df, [:public, :size]),
              :has_vacancy => sum => :yes,
              nrow => :n);
agg.no = agg.n .- agg.yes;
first(agg, 6)
```
:::

The two routes give identical $\hat{\boldsymbol{\beta}}$.

# Exercises

## Exercise 1 — link choice

Fit the same `has_vacancy ~ public + size + nace` model with logit, probit, and cloglog links. Compare:

1. fitted probabilities (scatter `logit` vs. `probit`);
2. AICs;
3. ROC AUC.

Does any link clearly dominate?

## Exercise 2 — calibration after adding NACE division

Refit the full model adding `nace_division` (the 2-digit code) on top of `nace`. Does AUC improve? Does calibration improve at the high-$\hat{\pi}$ end?

## Exercise 3 — separation diagnostic

For each NACE letter, compute the proportion of firms with `has_vacancy = 1`. Are there sectors with $0\%$ or $100\%$ vacancies? In a logistic regression with `nace` as a predictor, what would you expect to happen to the corresponding coefficient and its SE?

# Further reading

- **Theory.** Agresti, A. (2013). *Categorical Data Analysis* (3rd ed.). Wiley. Hosmer, D. W., Lemeshow, S., & Sturdivant, R. X. (2013). *Applied Logistic Regression* (3rd ed.). Wiley.
- **Marginal effects.** Arel-Bundock, V. (2024). *Marginal and Conditional Effects in R*. <https://marginaleffects.com/>.
- **LPM, robust SEs.** Wooldridge, J. M. (2010). *Econometric Analysis of Cross Section and Panel Data* (2nd ed.). MIT Press, ch. 15.
- **Diagnostics.** `DHARMa` (simulated residuals for binomial GLMs), `pROC`, `performance` (calibration, pseudo-$R^2$), `marginaleffects` (AME, MEM, MER).
- **Calibration.** Steyerberg, E. W. (2019). *Clinical Prediction Models* (2nd ed.). Springer, ch. 15.
