---
title: "Classification and regression trees"
author: "Maciej Beręsewicz"
format:
  html:
    embed-resources: true
    table-of-contents: true
    number-sections: true
    df-print: kable
editor: source
execute:
  eval: true
  warning: false
  message: false
filters:
  - ripper
ripper:
  output-dir: "extracted"
  script-links-position: "none"
  include-yaml: true
---

```{r setup, echo = FALSE}
JuliaCall::julia_setup(JULIA_HOME = "/Applications/Julia-1.11.app/Contents/Resources/julia/bin/")
knitr::opts_chunk$set(engine.path = list(
  python = "/opt/anaconda3/bin/python3.12",
  julia = "/Applications/Julia-1.11.app/Contents/Resources/julia/bin/"
))
```

# Packages

::: panel-tabset
## R

```{r pkg-install-r, eval = FALSE}
install.packages("rpart")
install.packages("rpart.plot")
install.packages("partykit")
```

```{r pkg-load-r}
library(rpart)
library(rpart.plot)
library(partykit)
set.seed(123)
```

## Python

```{bash pkg-install-py, eval = FALSE}
!pip install scikit-learn matplotlib
```

```{python pkg-load-py}
import numpy as np
import pandas as pd
import matplotlib.pyplot as plt
from sklearn.tree import DecisionTreeRegressor, DecisionTreeClassifier, plot_tree
from sklearn.linear_model import LinearRegression, LogisticRegression
np.random.seed(123)
```

## Julia

```{julia pkg-install-jl, eval = FALSE}
using Pkg
Pkg.add(["DecisionTree", "GLM", "DataFrames", "CSV", "CategoricalArrays",
         "Statistics", "Random", "Plots", "Distributions"])
```

```{julia pkg-load-jl}
using DecisionTree
using GLM
using DataFrames
using CSV
using CategoricalArrays
using Statistics
using Random
using Distributions
Random.seed!(123);
```
:::

# Why trees?

A classification or regression tree approximates $f(\mathbf{x}) = E[Y \mid X = \mathbf{x}]$ by a **piecewise-constant** function on a partition of the predictor space. This notebook walks through four examples of increasing complexity:

1. continuous $Y$ with one continuous $X$ that has a non-linear shape -- OLS vs.\ a regression tree;
2. binary $Y$ with one continuous $X$ -- logistic regression vs.\ a classification tree;
3. binary $Y$ with one binary $X$ pair -- where the response is an interaction (XOR-style);
4. the real Polish Job Vacancy Survey -- reproducing the toy tree from the slides.

After each example there is a short exercise.

# Example 1 -- continuous $Y$, continuous $X$ (non-linear)

The data-generating process is
$$
Y_i = \sin(2\pi X_i) + \varepsilon_i, \qquad X_i \sim U(0, 1), \qquad \varepsilon_i \sim N(0, 0.3^2).
$$

The true regression function is smooth and non-monotone. OLS in $X$ alone can only fit a line; the tree carves the $X$-axis into intervals and predicts the mean of $Y$ on each.

::: panel-tabset
## R

```{r ex1-sim-r}
n <- 300
x <- runif(n)
y <- sin(2 * pi * x) + rnorm(n, sd = 0.3)
d1 <- data.frame(x = x, y = y)
```

```{r ex1-fit-r}
fit_lm   <- lm(y ~ x, data = d1)
fit_tree <- rpart(y ~ x, data = d1,
                  method = "anova",                ## "anova" = regression tree (SSR splits); use "class" for classification
                  control = rpart.control(
                    cp       = 0.01,               ## complexity threshold: refuse a split unless it cuts the (scaled) error by >= cp
                    minsplit = 20))                ## a node with fewer than 20 obs is never split
```

```{r ex1-tree-r}
rpart.plot(fit_tree, type = 4, extra = 1, digits = 3)
```

```{r ex1-compare-r, fig.width = 7, fig.height = 4}
xs <- seq(0, 1, length = 400)
yhat_lm   <- predict(fit_lm,   newdata = data.frame(x = xs))
yhat_tree <- predict(fit_tree, newdata = data.frame(x = xs))

plot(x, y, pch = 19, col = "grey60", cex = 0.7,
     xlab = "x", ylab = "y", main = "OLS vs. regression tree")
lines(xs, sin(2 * pi * xs), col = "black", lwd = 2)
lines(xs, yhat_lm,   col = "red",  lwd = 2)
lines(xs, yhat_tree, col = "blue", lwd = 2)
legend("topright",
       legend = c("truth", "OLS", "rpart"),
       col    = c("black", "red", "blue"),
       lwd = 2, bty = "n")
```

In-sample MSE:

```{r ex1-mse-r}
c(OLS  = mean((d1$y - predict(fit_lm))^2),
  tree = mean((d1$y - predict(fit_tree))^2))
```

## Python

```{python ex1-sim-py}
n = 300
x = np.random.uniform(0, 1, n)
y = np.sin(2 * np.pi * x) + np.random.normal(0, 0.3, n)
X = x.reshape(-1, 1)
```

```{python ex1-fit-py}
fit_lm   = LinearRegression().fit(X, y)
fit_tree = DecisionTreeRegressor(
    max_depth         = 4,   ## hard cap on tree depth; None = grow until pure
    min_samples_split = 20   ## a node with fewer than 20 obs is never split
).fit(X, y)
```

```{python ex1-tree-py}
fig, ax = plt.subplots(figsize = (8, 5))
plot_tree(fit_tree, filled = True, rounded = True, feature_names = ["x"], ax = ax)
plt.show()
```

```{python ex1-compare-py}
xs = np.linspace(0, 1, 400).reshape(-1, 1)
plt.figure(figsize = (7, 4))
plt.scatter(x, y, color = "grey", s = 12)
plt.plot(xs, np.sin(2 * np.pi * xs),  color = "black", lw = 2, label = "truth")
plt.plot(xs, fit_lm.predict(xs),      color = "red",   lw = 2, label = "OLS")
plt.plot(xs, fit_tree.predict(xs),    color = "blue",  lw = 2, label = "tree")
plt.xlabel("x"); plt.ylabel("y"); plt.legend(); plt.title("OLS vs. regression tree")
plt.show()
```

```{python ex1-mse-py}
{"OLS":  float(((y - fit_lm.predict(X))**2).mean()),
 "tree": float(((y - fit_tree.predict(X))**2).mean())}
```

## Julia

```{julia ex1-sim-jl}
n = 300;
x = rand(n);
y = sin.(2 * pi .* x) .+ randn(n) .* 0.3;
X = reshape(x, n, 1);
```

```{julia ex1-fit-jl}
fit_lm   = lm(@formula(y ~ x), DataFrame(x = x, y = y));
fit_tree = DecisionTreeRegressor(
    max_depth         = 4,    ## hard cap on tree depth; -1 = grow until pure
    min_samples_split = 20    ## a node with fewer than 20 obs is never split
);
DecisionTree.fit!(fit_tree, X, y);
```

```{julia ex1-tree-jl}
print_tree(fit_tree, 3)
```

```{julia ex1-mse-jl}
yhat_lm   = GLM.predict(fit_lm);
yhat_tree = DecisionTree.predict(fit_tree, X);
(OLS  = mean((y .- yhat_lm).^2),
 tree = mean((y .- yhat_tree).^2))
```
:::

The tree fits a step function that follows the sine curve, while OLS misses the shape almost entirely. The in-sample MSE for the tree is much smaller -- the cost is a piecewise-constant fit with visible discontinuities at the splits.

## Exercise 1

Replace the sine wave with a **quadratic** trend, $Y_i = (X_i - 0.5)^2 + \varepsilon_i$, with the same noise level. Refit OLS and a regression tree. Two questions:

1. Does the tree still help, or does plain OLS now suffice?
2. Add the squared term $X^2$ to the OLS formula. How does its in-sample MSE compare to the tree's?

# Example 2 -- binary $Y$, continuous $X$

The DGP has a **non-monotone** success probability:
$$
P(Y_i = 1 \mid X_i) \;=\; \tfrac{1}{2} + 0.4 \cdot \sin(2\pi X_i), \qquad X_i \sim U(0, 1).
$$

A plain logistic regression in $X$ alone is constrained to a monotone S-curve and will fail to recover the bumps; a classification tree splits $X$ into bins matching the high- and low-probability regions.

::: panel-tabset
## R

```{r ex2-sim-r}
n <- 500
x <- runif(n)
p <- 0.5 + 0.4 * sin(2 * pi * x)
y <- rbinom(n, size = 1, prob = p)
d2 <- data.frame(x = x, y = factor(y))
```

```{r ex2-fit-r}
fit_logit <- glm(y ~ x, data = d2, family = binomial())
fit_ctree <- rpart(y ~ x, data = d2,
                   method = "class",                ## "class" = classification tree (Gini splits by default)
                   control = rpart.control(
                     cp        = 0.005,             ## smaller cp than the default 0.01 -> allow a deeper tree
                     minbucket = 20))               ## every leaf must contain at least 20 obs
```

```{r ex2-tree-r}
rpart.plot(fit_ctree, type = 4, extra = 104, digits = 3)
```

```{r ex2-compare-r, fig.width = 7, fig.height = 4}
xs <- seq(0, 1, length = 400)
p_logit <- predict(fit_logit, newdata = data.frame(x = xs), type = "response")
p_tree  <- predict(fit_ctree, newdata = data.frame(x = xs), type = "prob")[, "1"]

plot(xs, 0.5 + 0.4 * sin(2 * pi * xs),
     type = "l", lwd = 2, ylim = c(0, 1),
     xlab = "x", ylab = "P(Y = 1 | x)", main = "Logistic vs. classification tree")
lines(xs, p_logit, col = "red",  lwd = 2)
lines(xs, p_tree,  col = "blue", lwd = 2)
points(x, as.numeric(as.character(d2$y)), pch = "|", col = "grey60")
legend("topright",
       legend = c("truth", "logit", "tree"),
       col    = c("black", "red", "blue"),
       lwd = 2, bty = "n")
```

In-sample classification error:

```{r ex2-err-r}
err_logit <- mean((predict(fit_logit, type = "response") > 0.5) != (d2$y == "1"))
err_tree  <- mean(predict(fit_ctree, type = "class") != d2$y)
c(logit = err_logit, tree = err_tree)
```

## Python

```{python ex2-sim-py}
n = 500
x = np.random.uniform(0, 1, n)
p = 0.5 + 0.4 * np.sin(2 * np.pi * x)
y = np.random.binomial(1, p)
X = x.reshape(-1, 1)
```

```{python ex2-fit-py}
fit_logit = LogisticRegression().fit(X, y)
fit_ctree = DecisionTreeClassifier(
    max_depth        = 4,    ## hard cap on tree depth
    min_samples_leaf = 20    ## every leaf must contain at least 20 obs
).fit(X, y)
```

```{python ex2-tree-py}
fig, ax = plt.subplots(figsize = (8, 5))
plot_tree(fit_ctree, filled = True, rounded = True,
          feature_names = ["x"], class_names = ["0", "1"], ax = ax)
plt.show()
```

```{python ex2-compare-py}
xs = np.linspace(0, 1, 400).reshape(-1, 1)
plt.figure(figsize = (7, 4))
plt.plot(xs, 0.5 + 0.4 * np.sin(2 * np.pi * xs.ravel()),
         color = "black", lw = 2, label = "truth")
plt.plot(xs, fit_logit.predict_proba(xs)[:, 1], color = "red",  lw = 2, label = "logit")
plt.plot(xs, fit_ctree.predict_proba(xs)[:, 1], color = "blue", lw = 2, label = "tree")
plt.scatter(x, y, color = "grey", s = 4)
plt.xlabel("x"); plt.ylabel("P(Y = 1 | x)"); plt.legend()
plt.title("Logistic vs. classification tree")
plt.show()
```

```{python ex2-err-py}
{"logit": float((fit_logit.predict(X) != y).mean()),
 "tree":  float((fit_ctree.predict(X) != y).mean())}
```

## Julia

```{julia ex2-sim-jl}
n = 500;
x = rand(n);
p = 0.5 .+ 0.4 .* sin.(2 * pi .* x);
y = Int.(rand(n) .< p);
X = reshape(x, n, 1);
```

```{julia ex2-fit-jl}
fit_logit = glm(@formula(y ~ x), DataFrame(x = x, y = y), Binomial(), LogitLink());
fit_ctree = DecisionTreeClassifier(
    max_depth        = 4,     ## hard cap on tree depth
    min_samples_leaf = 20     ## every leaf must contain at least 20 obs
);
DecisionTree.fit!(fit_ctree, X, y);
```

```{julia ex2-tree-jl}
print_tree(fit_ctree, 3)
```

```{julia ex2-err-jl}
yhat_logit = Int.(GLM.predict(fit_logit) .> 0.5);
yhat_tree  = DecisionTree.predict(fit_ctree, X);
(logit = mean(yhat_logit .!= y),
 tree  = mean(yhat_tree  .!= y))
```
:::

The logistic curve flattens at $0.5$ -- it correctly notices that on average $P(Y = 1) = 0.5$, but cannot see the bumps. The tree splits $x$ into regions and reports the bump heights.

## Exercise 2

Change the DGP to $P(Y = 1 \mid X) = \tfrac{1}{2} + 0.4 \cdot \sin(4\pi X)$ -- the probability now oscillates **twice** as fast. Refit both models. How deep does the tree need to be (parameter `cp` / `max_depth`) before it picks up all four turning points? Plot the resulting probability curves on top of the truth.

# Example 3 -- binary $Y$, binary $X$ (interaction / XOR)

A pure interaction effect: $Y$ depends on whether $X_1$ and $X_2$ **agree**, not on either one alone.

|       | $X_2 = 0$ | $X_2 = 1$ |
|-------|-----------|-----------|
| $X_1 = 0$ | $P = 0.85$ | $P = 0.15$ |
| $X_1 = 1$ | $P = 0.15$ | $P = 0.85$ |

A logistic regression with **main effects only** sees both marginal effects as zero. A tree with depth 2 recovers the four cells exactly.

::: panel-tabset
## R

```{r ex3-sim-r}
n <- 1000
x1 <- rbinom(n, 1, 0.5)
x2 <- rbinom(n, 1, 0.5)
p  <- ifelse(x1 == x2, 0.85, 0.15)
y  <- rbinom(n, 1, p)
d3 <- data.frame(x1 = factor(x1), x2 = factor(x2), y = factor(y))
```

```{r ex3-table-r}
addmargins(prop.table(table(x1 = d3$x1, x2 = d3$x2, y = d3$y), c(1, 2)), 3)
```

```{r ex3-fit-r}
fit_logit_main <- glm(y ~ x1 + x2,    data = d3, family = binomial())   ## main effects only -> blind to XOR
fit_logit_int  <- glm(y ~ x1 * x2,    data = d3, family = binomial())   ## "x1 * x2" expands to x1 + x2 + x1:x2
fit_tree3      <- rpart(y ~ x1 + x2,  data = d3, method = "class")      ## defaults: cp = 0.01, minsplit = 20, minbucket = round(minsplit/3)
```

```{r ex3-tree-r}
rpart.plot(fit_tree3, type = 4, extra = 104, digits = 3)
```

```{r ex3-err-r}
err <- function(pred, y) mean(pred != y)
c(logit_main = err(predict(fit_logit_main, type = "response") > 0.5, d3$y == "1"),
  logit_int  = err(predict(fit_logit_int,  type = "response") > 0.5, d3$y == "1"),
  tree       = err(predict(fit_tree3, type = "class"), d3$y))
```

## Python

```{python ex3-sim-py}
n = 1000
x1 = np.random.binomial(1, 0.5, n)
x2 = np.random.binomial(1, 0.5, n)
p  = np.where(x1 == x2, 0.85, 0.15)
y  = np.random.binomial(1, p)
X  = np.column_stack([x1, x2])
```

```{python ex3-fit-py}
fit_logit_main = LogisticRegression(C = 1e6).fit(X, y)              ## C = 1 / penalty; large C ≈ unpenalised MLE
X_int = np.column_stack([x1, x2, x1 * x2])                          ## add the x1 * x2 interaction column by hand
fit_logit_int  = LogisticRegression(C = 1e6).fit(X_int, y)
fit_tree3      = DecisionTreeClassifier(max_depth = 2).fit(X, y)    ## depth 2 is exactly enough for a 2x2 XOR
```

```{python ex3-tree-py}
fig, ax = plt.subplots(figsize = (8, 5))
plot_tree(fit_tree3, filled = True, rounded = True,
          feature_names = ["x1", "x2"], class_names = ["0", "1"], ax = ax)
plt.show()
```

```{python ex3-err-py}
{"logit_main": float((fit_logit_main.predict(X)     != y).mean()),
 "logit_int":  float((fit_logit_int.predict(X_int)  != y).mean()),
 "tree":       float((fit_tree3.predict(X)          != y).mean())}
```

## Julia

```{julia ex3-sim-jl}
n = 1000;
x1 = rand(0:1, n);
x2 = rand(0:1, n);
p  = ifelse.(x1 .== x2, 0.85, 0.15);
y  = Int.(rand(n) .< p);
X  = hcat(x1, x2);
```

```{julia ex3-fit-jl}
df3 = DataFrame(x1 = x1, x2 = x2, y = y);
fit_logit_main = glm(@formula(y ~ x1 + x2), df3, Binomial(), LogitLink());   ## main effects only -> blind to XOR
fit_logit_int  = glm(@formula(y ~ x1 * x2), df3, Binomial(), LogitLink());   ## "x1 * x2" expands to x1 + x2 + x1&x2
fit_tree3 = DecisionTreeClassifier(max_depth = 2);                           ## depth 2 is exactly enough for a 2x2 XOR
DecisionTree.fit!(fit_tree3, X, y);
```

```{julia ex3-tree-jl}
print_tree(fit_tree3, 3)
```

```{julia ex3-err-jl}
yhat_main = Int.(GLM.predict(fit_logit_main) .> 0.5);
yhat_int  = Int.(GLM.predict(fit_logit_int)  .> 0.5);
yhat_tree = DecisionTree.predict(fit_tree3, X);
(logit_main = mean(yhat_main .!= y),
 logit_int  = mean(yhat_int  .!= y),
 tree       = mean(yhat_tree .!= y))
```
:::

The main-effects logistic predicts class $1$ in roughly half of the rows -- error rate $\approx 0.5$. The logistic with the explicit `x1:x2` interaction matches the tree exactly. **The tree did not need to be told about the interaction**: a depth-2 split in $x_1$ then $x_2$ recovers it from the data.

## Exercise 3

Switch to an *unbalanced* XOR: keep $P(Y = 1 \mid x_1 = x_2) = 0.85$, but lower $P(Y = 1 \mid x_1 \ne x_2)$ to $0.40$. Refit the three models and compare error rates again. Then add a **noise variable** $x_3 \sim$ Bernoulli$(0.5)$ that is independent of $Y$, refit the tree, and report which variable it splits on first.

# Application -- the Polish Job Vacancy Survey

Now move to the real data set used throughout the course.

::: panel-tabset
## R

```{r jvs-data-r}
df <- read.csv("data/polish-jvs.csv",
               colClasses = c("character", "factor", rep("character", 4), "numeric"))
df$size        <- factor(df$size, levels = c("Large", "Medium", "Small"))
df$public      <- factor(df$public)
df$nace        <- factor(df$nace)
df$has_vacancy <- as.integer(df$vacancies > 0)
nrow(df)
```

## Python

```{python jvs-data-py}
df = pd.read_csv("data/polish-jvs.csv",
                 dtype = {"id": str, "woj": str, "public": str,
                          "size": str, "nace_division": str, "nace": str})
df["size"]        = pd.Categorical(df["size"], categories = ["Large", "Medium", "Small"])
df["has_vacancy"] = (df["vacancies"] > 0).astype(int)
len(df)
```

## Julia

```{julia jvs-data-jl}
df = CSV.read("data/polish-jvs.csv", DataFrame,
              types = Dict(:id => String, :woj => String, :public => String,
                           :size => String, :nace_division => String, :nace => String));
df.size        = categorical(df.size, levels = ["Large", "Medium", "Small"]);
df.public      = categorical(df.public);
df.has_vacancy = ifelse.(df.vacancies .> 0, 1, 0);
nrow(df)
```
:::

## Reproducing the slide example -- a 3-leaf tree

The slides show a tree of depth 2 on `has_vacancy` using `size` and `public`. Force the same shape here:

::: panel-tabset
## R

```{r jvs-toy-r}
## Gotcha: with method = "class" on a heavily imbalanced binary outcome
## (88% zeros), rpart's cp filter is based on misclassification cost.
## Because no single split flips the majority class in any leaf, every
## split is scored as "zero improvement" and rpart returns the root only.
## Fix: fit as a regression tree on the 0/1 indicator -- for binary Y the
## SSR criterion is equivalent to Gini (var(Y) = p(1 - p) = Gini / 2).
jvs_toy <- rpart(has_vacancy ~ size + public, data = df,
                 method = "anova",                  ## regression on 0/1 = Gini classification for binary Y
                 control = rpart.control(
                   maxdepth = 2,                    ## stop at depth 2 (root = depth 0)
                   cp       = 0))                   ## let depth (not cp) bind; leaf values are p(has_vacancy = 1)
rpart.plot(jvs_toy, type = 4, extra = 101, digits = 3)
```

```{r jvs-toy-table-r}
aggregate(cbind(n = rep(1, nrow(df)),
                p_has = df$has_vacancy,
                mean_vac = df$vacancies)
          ~ size + public, data = df, FUN = sum) -> .agg
within(.agg, {p_has <- p_has / n; mean_vac <- mean_vac / n})
```

## Python

```{python jvs-toy-py}
from sklearn.preprocessing import OrdinalEncoder
enc = OrdinalEncoder(categories = [["Large", "Medium", "Small"], ["0", "1"]])
X_toy = enc.fit_transform(df[["size", "public"]].astype(str))
y_toy = df["has_vacancy"].to_numpy()
jvs_toy = DecisionTreeClassifier(max_depth = 2).fit(X_toy, y_toy)   ## depth 2 ⇒ at most 4 leaves; matches the slide
fig, ax = plt.subplots(figsize = (8, 5))
plot_tree(jvs_toy, filled = True, rounded = True,
          feature_names = ["size", "public"], class_names = ["0", "1"], ax = ax)
plt.show()
```

```{python jvs-toy-table-py}
(df.groupby(["size", "public"], observed = True)
   .agg(n = ("has_vacancy", "size"),
        p_has = ("has_vacancy", "mean"),
        mean_vac = ("vacancies", "mean"))
   .round(3))
```

## Julia

```{julia jvs-toy-jl}
size_code   = levelcode.(df.size);
public_code = levelcode.(df.public);
X_toy = hcat(size_code, public_code);             ## ";" suppresses printing the 57480x2 matrix
y_toy = df.has_vacancy;
jvs_toy = DecisionTreeClassifier(max_depth = 2)   ## depth 2 ⇒ at most 4 leaves; matches the slide
DecisionTree.fit!(jvs_toy, X_toy, y_toy);
print_tree(jvs_toy, 3)                            ## second argument = max depth to print (truncates display, not the model)
```

```{julia jvs-toy-table-jl}
combine(groupby(df, [:size, :public]),
        nrow => :n,
        :has_vacancy => mean => :p_has,
        :vacancies   => mean => :mean_vac)
```
:::

The leaf probabilities and means line up with the numbers reported in the slides: Large/private $\approx 0.19$, Large/public $\approx 0.17$, Small or Medium $\approx 0.07$.

## Exercise 4

Re-fit the tree **without** the depth restriction (use the default `cp = 0.01`) and let it use `size`, `public`, **and** `nace_division` ($\approx 90$ levels). Report (i) the number of leaves, (ii) the first three variables it splits on, and (iii) the in-sample classification error. Does `nace_division` ever appear in the top of the tree?

# Pruning with cost-complexity

A fully grown tree overfits. CART grows a large tree and prunes it back by minimising
$$
R_\alpha(T) \;=\; \sum_{m = 1}^{|T|} \sum_{i \in R_m} \ell(y_i, \hat y_{R_m}) \;+\; \alpha\,|T|.
$$

The complexity parameter $\alpha$ is chosen by cross-validation. `rpart` does this for you and stores the CV table in `$cptable`.

::: panel-tabset
## R

```{r prune-r}
## Same gotcha as above: fit as anova on the 0/1 indicator so that rpart
## actually grows a tree on this heavily imbalanced outcome.
jvs_big <- rpart(has_vacancy ~ size + public + nace_division + woj, data = df,
                 method = "anova",                  ## regression on 0/1; leaf values are p(has_vacancy = 1)
                 control = rpart.control(
                   cp        = 0.0005,              ## very small cp -> grow an over-large tree, prune it back below
                   minbucket = 50))                 ## min 50 obs per leaf, otherwise the deeper splits are unstable
printcp(jvs_big)                                    ## CV table of (cp, |T|, train error, xerror, xstd)
plotcp(jvs_big)                                     ## visualises the same table with the 1-SE bar
```

```{r prune-1se-r}
cp1se <- jvs_big$cptable[, "CP"][which.min(jvs_big$cptable[, "xerror"])]
jvs_pruned <- prune(jvs_big, cp = cp1se)
c(leaves_full   = sum(jvs_big$frame$var == "<leaf>"),
  leaves_pruned = sum(jvs_pruned$frame$var == "<leaf>"))
```

## Python

```{python prune-py}
from sklearn.tree import DecisionTreeClassifier
features_py = ["size", "public", "nace_division", "woj"]
enc_py = OrdinalEncoder(handle_unknown = "use_encoded_value", unknown_value = -1)
X_py   = enc_py.fit_transform(df[features_py].astype(str))
y_py   = df["has_vacancy"].to_numpy()

tree_full = DecisionTreeClassifier(
    min_samples_leaf = 50,        ## min 50 obs per leaf
    random_state     = 123        ## seed for tie-breaking among splits with equal impurity
)
path = tree_full.cost_complexity_pruning_path(X_py, y_py)    ## returns the (alpha, total_impurity) pruning path
alphas = path.ccp_alphas[::max(1, len(path.ccp_alphas) // 20)]   ## subsample ~20 alphas along the path for a quick scan

scores = []
for a in alphas:
    t = DecisionTreeClassifier(
        min_samples_leaf = 50,
        ccp_alpha        = a,     ## cost-complexity pruning weight; sklearn's name for rpart's `cp`
        random_state     = 123
    )
    t.fit(X_py, y_py)
    scores.append({"alpha": float(a),
                   "leaves": int(t.get_n_leaves()),
                   "train_err": float((t.predict(X_py) != y_py).mean())})
pd.DataFrame(scores)
```

## Julia

```{julia prune-jl}
size_c = levelcode.(df.size);
pub_c  = levelcode.(df.public);
nace_c = levelcode.(categorical(df.nace_division));
woj_c  = levelcode.(categorical(df.woj));
X_full = hcat(size_c, pub_c, nace_c, woj_c);          ## ";" suppresses the big matrix print
y_full = df.has_vacancy;

depths = [2, 4, 6, 8, 10]
results = DataFrame(max_depth = Int[], train_err = Float64[])
for d in depths
    t = DecisionTreeClassifier(
        max_depth        = d,    ## sweep depth as a proxy for cost-complexity pruning (DecisionTree.jl has no ccp_alpha)
        min_samples_leaf = 50    ## min 50 obs per leaf
    )
    DecisionTree.fit!(t, X_full, y_full)
    push!(results, (d, mean(DecisionTree.predict(t, X_full) .!= y_full)))
end
results
```
:::

The cross-validated pruning curve shows the classic U-shape: a tree that is too small under-fits, a tree that is too large over-fits, and the **1-SE rule** picks the smallest tree within one standard error of the best CV error.

## Exercise 5

Fit a **regression tree** for `vacancies` (not `has_vacancy`) with `method = "anova"` and the same predictors. Plot the pruning curve and select the tree at the 1-SE rule. What is the MSE of this pruned tree compared to a Poisson GLM with the same predictors?

# Conditional inference trees and ensembles (teaser)

Two extensions that fix the biggest weakness of CART (selection bias toward high-cardinality predictors and high variance of a single tree):

* **Conditional inference trees** (`partykit::ctree`) -- decouple variable selection from split selection by using permutation tests.
* **Random forest** -- average many trees fit on bootstrap samples with a random subset of predictors at each split.

::: panel-tabset
## R

```{r ctree-r}
ct <- partykit::ctree(factor(has_vacancy) ~ size + public + nace,
                      data = df,
                      control = partykit::ctree_control(
                        maxdepth     = 3,           ## maximum tree depth
                        mincriterion = 0.99))       ## 1 - p-value cut-off; only split when the permutation test p < 0.01
print(ct)
```

## Python

```{python rf-py}
from sklearn.ensemble import RandomForestClassifier
rf = RandomForestClassifier(
    n_estimators     = 200,   ## number of bootstrap trees in the forest
    min_samples_leaf = 50,    ## min 50 obs per leaf in every tree
    random_state     = 123,   ## seed for bootstrapping and the random feature subset at each split
    n_jobs           = -1     ## -1 = use all available CPU cores
).fit(X_py, y_py)
imp = pd.Series(rf.feature_importances_, index = features_py).sort_values(ascending = False)
imp.round(3)
```

## Julia

```{julia rf-jl}
rf = RandomForestClassifier(
    n_trees          = 200,    ## number of trees in the forest
    partial_sampling = 0.7,    ## fraction of training rows used to grow each tree (bootstrap subsample fraction)
    min_samples_leaf = 50      ## min 50 obs per leaf in every tree
);
DecisionTree.fit!(rf, X_full, y_full);
mean(DecisionTree.predict(rf, X_full) .!= y_full)
```
:::

## Exercise 6

Compare a **single classification tree**, a **random forest**, and a **logistic regression** on the same set of predictors. Split the JVS data into a 70/30 train-test split, fit each model on the training set, and report **test-set accuracy** and **AUC** for each. Which model wins?

# Further reading

- **Textbook.** Hastie, T., Tibshirani, R., & Friedman, J. (2009). *The Elements of Statistical Learning* (2nd ed.), Chapter 9. Springer. doi:10.1007/978-0-387-84858-7.
- **Gentle introduction.** James, G., Witten, D., Hastie, T., & Tibshirani, R. (2021). *An Introduction to Statistical Learning* (2nd ed.), Chapter 8. Springer. doi:10.1007/978-1-0716-1418-1.
- **Original reference.** Breiman, L., Friedman, J. H., Olshen, R. A., & Stone, C. J. (1984). *Classification and Regression Trees*. Wadsworth.
- **R packages.** Therneau, T. & Atkinson, B. -- `rpart` vignette on CRAN; Hothorn, T., Hornik, K., & Zeileis, A. (2006). Unbiased recursive partitioning. *JCGS*, 15(3), 651–674. doi:10.1198/106186006X133933 -- the `partykit::ctree` paper.
- **Ensembles.** Breiman, L. (2001). Random forests. *Machine Learning*, 45(1), 5–32. doi:10.1023/A:1010933404324. Friedman, J. H. (2001). Greedy function approximation: A gradient boosting machine. *Annals of Statistics*, 29(5), 1189–1232. doi:10.1214/aos/1013203451. Chen, T., & Guestrin, C. (2016). XGBoost. *KDD '16*, 785–794. doi:10.1145/2939672.2939785.
- **Fifty-year retrospective.** Loh, W.-Y. (2014). Fifty years of classification and regression trees. *International Statistical Review*, 82(3), 329–348. doi:10.1111/insr.12016.
