---
title: "Problem Set 1: Vacancy Analysis"
author: "Categorical Data Analysis 2025/26"
format:
  html:
    embed-resources: true
    number-sections: true
    df-print: kable
---

## Instructions

- Use your **student ID number** (nr albumu) as the seed — every student works with a **unique dataset**.
- You may use **R, Python, or Julia** (pick one and use it consistently).
- **Total: 10 points. Deadline: one week from assignment date.**

::: {.callout-warning}
## Submission requirements

1. Submit a **single self-contained HTML file** via **Moodle**.
2. The file name must follow the pattern: `hw01-<student_id>.html` (e.g., `hw01-130149.html`).
3. The file **must contain**:
   - Your **student ID** and **NACE section letter** clearly stated at the top.
   - All **code** used to produce the answers (visible in the document).
   - All **outputs** (numbers, tables, plots) rendered inline.
   - Written answers to interpretation questions.
4. **How to produce the HTML file:**
   - **R users**: Write your solution in a `.qmd` or `.Rmd` file, then render with `quarto render` or `knit`.
   - **Python users**: Write in a Jupyter notebook (`.ipynb`), then export: *File → Download as → HTML*.
   - **Julia users**: Write in a `.qmd` file with Julia engine, then render with `quarto render`.
5. **Before submitting**, verify that the HTML file opens in a browser and shows all outputs correctly.
6. Submissions that do not contain visible code and outputs will **not be graded**.

**How to use this file:**

- **R users**: Save this file as `hw01-<student_id>.qmd`. Delete the Python code blocks, keep the R blocks. Fill in the R code and render with `quarto render hw01-<student_id>.qmd`.
- **Python users**: Use the `.ipynb` version provided. Delete the R code cells, keep the Python cells. Fill in the code and export as HTML: *File → Download as → HTML*.
:::

## Step 0: Set your seed and prepare data

Replace `XXXXX` with your student ID number. Use **one** of the following (R or Python) and keep using the same language throughout.

::: panel-tabset
### R

```{r, eval = F}
## Replace XXXXX with your student ID
set.seed(XXXXX)
jvs <- read.csv("polish-jvs.csv")
my_data <- jvs[sample(nrow(jvs), 3000), ]
top3 <- names(sort(table(my_data$nace), decreasing = TRUE))[1:3]
my_section <- sample(top3, 1)
cat("My NACE section:", my_section, "\n")
section_data <- my_data[my_data$nace == my_section, ]
cat("Number of rows:", nrow(section_data), "\n")
```

### Python

```{python, eval = F}
## Replace XXXXX with your student ID
import numpy as np
import pandas as pd
import scipy.stats as st
from scipy.optimize import minimize

STUDENT_ID = XXXXX

np.random.seed(STUDENT_ID)
jvs = pd.read_csv("polish-jvs.csv")
my_data = jvs.sample(3000)
top3 = my_data['nace'].value_counts().head(3).index.tolist()
my_section = np.random.choice(top3, 1)[0]
print("My NACE section:", my_section)
section_data = my_data[my_data['nace'] == my_section]
print("Number of rows:", len(section_data))
```
:::

**My student ID:** *XXXXX*

**My NACE section:** *...*

**Number of rows in section_data:** *...*


## Part A: Categorical data (2 points)

### a) Frequency table of `size` (1 pt)

Create a frequency table of the `size` variable. Which size category is the most common?

```{r}
## Your code here (R)

```

```{python}
## Your code here (Python)

```

**Answer:** The most common size is *...* with *...* observations.


### b) Two-way table `size` x `public` (1 pt)

Create a two-way table of `size` x `public`. What proportion of companies in your section are both **Small** and **private** (`public == 0`)? Report as a percentage with 2 decimal places.

```{r}
## Your code here (R)

```

```{python}
## Your code here (Python)

```

**Answer:** The proportion of Small and private companies is *...* %.


## Part B: Distribution of vacancies (3 points)

### a) Sample statistics (1 pt)

Compute the sample mean ($\bar{x}$), sample variance ($s^2$), and dispersion index ($D = s^2 / \bar{x}$) for the `vacancies` variable.

```{r}
## Your code here (R)

```

```{python}
## Your code here (Python)

```

**Answer:**

- Sample mean: *...*
- Sample variance: *...*
- Dispersion index: *...*


### b) Zero proportion comparison (1 pt)

Compute the observed proportion of zeros and compare with the theoretical Poisson probability $P(X = 0) = e^{-\bar{x}}$.

```{r}
## Your code here (R)

```

```{python}
## Your code here (Python)

```

**Answer:**

- Observed P(X=0): *...*
- Theoretical Poisson P(X=0): *...*


### c) Observed vs. expected frequency table (1 pt)

Create a comparison table for values 0, 1, 2, 3, 4, 5+. Compute expected frequencies under $\text{Poisson}(\hat{\lambda} = \bar{x})$.

```{r}
## Your code here (R)

```

```{python}
## Your code here (Python)

```

| Vacancies | Observed | Expected (Poisson) |
|-----------|----------|-------------------|
| 0         | *...*    | *...*             |
| 1         | *...*    | *...*             |
| 2         | *...*    | *...*             |
| 3         | *...*    | *...*             |
| 4         | *...*    | *...*             |
| 5+        | *...*    | *...*             |


## Part C: Maximum likelihood estimation (3 points)

### a) Closed-form MLE (1 pt)

Write the Poisson log-likelihood function and compute the MLE using the closed-form formula. Report $\hat{\lambda}$ and its standard error $\text{SE} = \sqrt{\hat{\lambda}/n}$.

```{r}
## Your code here (R)

```

```{python}
## Your code here (Python)

```

**Answer:**

- $\hat{\lambda}$ (closed form) = *...*
- SE = *...*


### b) Numerical optimization (1 pt)

Implement the log-likelihood in code and find $\hat{\lambda}$ by numerical optimization. Verify it matches part (a).

```r
## Define the negative log-likelihood function (R)
## Your code here

## Minimize it (e.g., optim with method="Brent")
## Your code here

```

```{python}
## Define the negative log-likelihood function (Python)
## Your code here

## Minimize it (e.g., scipy.optimize.minimize)
## Your code here

```

**Answer:** $\hat{\lambda}$ (numerical) = *...*


### c) Score function verification (1 pt)

Compute the score function (gradient) at $\hat{\lambda}$. Verify it is approximately zero.

The score function for the Poisson is:

$$S(\lambda) = \frac{\partial l}{\partial \lambda} = \frac{\sum_i x_i}{\lambda} - n$$

```{r}
## Your code here (R)

```

```{python}
## Your code here (Python)

```

**Answer:** Score at MLE = *...* (should be ≈ 0)


## Part D: Interpretation (2 points)

### a) Is the Poisson model adequate? (1 pt)

Is the Poisson distribution a good model for the vacancies in your NACE section? In your answer, consider: (1) what value of the dispersion index would you expect if Poisson were correct, and what did you observe? (2) Do the observed and expected frequencies from Part B(c) agree — where are the largest discrepancies? Justify in 2-3 sentences.

**Answer:** *Write your interpretation here...*


### b) Alternative distribution (1 pt)

If the Poisson model is not adequate, what alternative distribution would you suggest and why? What property of this alternative addresses the problem you identified?

**Answer:** *Write your interpretation here...*
