---
title: "Problem Set 2: Retail store analysis"
author: "Categorical Data Analysis 2025/26"
format:
  html:
    embed-resources: true
    number-sections: true
    df-print: kable
---

## Instructions

- Use the **numeric prefix of your university email** as the seed --- e.g. `12345` from `12345@student.ue.poznan.pl`. Every student works with a **unique simulated dataset**.
- You may use **R, Python, or Julia** (pick one and use it consistently for the analysis tabs).
- **Total: 10 points. Deadline: TBA.**

::: {.callout-warning}
## Submission requirements

1. Submit a **single self-contained HTML file** via **Moodle**.
2. The file name must follow the pattern: `hw02-<id>.html` (e.g., `hw02-12345.html`).
3. The file **must contain**:
   - Your **student identifier** (the numeric prefix of your email) clearly stated at the top.
   - All **code** used to produce the answers (visible in the document).
   - All **outputs** (numbers, tables, plots) rendered inline.
   - Written answers to interpretation questions.
4. **How to produce the HTML file:**
   - **R users**: Save this file as `hw02-<id>.qmd`, fill in the R blocks (delete Python/Julia blocks), and render with `quarto render hw02-<id>.qmd`.
   - **Python users**: Use the `.ipynb` version provided (convert from this qmd if needed). Fill in the Python cells and export: *File → Download as → HTML*.
   - **Julia users**: Fill in the Julia blocks in the qmd and render with `quarto render` using the Julia engine.
5. **Before submitting**, verify that the HTML file opens in a browser and shows all outputs correctly.
6. Submissions that do not contain visible code and outputs will **not be graded**.

**All students must run the R setup block in Step 0 to generate the personalised dataset CSV.** Python and Julia users load that CSV in their own analysis blocks.
:::

## Step 0: simulate your retail-store dataset

Replace `XXXXX` with the numeric prefix of your email. The R block below generates a personalised retail-store dataset and writes it to `hw02-data-<id>.csv` in the working directory. **All users (R/Python/Julia) must run this R block once.**

```{r, setup, eval = FALSE}
## Replace XXXXX with the numeric prefix of your university email
STUDENT_ID <- XXXXX
set.seed(STUDENT_ID)

N <- 1500
size <- factor(
  sample(c("Small", "Medium", "Large"), N, replace = TRUE,
         prob = c(0.5, 0.3, 0.2)),
  levels = c("Small", "Medium", "Large")
)
region <- factor(sample(c("North", "South", "East", "West"), N, replace = TRUE),
                 levels = c("North", "South", "East", "West"))
loyalty_program <- rbinom(N, 1, 0.4)
experience_years <- runif(N, 0, 20)

size_eff   <- c(Small = 0, Medium = 5, Large = 12)[as.character(size)]
region_eff <- c(North = 0, South = 3, East = -2, West = 1)[as.character(region)]

monthly_revenue <- 20 + size_eff + region_eff +
  2   * loyalty_program +
  0.5 * experience_years +
  0.3 * loyalty_program * experience_years +
  rnorm(N, 0, 5)

mu_c <- exp(0.8 + 0.2 * (size == "Large") - 0.1 * (size == "Small"))
complaints <- rnbinom(N, mu = mu_c, size = 2)

retail <- data.frame(
  store_id = 1:N, size, region, loyalty_program,
  experience_years, complaints, monthly_revenue
)

write.csv(retail, paste0("hw02-data-", STUDENT_ID, ".csv"), row.names = FALSE)
cat("Simulated", nrow(retail), "rows for student", STUDENT_ID, "\n")
str(retail)
```

Load the CSV in the language you will use for the analysis:

::: panel-tabset
### R

```{r, eval=FALSE}
library(MASS)
library(marginaleffects)
retail <- read.csv(paste0("hw02-data-", STUDENT_ID, ".csv"))
retail$size   <- factor(retail$size,   levels = c("Small", "Medium", "Large"))
retail$region <- factor(retail$region, levels = c("North", "South", "East", "West"))
```

### Python

```{python, eval=FALSE}
import numpy as np
import pandas as pd
import statsmodels.api as sm
import statsmodels.formula.api as smf
from scipy import stats
import marginaleffects as me

STUDENT_ID = XXXXX
retail = pd.read_csv(f"hw02-data-{STUDENT_ID}.csv")
retail["size"]   = pd.Categorical(retail["size"],
                                  categories=["Small", "Medium", "Large"])
retail["region"] = pd.Categorical(retail["region"],
                                  categories=["North", "South", "East", "West"])
```

### Julia

```{julia, eval=FALSE}
using CSV, DataFrames, GLM, Distributions, StatsBase, Effects, CategoricalArrays

STUDENT_ID = XXXXX
retail = CSV.read("hw02-data-$(STUDENT_ID).csv", DataFrame)
retail.size   = categorical(retail.size;
                            levels = ["Small", "Medium", "Large"])
retail.region = categorical(retail.region;
                            levels = ["North", "South", "East", "West"])
```
:::

**My student identifier:** *XXXXX*

**Number of rows:** *...*


## Part A: Goodness of fit (3 points)

### a) Dispersion diagnostics (1 pt)

Compute the sample mean, variance, and dispersion index ($D = s^2 / \bar{x}$) of the `complaints` variable. What does the value of $D$ suggest about the adequacy of a Poisson model?

::: panel-tabset
#### R

```{r}
## Your code here (R)

```

#### Python

```{python}
## Your code here (Python)

```

#### Julia

```{julia}
## Your code here (Julia)

```
:::

**Answer:**

- Mean: *...*
- Variance: *...*
- Dispersion index: *...*


### b) Poisson $\chi^2$ goodness-of-fit test (1 pt)

Fit a Poisson distribution to `complaints` by MLE (closed-form $\hat\lambda = \bar x$). Build an observed-vs-expected frequency table using categories `{0, 1, 2, 3, 4+}` and compute the Pearson $\chi^2$ statistic, degrees of freedom (remember to subtract one for the estimated parameter), and the p-value.

::: panel-tabset
#### R

```{r}
## Your code here (R)

```

#### Python

```{python}
## Your code here (Python)

```

#### Julia

```{julia}
## Your code here (Julia)

```
:::

**Answer:**

- $\hat\lambda$: *...*
- $\chi^2$ statistic: *...*
- Degrees of freedom: *...*
- p-value: *...*
- Decision at $\alpha = 0.05$: *reject / do not reject Poisson*


### c) Negative binomial model + AIC (1 pt)

Fit an intercept-only Negative Binomial model to `complaints`. Report the AIC of both the Poisson and the Negative Binomial models. Which model is preferred?

::: panel-tabset
#### R

```{r}
## Your code here (R)
## Hint: fitdistr, fitdistrplus

```

#### Python

```{python}
## Your code here (Python)

```

#### Julia

```{julia}
## Your code here (Julia)

```
:::

**Answer:**

- Poisson AIC: *...*
- Negative Binomial AIC: *...*
- Preferred model: *...*


## Part B: Linear regression (4 points)

### a) Additive model, treatment contrasts (1 pt)

Fit the linear model
$$ \texttt{monthly\_revenue} \sim \texttt{size} + \texttt{region} + \texttt{loyalty\_program} + \texttt{experience\_years}. $$
Use **treatment contrasts** (the default) with `Small` as the reference level for `size`. Report the model's $R^2$ and the coefficient on `sizeMedium`.

::: panel-tabset
#### R

```{r}
## Your code here (R)

```

#### Python

```{python}
## Your code here (Python)
## Hint: smf.ols(...).fit()

```

#### Julia

```{julia}
## Your code here (Julia)
## Hint: lm(@formula(...), retail)

```
:::

**Answer:**

- $R^2$: *...*
- `sizeMedium` coefficient: *...*


### b) Sum contrasts for `size` (1 pt)

Refit the same additive model but with `size` coded using **sum contrasts** — `contr.sum` in R, `C(size, Sum)` in Python, `EffectsCoding(base = "Large")` in Julia (the `base` argument makes `Large` the reference that gets $-1$, matching R / Python conventions). Report the intercept and the coefficient on the first sum-coded level (`size1` in R, corresponding to `Small`).

::: panel-tabset
#### R

```{r}
## Your code here (R)

```

#### Python

```{python}
## Your code here (Python)

```

#### Julia

```{julia}
## Your code here (Julia)

```
:::

**Answer:**

- Intercept (sum contrasts): *...*
- `size1` coefficient: *...*


### c) Relevel `region` to "West" (1 pt)

Using the original treatment contrasts, relevel `region` so that `"West"` is the reference. Refit the additive model and report the coefficient on `regionNorth`.

::: panel-tabset
#### R

```{r}
## Your code here (R)
## Hint: relevel

```

#### Python

```{python}
## Your code here (Python)
## Hint: C()

```

#### Julia

```{julia}
## Your code here (Julia)
## Hint: contrasts = Dict(...)

```
:::

**Answer:**

- `regionNorth` coefficient (West ref): *...*


### d) Interaction model (1 pt)

Fit the model
$$ \texttt{monthly\_revenue} \sim \texttt{size} + \texttt{region} + \texttt{loyalty\_program} \times \texttt{experience\_years}. $$
Report the coefficient of the interaction term `loyalty_program:experience_years`.

::: panel-tabset
#### R

```{r}
## Your code here (R)

```

#### Python

```{python}
## Your code here (Python)

```

#### Julia

```{julia}
## Your code here (Julia)

```
:::

**Answer:**

- `loyalty_program:experience_years` coefficient: *...*


## Part C: Marginal effects (3 points)

Use the **interaction model from Part B(d)** throughout Part C. Use `marginaleffects` in R and Python, or `Effects.jl` in Julia.

### a) Overall AME of `experience_years` (1 pt)

Compute the **Average Marginal Effect** of `experience_years` on `monthly_revenue`, averaging across every row of your dataset.

::: panel-tabset
#### R

```{r}
## Your code here (R)
## Hint: avg_slopes()

```

#### Python

```{python}
## Your code here (Python)
## Hint: me.avg_slopes()

```

#### Julia

```{julia}
## Your code here (Julia)
## Hint: effects()

```
:::

**Answer:** AME of `experience_years`: *...*


### b) AME stratified by `loyalty_program` (1 pt)

Compute the AME of `experience_years` **separately** for stores with `loyalty_program = 0` and `loyalty_program = 1`. What does the comparison tell you about the interaction?

::: panel-tabset
#### R

```{r}
## Your code here (R)
## Hint: avg_slopes()

```

#### Python

```{python}
## Your code here (Python)

```

#### Julia

```{julia}
## Your code here (Julia)

```
:::

**Answer:**

- AME of `experience_years` at `loyalty_program = 0`: *...*
- AME of `experience_years` at `loyalty_program = 1`: *...*


### c) Prediction at representative values (MERV) (1 pt)

Compute the predicted `monthly_revenue` for a store with `size = "Small"`, `region = "North"`, `loyalty_program = 0`, `experience_years = 10`.

::: panel-tabset
#### R

```{r}
## Your code here (R)
## Hint: predictions()

```

#### Python

```{python}
## Your code here (Python)

```

#### Julia

```{julia}
## Your code here (Julia)

```
:::

**Answer:** Predicted `monthly_revenue`: *...*


## Interpretation (required, manually reviewed)

Based on the interaction coefficient from B(d) and the stratified AMEs from C(b), explain in 2--3 sentences whether loyalty-programme membership **amplifies**, **dampens**, or **does not change** the effect of experience on monthly revenue. Quote specific numbers from your answers.

**Answer:** *Write your interpretation here...*
