---
title: "Problem Set 3: Insurance claims --- zero-inflated counts"
author: "Categorical Data Analysis 2025/26"
format:
  html:
    embed-resources: true
    number-sections: true
    df-print: kable
---

## Instructions

- Use the **numeric prefix of your university email** as the seed --- e.g. `12345` from `12345@student.ue.poznan.pl`. Every student works with a **unique simulated dataset**.
- You may use **R or Python** (pick one and use it consistently for the analysis tabs).
- **Total: 7 points. Deadline: 9.06.2026.**

::: {.callout-warning}
## Submission requirements

1. Submit a **single self-contained HTML file** via **Moodle**.
2. The file name must follow the pattern: `hw03-<id>.html` (e.g., `hw03-12345.html`).
3. The file **must contain**:
   - Your **student identifier** (the numeric prefix of your email) clearly stated at the top.
   - All **code** used to produce the answers (visible in the document).
   - All **outputs** (numbers, tables, plots) rendered inline.
   - Written answers to interpretation questions.
4. **How to produce the HTML file:**
   - **R users**: Save this file as `hw03-<id>.qmd`, fill in the R blocks (delete Python blocks), and render with `quarto render hw03-<id>.qmd`.
   - **Python users**: Use the `.ipynb` version provided (convert from this qmd if needed). Fill in the Python cells and export: *File → Download as → HTML*.
5. **Before submitting**, verify that the HTML file opens in a browser and shows all outputs correctly.
6. Submissions that do not contain visible code and outputs will **not be graded**.

**All students must run the R setup block in Step 0 to generate the personalised dataset CSV.** Python users load that CSV in their own analysis blocks.
:::

## Step 0: simulate your insurance-claims dataset

Replace `XXXXX` with the numeric prefix of your email. The R block below generates a personalised insurance-claims dataset and writes it to `hw03-data-<id>.csv` in the working directory. **All users (R/Python) must run this R block once.**

```{r, setup, eval = FALSE}
## Replace XXXXX with the numeric prefix of your university email
STUDENT_ID <- XXXXX
set.seed(STUDENT_ID)

N <- 1200
age_group <- factor(sample(c("Young","Middle","Senior"), N, replace = TRUE,
                           prob = c(0.30, 0.45, 0.25)),
                    levels = c("Young","Middle","Senior"))
vehicle <- factor(sample(c("Car","Van","Truck"), N, replace = TRUE,
                         prob = c(0.60, 0.25, 0.15)),
                  levels = c("Car","Van","Truck"))
urban <- rbinom(N, 1, 0.55)
policy_years <- round(runif(N, 0, 15), 1)
zi_lin <- 0.4 - 0.07 * policy_years - 0.5 * urban
pz <- plogis(zi_lin)
struct_zero <- rbinom(N, 1, pz)
age_eff <- c(Young = 0.5, Middle = 0, Senior = -0.3)[as.character(age_group)]
veh_eff <- c(Car = 0, Van = 0.2, Truck = 0.5)[as.character(vehicle)]
mu <- exp(-0.2 + age_eff + veh_eff + 0.3 * urban - 0.02 * policy_years)
claims <- ifelse(struct_zero == 1L, 0L, rpois(N, mu))
insurance <- data.frame(customer_id = 1:N, age_group, vehicle, urban,
                        policy_years, claims)
write.csv(insurance, paste0("hw03-data-", STUDENT_ID, ".csv"), row.names = FALSE)
cat("Simulated", nrow(insurance), "rows for student", STUDENT_ID, "\n")
str(insurance)
```

Load the CSV in the language you will use for the analysis:

::: panel-tabset
### R

```{r, eval=FALSE}
library(pscl)
library(lmtest)
library(marginaleffects)
insurance <- read.csv(paste0("hw03-data-", STUDENT_ID, ".csv"))
insurance$age_group <- factor(insurance$age_group,
                              levels = c("Young", "Middle", "Senior"))
insurance$vehicle   <- factor(insurance$vehicle,
                              levels = c("Car", "Van", "Truck"))
```

### Python

```{python, eval=FALSE}
import numpy as np
import pandas as pd
import statsmodels.api as sm
from patsy import dmatrices

STUDENT_ID = XXXXX
insurance = pd.read_csv(f"hw03-data-{STUDENT_ID}.csv")
insurance["age_group"] = pd.Categorical(insurance["age_group"],
                                        categories=["Young", "Middle", "Senior"])
insurance["vehicle"]   = pd.Categorical(insurance["vehicle"],
                                        categories=["Car", "Van", "Truck"])
```
:::

**My student identifier:** *XXXXX*

**Number of rows:** *...*

**Observed share of zeros:** *...*


## Part A: Model selection (3 points)

### a) Poisson GLM and predicted zeros (1 pt)

Fit a Poisson GLM
$$ \texttt{claims} \sim \texttt{age\_group} + \texttt{vehicle} + \texttt{urban} + \texttt{policy\_years}. $$
Report its **AIC** and the **mean predicted $P(Y=0)$** under Poisson (the mean of $\exp(-\hat\mu_i)$ over all rows). Compare this mean predicted $P(Y=0)$ to the **observed share of zeros** --- does Poisson under-predict the zeros?

::: panel-tabset
#### R

```{r}
## Your code here (R)

```

#### Python

```{python}
## Your code here (Python)

```
:::

**Answer:**

- Poisson AIC: *...*
- Observed share of zeros: *...*
- Mean predicted $P(Y=0)$ under Poisson: *...*
- Does Poisson under-predict zeros? *yes / no*


### b) Zero-Inflated Poisson (1 pt)

Fit a Zero-Inflated Poisson (ZIP) with **count formula** `age_group + vehicle + urban + policy_years` and **zero-inflation formula** `urban + policy_years`. Report its **AIC**.

::: panel-tabset
#### R

```{r}
## Your code here (R)

```

#### Python

```{python}
## Your code here (Python)

```
:::

**Answer:**

- ZIP AIC: *...*


### c) Comparing PO and ZIP (1 pt)

Compare the Poisson and ZIP models via **AIC**, **BIC**, and a **likelihood-ratio test**. In R, run `lmtest::lrtest(po, zip)`. In Python, compute the LR test manually: LR $= 2(\ell_{\text{ZIP}} - \ell_{\text{PO}})$, with degrees of freedom equal to the difference in the number of estimated parameters, and the p-value from `scipy.stats.chi2.sf`. Which model is preferred and why? Also show **side-by-side rootograms** for the two models.

::: panel-tabset
#### R

```{r}
## Your code here (R)

```

#### Python

```{python}
## Your code here (Python)
## Manual LR test: 

```
:::

**Answer:**

- Poisson AIC / BIC: *...* / *...*
- ZIP AIC / BIC: *...* / *...*
- LR test: statistic *...*, df *...*, p-value *...*
- Preferred model and why: *...*

*Caveat:* Poisson is a boundary case of ZIP (the zero-inflation probability $\to 0$ lies on the edge of the parameter space), so the LRT p-value is only approximate/conservative and should be read together with AIC/BIC.


## Part B: Comparing PO and ZIP estimates (2 points)

### a) The `urban` coefficient in PO vs ZIP (1 pt)

Report the coefficient on `urban` from the **Poisson** model and the **count-component** coefficient on `urban` from the **ZIP** model. Are they similar? Explain in one sentence why they can differ.

::: panel-tabset
#### R

```{r}
## Your code here (R)

```

#### Python

```{python}
## Your code here (Python)

```
:::

**Answer:**

- Poisson `urban` coefficient: *...*
- ZIP count-component `urban` coefficient: *...*
- Are they similar? Why can they differ? *...*


### b) Reproducing the zeros (1 pt)

Report the **observed share of zeros**, the **mean predicted $P(Y=0)$ under Poisson**, and the **mean predicted $P(Y=0)$ under ZIP**. Which model reproduces the zeros better?

::: panel-tabset
#### R

```{r}
## Your code here (R)

```

#### Python

```{python}
## Your code here (Python)

```
:::

**Answer:**

- Observed share of zeros: *...*
- Mean predicted $P(Y=0)$ under Poisson: *...*
- Mean predicted $P(Y=0)$ under ZIP: *...*
- Which model reproduces zeros better? *...*


## Part C: Marginal effects (2 points)

Use the **ZIP model from A(b)** throughout Part C.

### a) AME of `urban` on E(Y|X) (1 pt)

Compute the **Average Marginal Effect** of `urban` on the expected number of claims $E(Y \mid X)$.

::: panel-tabset
#### R

```{r}
## Your code here (R)

```

#### Python

```{python}
## Your code here (Python)

```
:::

**Answer:** AME of `urban` on $E(Y\mid X)$: *...*


### b) AME of `policy_years` on P(Y=0) (1 pt)

Compute the **Average Marginal Effect** of `policy_years` on the probability of a zero $P(Y=0)$, from the **zero-inflation** part of the ZIP model. Interpret the sign in one sentence.

::: panel-tabset
#### R

```{r}
## Your code here (R)

```

#### Python

```{python}
## Your code here (Python)

```
:::

**Answer:**

- AME of `policy_years` on $P(Y=0)$: *...*
- Interpretation of the sign: *...*


## Interpretation (required, manually reviewed)

In 2--3 sentences, based on the AIC/BIC/LR-test results and the zero-reproduction comparison, state whether **zero-inflation is warranted** for this dataset and what the **structural zeros** plausibly represent (e.g. customers with no real claim exposure / very safe drivers). Quote specific numbers from your answers.

**Answer:** *Write your interpretation here...*
