---
title: "CDA: case study"
format:
  html:
    self-contained: true
    table-of-contents: true
    number-sections: true
    df-print: kable
editor: source
execute:
  eval: true
  warning: false
  message: false
  echo: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(engine.path = list(
  python = "/opt/anaconda3/bin/python3.12"
))
```

# Introduction

We analyse days of absenteeism (`daysabs`) for 316 students from two Los
Angeles high schools (`school`). For each student we also have gender
(`gender`), ethnicity (`ethnic`), and standardised math (`mathnce`) and
language (`langnce`) test scores. A few rows of the dataset are shown
below.

::: panel-tabset
## R

```{r}
library(readxl)
df <- read_excel("school-absence-for-lecture.xlsx")
head(df)
```

## Python

```{python}
import pandas as pd
df = pd.read_excel("school-absence-for-lecture.xlsx")
df.head()
```
:::

# Research questions

1.  What distribution best describes the number of absent days?
2.  Do male and female students differ in average absenteeism?
3.  Are standardised test scores related to absenteeism?
4.  When all variables are considered together, which one matters most?
5.  What predicts whether a student misses **at least one** day?

# Tasks

This case study is a chance to put together what you have seen so far.
Each step has a short hint -- think about which tools from earlier
lectures fit the situation, rather than reaching for a specific name.
The whole study is sized for one lecture slot (**90 min**); suggested
timings are shown next to each task.

## Part 1 -- get to know the data *(≈20 min)*

1.  *(≈3 min)* Read the Excel file into your session.

2.  *(≈5 min)* Tabulate every categorical variable (`gender`,
    `ethnic`, `school`). *Hint:* how many observations end up in the
    smallest category? Very small groups give unstable coefficient
    estimates and can even cause complete separation in a binary
    model -- **decide what to do with such a group before fitting
    anything**.

3.  *(≈5 min)* Plot the distribution of `daysabs`. *Hint:* what does
    the shape tell you about the outcome? Are zeros over-represented?

4.  *(≈7 min)* Visualise `daysabs` against each of the predictors.
    *Hint:* pick a plot type that respects whether the predictor is
    categorical or continuous.

## Part 2 -- modelling the count *(≈35 min)*

5.  *(≈10 min)* Fit at least two regression models for `daysabs`.
    *Hint:* the outcome is a non-negative integer, often with a long
    right tail -- which model families were designed for this kind of
    response?

6.  *(≈8 min)* Compare how well each model fits. *Hint:* think about
    the mean--variance relationship each model assumes and whether
    the data agree. Information criteria can summarise the trade-off.

7.  *(≈8 min)* Pick one model and explain, in plain language, what
    the estimated coefficients say on the natural scale of the
    response.

8.  *(≈9 min)* Test whether each variable carries independent
    information. *Hint:* a single coefficient is rarely the right
    unit when a variable enters the model with several levels -- what
    type of test combines them?

## Part 3 -- modelling whether any absence occurred *(≈35 min)*

9.  *(≈3 min)* Create a new variable `at_least_one_absence` equal to
    1 when `daysabs >= 1` and 0 otherwise.

10. *(≈10 min)* Build a model relating this 0/1 outcome to every
    predictor except `daysabs`. *Hint:* the response is binary --
    which model family fits?

11. *(≈12 min)* Translate the estimated effects into a more intuitive
    scale than the linear predictor. *Hint:* which average quantity
    tells you, in the units of the response, how much each variable
    matters?

12. *(≈10 min)* Identify which predictors significantly improve the
    model. *Hint:* the same reasoning as task 8, applied here.
