---
title: "Categorical data"
format: 
  html:
    self-contained: true
    table-of-contents: true
    number-sections: true
editor: source
execute: 
  eval: true
  warning: false
  message: false
---

```{r setup, echo = FALSE}
JuliaCall::julia_setup(JULIA_HOME = "/Applications/Julia-1.11.app/Contents/Resources/julia/bin/")
knitr::opts_chunk$set(engine.path = list(
  python = "/opt/anaconda3/bin/python3.12",
  julia = "/Applications/Julia-1.11.app/Contents/Resources/julia/bin/"
))
```

# Coding categorical data

## Example 1: manual dummy coding

::: panel-tabset
## R

```{r}
df <- data.frame(group = c("A", "B", "C", "B", "A"))
df$group_B <- as.integer(df$group == "B")
df$group_C <- as.integer(df$group == "C")
df
```

## Python

```{python}
import pandas as pd

df = pd.DataFrame({"group": ["A", "B", "C", "B", "A"]})
df["group_B"] = (df["group"] == "B").astype(int)
df["group_C"] = (df["group"] == "C").astype(int)
df
```

## Julia

```{julia}
using DataFrames

df = DataFrame(group = ["A", "B", "C", "B", "A"]);
df.group_B = Int.(df.group .== "B");
df.group_C = Int.(df.group .== "C");
df
```
:::

## Example 2: factor coding via model.matrix()

::: panel-tabset
## R

```{r}
df$group <- factor(df$group)   # reference: A (alphabetical)
model.matrix(~ group, data = df)
```

change reference level

```{r}
df$group <- relevel(df$group, ref = "B")
model.matrix(~ group, data = df)
```

## Python

```{python}
from patsy import dmatrix

df["group"] = pd.Categorical(df["group"])   # reference: A (alphabetical)
dmatrix("C(group)", df, return_type="dataframe")
```

change reference level

```{python}
dmatrix("C(group, Treatment(reference='B'))", df, return_type="dataframe")
```

## Julia

```{julia}
using StatsModels, CategoricalArrays

df.group = categorical(df.group)   # reference: A (alphabetical)
modelmatrix(@formula(0 ~ group), df)
```

change reference level

```{julia, eval=F}
# TBA
```
:::

## Example 3: contrast schemes

::: panel-tabset
## R

```{r}
df <- data.frame(group = factor(c("A", "B", "C", "B", "A")))
contrasts(df$group)
```

```{r}
contrasts(df$group) <- contr.treatment(3)
model.matrix(~ group, data = df)
```

```{r}
# effects (sum) coding
contrasts(df$group) <- contr.sum(3)
model.matrix(~ group, data = df)
```

```{r}
# polynomial coding (ordinal only!)
df$size <- factor(c("S", "M", "L", "M", "S"),
                  levels   = c("S", "M", "L"),
                  ordered  = TRUE)
contrasts(df$size) <- contr.poly(3)
model.matrix(~ size, data = df)
```

## Python

```{python}
import pandas as pd
from patsy import dmatrix

df = pd.DataFrame({"group": pd.Categorical(["A", "B", "C", "B", "A"])})

# treatment (default)
dmatrix("C(group)", df, return_type="dataframe")
```

```{python}
# treatment coding -- explicit
dmatrix("C(group, Treatment)", df, return_type="dataframe")
```

```{python}
# effects (sum) coding
dmatrix("C(group, Sum)", df, return_type="dataframe")
```

```{python}
# polynomial coding (ordinal only!)
df["size"] = pd.Categorical(["S", "M", "L", "M", "S"],
                             categories=["S", "M", "L"],
                             ordered=True)
dmatrix("C(size, Poly)", df, return_type="dataframe")
```

## Julia

```{julia, eval=F}
# TBA
```
:::

# Exercises

## Exercise 1

Create the following dataset:

| id  | education |
|-----|-----------|
| 1   | primary   |
| 2   | secondary |
| 3   | tertiary  |
| 4   | secondary |
| 5   | primary   |
| 6   | tertiary  |

Exercises:

1.  Manually create $k-1 = 2$ dummy variables with `primary` as the
    reference level.
2.  Declare `education` as a factor/categorical variable with `primary`
    as the reference level and inspect the model matrix.
3.  Verify that (1) and (2) produce the same indicator columns.

## Exercise 2

Using the built-in `mtcars` dataset, focus on the `cyl` variable (number
of cylinders: 4, 6, 8).

Where to find `mtcars`:

-   **R** -- built-in, load with `data(mtcars)` or use directly
-   **Python** -- available via `statsmodels`:
    `sm.datasets.get_rdataset("mtcars").data`
-   **Julia** -- available via `RDatasets.jl`:
    `dataset("datasets", "mtcars")`

1.  Declare `cyl` as a factor/categorical variable and inspect the
    default model matrix. Which level is the reference? How many dummy
    columns are there and why?
2.  Change the reference level to `8` and inspect the model matrix
    again. How do the coefficients change?
3.  Apply effects (sum) coding and inspect the model matrix. How do the
    columns differ from (1)?
