---
title: "Contingency tables and Cramer's V"
author: "Maciej Beręsewicz"
format: 
  html:
    embed-resources: true
    table-of-contents: true
    number-sections: true
    df-print: kable
editor: source
execute: 
  eval: true
  warning: false
  message: false
---

```{r setup, echo = FALSE}
JuliaCall::julia_setup(JULIA_HOME = "/Applications/Julia-1.11.app/Contents/Resources/julia/bin/")
knitr::opts_chunk$set(engine.path = list(
  python = "/opt/anaconda3/bin/python3.12",
  julia = "/Applications/Julia-1.11.app/Contents/Resources/julia/bin/"
))
```

## Setup

::: panel-tabset
## R

Install `vcd` packages if necessary.

```{r, eval = FALSE}
install.packages("vcd")
```

```{r}
library(vcd)
```

Read in the dataset for the lecture.

```{r}
df <- read.csv("../data/polish-jvs.csv", colClasses = c(rep("character", 6), "numeric"))
head(df)
```

## Python

Load modules

```{bash, eval = F}
!pip install pingouin 
```

```{python}
import numpy as np
import scipy.stats as st
import pingouin as pg
import pandas as pd
import matplotlib.pyplot as plt
from statsmodels.graphics.mosaicplot import mosaic
from scipy.stats import chi2_contingency, fisher_exact, chi2
```

```{python}
df = pd.read_csv("../data/polish-jvs.csv", dtype={"id": np.int64, "woj":str, "public":str,"size": str, "nace_division": str, "nace": str})
df.head()
```

## Julia

Install modules

```{julia, eval = F}
using Pkg
Pkg.add(["HypothesisTests", "StatsBase", "CSV", "FreqTables"])
```

Load packages

```{julia}
using Random
using Distributions
using HypothesisTests
using StatsBase
using DataFrames
using CSV
using FreqTables
```

```{julia}
df=CSV.read("../data/polish-jvs.csv", DataFrame, 
            types = Dict(:id => Int64, :woj=> String, :public=> String, 
                         :size => String, :nace_division => String, :nace => String));
first(df, 5)
```
:::

## Contingency tables -- $\chi^2$ test and Cramer's V correlation coefficient

Is there an association between size, and having at least one vacancy?
If yes, what is the size of this effect (i.e. correlation)?

::: panel-tabset
## R

```{r}
df$vac <- df$vacancies > 0
table(df$size, df$vac)
tab1 <- with(df, table(size, vac))
tab1

prop.table(tab1, margin=1)
prop.table(tab1, margin=1) |> addmargins()
prop.table(tab1, margin=2)
prop.table(tab1, margin=2) |> addmargins()
mosaicplot(tab1)
chisq.test(tab1)
```

```{r}
assocstats(tab1) ## vcd
```

```{r}
xtabs(~size+vac, df) |> summary()
tab1 |> summary()
```

## Python

```{python}
df["vac"] = np.where(df["vacancies"] > 0, True, False)
mosaic(df, ['size', 'vac'])
plt.show()
tab1 = st.contingency.crosstab(df["size"], df["vac"])
row_props = tab1.count / tab1.count.sum(axis=1, keepdims=True)
print(row_props)
col_props = tab1.count / tab1.count.sum(axis=0, keepdims=True)
print(col_props)
res1=st.chi2_contingency(tab1.count)
[res1.statistic, np.sqrt(res1.statistic/np.sum(tab1.count))]
pg.chi2_independence(df, x = "size", y = "vac")
```

## Julia

```{julia}
df.vac = ifelse.(df.vacancies .> 0, true, false);
tab1=freqtable(df, :size, :vac)
res1=ChisqTest(tab1)
sqrt(res1.stat / sum(tab1))
```
:::

## Additional code

Codes for the slide "When $\chi^2$ approximation fails -- example"

::: panel-tabset
## R

```{r}
tab <- matrix(c(1, 5, 9, 5), nrow = 2, byrow = TRUE,
              dimnames = list(c("Treatment", "Control"),
                              c("Success", "Failure")))
tab

cat("\nExpected counts:\n")
chisq.test(tab, correct = FALSE)$expected

pearson <- chisq.test(tab, correct = FALSE)
yates   <- chisq.test(tab, correct = TRUE)
fisher  <- fisher.test(tab)

obs <- as.vector(tab)
exp_counts <- as.vector(chisq.test(tab, correct = FALSE)$expected)
G2 <- 2 * sum(obs * log(obs / exp_counts))
p_G2 <- 1 - pchisq(G2, df = 1)

results <- data.frame(
  Test      = c("Pearson chi2", "Yates chi2", "G2", "Fisher exact"),
  Statistic = c(pearson$statistic, yates$statistic, G2, NA),
  p_value   = c(pearson$p.value, yates$p.value, p_G2, fisher$p.value)
)
results
```

## Python

```{python}
tab = np.array([[1, 9],
                [5, 5]])
tab_df = pd.DataFrame(tab,
                       index=["Treatment", "Control"],
                       columns=["Success", "Failure"])
print(tab_df)

# Expected counts
_, _, _, expected = chi2_contingency(tab, correction=False)
print("\nExpected counts:")
print(pd.DataFrame(expected,
                    index=["Treatment", "Control"],
                    columns=["Success", "Failure"]))

# Pearson chi-squared (no correction)
pearson_stat, pearson_p, _, _ = chi2_contingency(tab, correction=False)

# Yates' continuity correction
yates_stat, yates_p, _, _ = chi2_contingency(tab, correction=True)

# Fisher's exact test
fisher_or, fisher_p = fisher_exact(tab)

# G^2 (log-likelihood ratio)
obs = tab.flatten()
exp_counts = expected.flatten()
G2 = 2 * np.sum(obs * np.log(obs / exp_counts))
p_G2 = 1 - chi2.cdf(G2, df=1)

results = pd.DataFrame({
    "Test":      ["Pearson chi2", "Yates chi2", "G2", "Fisher exact"],
    "Statistic": [pearson_stat, yates_stat, G2, np.nan],
    "p_value":   [pearson_p, yates_p, p_G2, fisher_p]
})
print("\n", results)
```

## Julia

```{julia}
tab = [1 9;
       5 5]
println("Observed counts:")
display(tab)

# Expected counts
n = sum(tab)
row_sums = sum(tab, dims=2)
col_sums = sum(tab, dims=1)
expected = row_sums * col_sums / n
println("\nExpected counts:")
display(expected)

# Pearson chi-squared (no correction)
pearson = ChisqTest(tab)
pearson_stat = pearson.stat
pearson_p = pvalue(pearson)

# Yates' continuity correction
obs = vec(tab)
exp_counts = vec(expected)
yates_stat = sum((abs.(obs .- exp_counts) .- 0.5).^2 ./ exp_counts)
yates_p = 1 - cdf(Chisq(1), yates_stat)

# G^2 (log-likelihood ratio)
G2 = 2 * sum(obs .* log.(obs ./ exp_counts))
p_G2 = 1 - cdf(Chisq(1), G2)

# Fisher's exact test
fisher = FisherExactTest(tab[1,1], tab[1,2], tab[2,1], tab[2,2])
fisher_p = pvalue(fisher)

results = DataFrame(
    Test      = ["Pearson chi2", "Yates chi2", "G2", "Fisher exact"],
    Statistic = [pearson_stat, yates_stat, G2, missing],
    p_value   = [pearson_p, yates_p, p_G2, fisher_p]
)
println("\n")
display(results)
```

:::

