---
title: "Discrete distributions"
author: "Maciej Beręsewicz"
format: 
  html:
    embed-resources: true
    table-of-contents: true
    number-sections: true
editor: source
execute: 
  eval: true
  warning: false
  message: false
---

```{r setup, echo = FALSE}
JuliaCall::julia_setup(JULIA_HOME = "/Applications/Julia-1.11.app/Contents/Resources/julia/bin/")
knitr::opts_chunk$set(engine.path = list(
  python = "/opt/anaconda3/bin/python3.12",
  julia = "/Applications/Julia-1.11.app/Contents/Resources/julia/bin/"
))
```

## Setup

::: panel-tabset
## R

There is no need to install or load packages. All functionalities are
build in. One may consider installing `distributions3` package which
mimics `Distributions.jl` package.

```{r, eval = FALSE}
install.packages("distributions3")
```

```{r}
library(distributions3)
```

## Python

Load modules

```{python}
import scipy.stats as st
import numpy as np
import pandas as pd
```

## Julia

Install modules

```{julia, eval = F}
using Pkg
Pkg.add(["DataFrames", "Distributions"])
```

Load packages

```{julia}
using DataFrames
using Distributions
using Random
```
:::

## Exercises from lectures

::: panel-tabset
## R

Exercise 1: generate 10 pseudo-random numbers from Bernoulli(0.7)

```{r}
set.seed(20230228)
rbinom(n=10,size=1,prob=0.7)
```

Exercise 2: generate 10 pseudo-random numbers from Binomial(50,0.7)

```{r}
set.seed(20230228)
rbinom(n=10,size=50,prob=0.7)
```

Exercise 3: generate 10 pseudo-random numbers from Poisson(1.5)

```{r}
set.seed(20230228)
rpois(n=10,lambda=1.5)
rpois(10,1.5)
```

## R (`distributions3`)

Exercise 1: generate 10 pseudo-random numbers from Bernoulli(0.7)

```{r}
set.seed(20230228)
X <- Bernoulli(0.7)
random(X, 10)
```

Exercise 2: generate 10 pseudo-random numbers from Binomial(50,0.7)

```{r}
set.seed(20230228)
X <- Binomial(50, 0.7)
random(X, 10)
```

Exercise 3: generate 10 pseudo-random numbers from Poisson(1.5)

```{r}
set.seed(20230228)
X <- Poisson(1.5)
random(X, 10)
```

## Python

Exercise 1: generate 10 pseudo-random numbers from Bernoulli(0.7)

```{python}
np.random.seed(20230228)
st.bernoulli(0.7).rvs(10)
```

Exercise 2: generate 10 pseudo-random numbers from Binomial(50,0.7)

```{python}
np.random.seed(20230228)
st.binom(50,0.7).rvs(10)
```

Exercise 3: generate 10 pseudo-random numbers from Poisson(1.5)

```{python}
np.random.seed(20230228)
st.poisson(1.5).rvs(10)
```

## Julia

Exercise 1: generate 10 pseudo-random numbers from Bernoulli(0.7)

```{julia}
Random.seed!(20230228);
rand(Bernoulli(0.7),10)
```

Exercise 2: generate 10 pseudo-random numbers from Binomial(50,0.7)

```{julia}
Random.seed!(20230228);
rand(Binomial(50,0.7),10)
```

Exercise 3: generate 10 pseudo-random numbers from Poisson(1.5)

```{julia}
Random.seed!(20230228);
rand(Poisson(1.5),10)
```
:::

## Exercises from lectures

### Exercise 1

Assume that football player with success rate 0.4 shot 10 times on goal.
Let $X$ be a random variable denoting number of successful scores.
Please find:

-   Distribution of $X$
-   Probability that football player score exactly 4 times ($P(X=4)$)
-   Probability that football player score at least 7 times
    ($P(X>=7) = 1- P(X <= 6)$)

::: panel-tabset
## R

```{r}

```

```{r}

```

```{r}

```

## R (`distributions3`)

```{r}

```

```{r}

```

```{r}

```

```{r}

```

## Python

```{python}

```

```{python}

```

```{python}

```

## Julia

Declare distribution

```{julia}

```

```{julia}

```

```{julia}

```

```{julia}

```
:::

### Exercise 2

Number of car accidents in one day in some city follows Poisson
distribution with expected value $\lambda=2$. Find the probability that
at most 4 car accidents happen.

::: panel-tabset
## R

```{r}

```

## R (`distributions3`)

```{r}

```

## Python

```{python}

```

## Julia

```{julia}

```
:::

### Exercise 3

Read in ``

