---
title: "Discrete distributions"
author: "Maciej Beręsewicz"
format: 
  html:
    embed-resources: true
    table-of-contents: true
    number-sections: true
editor: source
execute: 
  eval: true
  warning: false
  message: false
---

```{r setup, echo = FALSE}
JuliaCall::julia_setup(JULIA_HOME = "/Applications/Julia-1.11.app/Contents/Resources/julia/bin/")
knitr::opts_chunk$set(engine.path = list(
  python = "/opt/anaconda3/bin/python3.12",
  julia = "/Applications/Julia-1.11.app/Contents/Resources/julia/bin/"
))
```

## Setup

::: panel-tabset
## R

There is no need to install or load packages. All functionalities are
build in. One may consider installing `distributions3` package which
mimics `Distributions.jl` package.

```{r, eval = FALSE}
install.packages("distributions3")
```

```{r}
library(distributions3)
```

## Python

Load modules

```{python}
import scipy.stats as st
import numpy as np
import pandas as pd
```

## Julia

Install modules

```{julia, eval = F}
using Pkg
Pkg.add(["DataFrames", "Distributions"])
```

Load packages

```{julia}
using DataFrames
using Distributions
using Random
```
:::

## Exercises from lectures

::: panel-tabset
## R

Exercise 1: generate 10 pseudo-random numbers from Bernoulli(0.7)

```{r}
set.seed(20230228)
rbinom(n=10,size=1,prob=0.7)
```

Exercise 2: generate 10 pseudo-random numbers from Binomial(50,0.7)

```{r}
set.seed(20230228)
rbinom(n=10,size=50,prob=0.7)
```

Exercise 3: generate 10 pseudo-random numbers from Poisson(1.5)

```{r}
set.seed(20230228)
rpois(n=10,lambda=1.5)
rpois(10,1.5)
```

## R (`distributions3`)

Exercise 1: generate 10 pseudo-random numbers from Bernoulli(0.7)

```{r}
set.seed(20230228)
X <- Bernoulli(0.7)
random(X, 10)
```

Exercise 2: generate 10 pseudo-random numbers from Binomial(50,0.7)

```{r}
set.seed(20230228)
X <- Binomial(50, 0.7)
random(X, 10)
```

Exercise 3: generate 10 pseudo-random numbers from Poisson(1.5)

```{r}
set.seed(20230228)
X <- Poisson(1.5)
random(X, 10)
```

## Python

Exercise 1: generate 10 pseudo-random numbers from Bernoulli(0.7)

```{python}
np.random.seed(20230228)
st.bernoulli(0.7).rvs(10)
```

Exercise 2: generate 10 pseudo-random numbers from Binomial(50,0.7)

```{python}
np.random.seed(20230228)
st.binom(50,0.7).rvs(10)
```

Exercise 3: generate 10 pseudo-random numbers from Poisson(1.5)

```{python}
np.random.seed(20230228)
st.poisson(1.5).rvs(10)
```

## Julia

Exercise 1: generate 10 pseudo-random numbers from Bernoulli(0.7)

```{julia}
Random.seed!(20230228);
rand(Bernoulli(0.7),10)
```

Exercise 2: generate 10 pseudo-random numbers from Binomial(50,0.7)

```{julia}
Random.seed!(20230228);
rand(Binomial(50,0.7),10)
```

Exercise 3: generate 10 pseudo-random numbers from Poisson(1.5)

```{julia}
Random.seed!(20230228);
rand(Poisson(1.5),10)
```
:::

## Exercises from lectures

### Exercise 1

Assume that football player with success rate 0.4 shot 10 times on goal.
Let $X$ be a random variable denoting number of successful scores.
Please find:

-   Distribution of $X$
-   Probability that football player score exactly 4 times ($P(X=4)$)
-   Probability that football player score at least 7 times
    ($P(X>=7) = 1- P(X <= 6)$)

::: panel-tabset
## R

```{r}

```

```{r}

```

```{r}

```

## R (`distributions3`)

```{r}

```

```{r}

```

```{r}

```

```{r}

```

## Python

```{python}

```

```{python}

```

```{python}

```

## Julia

Declare distribution

```{julia}

```

```{julia}

```

```{julia}

```

```{julia}

```
:::

### Exercise 2

Number of car accidents in one day in some city follows Poisson
distribution with expected value $\lambda=2$. Find the probability that
at most 4 car accidents happen.

::: panel-tabset
## R

```{r}

```

## R (`distributions3`)

```{r}

```

## Python

```{python}

```

## Julia

```{julia}

```
:::

### Exercise 3

Read in `data/polish-jvs.csv` dataset (Polish Job Vacancy Survey) and do the following:

1. Calculate the average number of vacancies ($\hat{\lambda}$) from the data. Using the Poisson distribution with this $\hat{\lambda}$, calculate the probability that a randomly selected entity reports exactly 0 vacancies ($P(X=0)$). Compare with the observed proportion of zeros in the data.
2. Create a binary variable `has_vacancy` equal to 1 if `vacancies > 0` and 0 otherwise. Calculate $\hat{p}$ as the proportion of entities with at least one vacancy. Using the Binomial distribution with $n=10$ and this $\hat{p}$, calculate the probability that out of 10 randomly selected entities, at least 3 report a vacancy ($P(Y \geq 3)$).
3. Generate 1000 random draws from Poisson($\hat{\lambda}$) distribution (where $\hat{\lambda}$ is the mean from point 1). Compare the distribution of simulated values with the observed distribution of vacancies (e.g. using a table of frequencies or a bar plot).

::: panel-tabset
## R

```{r}

```

```{r}

```

```{r}

```

## R (`distributions3`)

```{r}

```

```{r}

```

```{r}

```

## Python

```{python}

```

```{python}

```

```{python}

```

## Julia

```{julia}

```

```{julia}

```

```{julia}

```
:::

## Advanced example: Poisson-Multinomial connection

Using the `polish-jvs.csv` dataset, consider 3 NACE sections: **C** (manufacturing), **G** (trade), **P** (education). If vacancies in each section follow independent Poisson distributions, then conditional on the total number of vacancies, the allocation across sections follows a Multinomial distribution.

1. For each section, calculate the per-entity mean $\hat{\lambda}_j$ and the total Poisson rate $\hat{\Lambda}_j = n_j \cdot \hat{\lambda}_j$ (where $n_j$ is the number of entities in section $j$). Compute the theoretical proportions $\hat{\pi}_j = \hat{\Lambda}_j / \sum_k \hat{\Lambda}_k$.
2. Simulate 1000 independent Poisson draws from $\mathrm{Pois}(\hat{\Lambda}_j)$ for each section. For each draw compute the total $N = X_C + X_G + X_P$ and conditional proportions $X_j / N$. Compare mean conditional proportions with $\hat{\pi}$.
3. Draw from $\mathrm{Mult}(N_{\text{obs}}, \hat{\pi})$ where $N_{\text{obs}}$ is the observed total vacancies across the 3 sections. Compare the expected counts $N_{\text{obs}} \cdot \hat{\pi}_j$ with the observed totals.

::: panel-tabset
## R

```{r}
## read data and filter 3 sections
jvs <- read.csv("../data/polish-jvs.csv")
jvs3 <- jvs[jvs$nace %in% c("C", "G", "P"), ]

## step 1: estimate per-entity lambdas, total rates (Lambda), and pi
lambda_hat <- tapply(jvs3$vacancies, jvs3$nace, mean)
n_entities <- tapply(jvs3$vacancies, jvs3$nace, length)
Lambda_hat <- n_entities * lambda_hat  ## total Poisson rates
pi_hat <- Lambda_hat / sum(Lambda_hat)
lambda_hat
Lambda_hat
pi_hat
```

```{r}
## step 2: simulate 1000 independent Poisson draws per section
set.seed(123)
n_sim <- 1000
sim_C <- rpois(n_sim, Lambda_hat["C"])
sim_G <- rpois(n_sim, Lambda_hat["G"])
sim_P <- rpois(n_sim, Lambda_hat["P"])

## conditional proportions (exclude cases where total = 0)
totals <- sim_C + sim_G + sim_P
valid <- totals > 0
cond_prop <- data.frame(
  C = mean(sim_C[valid] / totals[valid]),
  G = mean(sim_G[valid] / totals[valid]),
  P = mean(sim_P[valid] / totals[valid])
)
## compare with theoretical pi
rbind(simulated = unlist(cond_prop), theoretical = pi_hat)
```

```{r}
## step 3: multinomial -- expected vs observed
obs_totals <- tapply(jvs3$vacancies, jvs3$nace, sum)
n_total <- sum(obs_totals)
expected <- n_total * pi_hat

rbind(observed = obs_totals, expected = expected)

## simulate from multinomial
set.seed(123)
mult_draws <- rmultinom(1000, size = n_total, prob = pi_hat)
rownames(mult_draws) <- names(pi_hat)
## compare means of multinomial draws with observed
data.frame(
  observed = obs_totals,
  mult_mean = rowMeans(mult_draws)
)
```

## Python

```{python}
import pandas as pd
import numpy as np
from scipy import stats

## read data and filter 3 sections
jvs = pd.read_csv("../data/polish-jvs.csv")
jvs3 = jvs[jvs["nace"].isin(["C", "G", "P"])]

## step 1: estimate per-entity lambdas, total rates (Lambda), and pi
lambda_hat = jvs3.groupby("nace")["vacancies"].mean()
n_entities = jvs3.groupby("nace")["vacancies"].count()
Lambda_hat = n_entities * lambda_hat  ## total Poisson rates
pi_hat = Lambda_hat / Lambda_hat.sum()
print("Lambda (per-entity):", lambda_hat.to_dict())
print("Lambda (total):", Lambda_hat.to_dict())
print("Pi:", pi_hat.to_dict())
```

```{python}
## step 2: simulate 1000 independent Poisson draws per section
np.random.seed(123)
n_sim = 1000
sim = {s: stats.poisson(Lambda_hat[s]).rvs(n_sim) for s in ["C", "G", "P"]}
totals = sim["C"] + sim["G"] + sim["P"]
valid = totals > 0

cond_prop = {s: np.mean(sim[s][valid] / totals[valid]) for s in ["C", "G", "P"]}
print("Simulated proportions:", {k: round(v, 4) for k, v in cond_prop.items()})
print("Theoretical pi:", {k: round(v, 4) for k, v in pi_hat.items()})
```

```{python}
## step 3: multinomial -- expected vs observed
obs_totals = jvs3.groupby("nace")["vacancies"].sum()
n_total = obs_totals.sum()
expected = n_total * pi_hat

print("Observed:", obs_totals.to_dict())
print("Expected:", {k: round(v, 1) for k, v in expected.items()})

## simulate from multinomial
np.random.seed(123)
mult_draws = np.random.multinomial(n_total, pi_hat.values, size=1000)
print("Multinomial means:", dict(zip(pi_hat.index, mult_draws.mean(axis=0).round(1))))
```

## Julia

```{julia}
using CSV, DataFrames, Distributions, Random, Statistics

## read data and filter 3 sections
jvs = CSV.read("../data/polish-jvs.csv", DataFrame)
jvs3 = filter(r -> r.nace in ["C", "G", "P"], jvs)

## step 1: estimate per-entity lambdas, total rates (Lambda), and pi
lambda_hat = combine(groupby(jvs3, :nace),
    :vacancies => mean => :lambda,
    :vacancies => length => :n_entities)
sort!(lambda_hat, :nace)
lambda_hat.Lambda = lambda_hat.n_entities .* lambda_hat.lambda  ## total Poisson rates
lambda_hat.pi = lambda_hat.Lambda ./ sum(lambda_hat.Lambda)
lambda_hat
```

```{julia}
## step 2: simulate 1000 independent Poisson draws per section
Random.seed!(123)
n_sim = 1000
sims = Dict(r.nace => rand(Poisson(r.Lambda), n_sim) for r in eachrow(lambda_hat))
totals = sims["C"] .+ sims["G"] .+ sims["P"]
valid = totals .> 0

cond_prop = Dict(s => mean(sims[s][valid] ./ totals[valid]) for s in ["C", "G", "P"])
println("Simulated: ", cond_prop)
println("Theoretical: ", Dict(r.nace => round(r.pi, digits=4) for r in eachrow(lambda_hat)))
```

```{julia}
## step 3: multinomial -- expected vs observed
obs_totals = combine(groupby(jvs3, :nace), :vacancies => sum => :total)
sort!(obs_totals, :nace)
n_total = sum(obs_totals.total)
expected = n_total .* lambda_hat.pi

DataFrame(nace = lambda_hat.nace, observed = obs_totals.total, expected = round.(expected, digits=1))
```
:::

